---
title: "Proteomics preliminary analysis"
output: html_document
date: "2025-09-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load libraries, echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(readxl)
library(OCMSutility)
source("~/devel/MIGTranscriptomeExplorer/R/MIGTranscriptome_plotting.R")
source("~/devel/MIGTranscriptomeExplorer/R/MIGTranscriptome_retrieval.R")

```


# Overview

These data were generated from RNALater stored ileal biopsy samples from patients with PSC-UC, UC or healthy controls. The main aim of this project was to look for patterns that we have observed previously in scRNA-seq data with respect to decreases in genes involved in cellular detoxification (e.g. GSTA1, Metallothioneins etc). I am using data that Anna provided to give a basic QC and having a preliminary look at differential abundance between groups.

## Age and sex of participants

```{r age and sex, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}

group_colors <- c(blues9[5], "red3", "grey")
names(group_colors) <- c("PSC-UC", "UC", "HC")


# read metadata
prot_meta <- read.csv("~/archive/proj080/metadata/NISEED23_metadata.txt",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep="\t",
                      colClasses = c(rep(NA, 28), "NULL"))
colnames(prot_meta)[10] <- "diagnosis"
prot_meta <- prot_meta[prot_meta$UID != "",]
rownames(prot_meta) <- prot_meta$UID

prot_meta$Collection.Date..Date. <- ifelse(prot_meta$Collection.Date..Date. == "", NA , prot_meta$Collection.Date..Date.)
prot_meta$age <- as.numeric(stringi::stri_sub(prot_meta$Collection.Date..Date., -4)) - as.numeric(prot_meta$YOB..date.)

prot_meta <- prot_meta[prot_meta$diagnosis != "",]

# remove the cd one
prot_meta <- prot_meta[prot_meta$diagnosis != "PSC-CD",]

age_p <- ggplot(prot_meta, aes(x = diagnosis, y = age, color = diagnosis)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  theme_classic() +
  scale_color_manual(values = group_colors) +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3)))

sex_p <- ggplot(prot_meta, aes(x = diagnosis, fill = Sex..dropdown..M..F., alpha = 0.5)) +
  geom_bar(color = "black") +
  theme_classic() +
  scale_fill_manual(values = c("red3", "black"))

gridExtra::grid.arrange(age_p, sex_p, nrow = 1) 
```

```{r metadata table, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}

hc <- subset(prot_meta, subset = diagnosis == "HC")
psc <- subset(prot_meta, subset = diagnosis == "PSC-UC")
uc <- subset(prot_meta, subset = diagnosis == "UC")

datas <- list(hc, psc, uc)
names(datas) <- c("HC", "PSC", "UC")

# Medications - broad
biologics <- c("Adalimuab", "Adalimuab", "Ustekinumab", "Vedolizumab")
asas <- c("Pentasa", "Octasa", "Mezavant")
aza <- c("Azathioprine")
urso <- c("URSO", "UCDA", "UCDA")

metadata_summary <- list()
for (i in 1:length(datas)){
  dat <- datas[[i]]
  sex <- dat$Sex..dropdown..M..F.
  percent_f <- length(sex[sex == "F"])/length(sex)
  meds <- strsplit(dat$Medication..text., "; ")
  bs <- lapply(meds, FUN = function(x) intersect(x, biologics))
  as <- lapply(meds, FUN = function(x) intersect(x, asas))
  az <- lapply(meds, FUN = function(x) intersect(x, aza))
  ur <- lapply(meds, FUN = function(x) intersect(x, urso))
  
  bs <- unlist(lapply(bs, function(x) length(x) != 0))
  bs <- length(bs[bs == TRUE])/length(bs)
  as <- unlist(lapply(as, function(x) length(x) != 0))
  as <- length(as[as == TRUE])/length(as)
  az <- unlist(lapply(az, function(x) length(x) != 0))
  az <- length(az[az == TRUE])/length(az)
  ur <- unlist(lapply(ur, function(x) length(x) != 0))
  ur <- length(ur[ur == TRUE])/length(ur)
  
  age <- median(dat$age, na.rm = TRUE)
  sd_age <- sd(dat$age, na.rm = TRUE)
  alp <- median(dat$ALP, na.rm = TRUE)
  sd_alp <- sd(dat$ALP, na.rm = TRUE)
  
  metadata_summary[[names(datas)[i]]] <- data.frame(percent_F = round(percent_f,2),
                                                    age = age,
                                                    sd_age = sd_age,
                                                    alp = alp,
                                                    sd_alp = sd_alp,
                                                    Aza = az,
                                                    Urso = ur,
                                                    Asa = as,
                                                    Biologics = bs,
                                                    diagnosis = names(datas)[i]) 
}


metadata_summary <- dplyr::bind_rows(metadata_summary)

knitr::kable(metadata_summary)

```



## Liver biochemistry {.tabset .tabset-pills}

One of the main things that we are looking at is the effect of cholestasis on gene expression in the ileum. So here we look at liver biochemistry measurements to get an idea of the extent of cholestasis in the patients that we have data for.

### ALP

```{r alp, echo = FALSE, message = FALSE, warning = FALSE}

alp <- ggplot(prot_meta, aes(x = diagnosis, y = ALP, color = diagnosis)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  scale_color_manual(values = group_colors) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(2,3), c(1,3)), size = 3,
                             label = "p.signif") +
  ylab("ALP U/L") +
  theme(legend.position = "none") +
  geom_hline(yintercept = 162, linetype = "dashed")

alp
ggsave(alp, file = "export.dir/alp.pdf", height = 2.5, width = 2.5)
```

### ALT

```{r alt, echo = FALSE, message = FALSE, warning = FALSE}

alt <- ggplot(prot_meta, aes(x = diagnosis, y = ALT, color = diagnosis)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  scale_color_manual(values = group_colors) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(2,3), c(1,3)), size = 3,
                             label = "p.signif")

alt
```


### Bilirubin


```{r bilirubin, echo = FALSE, message = FALSE, warning = FALSE}

bil <- ggplot(prot_meta, aes(x = diagnosis, y = Bilirubin, color = diagnosis)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  scale_color_manual(values = group_colors) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(2,3), c(1,3)), size = 3,
                             label = "p.signif")

bil
```


## ALP vs. ALT

There isn't consistency between liver function tests particularly but I will have a look at the relationship between ALP and ALT anyway - likely they are correlated in the PSC patients.

```{r alt vs alp, echo = FALSE, message = FALSE, warning = FALSE}

psc_meta <- prot_meta[prot_meta$diagnosis == "PSC-UC",]

# upper limit of the HC
alp_up <- max(prot_meta$ALP[prot_meta$diagnosis == "HC"], na.rm = TRUE)
alt_up <- mean(prot_meta$ALT[prot_meta$diagnosis == "HC"], na.rm = TRUE)

ggplot(prot_meta, aes(x = ALP, y = ALT)) +
  geom_point(size = 3) +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor() +
  geom_hline(yintercept = alt_up, linetype = "dashed") +
  geom_vline(xintercept = alp_up*1.5, linetype = "dashed")


# Write out this subsetted and changed metadata
if (!dir.exists("export.dir")){
  dir.create("export.dir")
}
write.table(prot_meta,
            file = "export.dir/metadata_clean.tsv",
            row.names = FALSE,
            quote = FALSE,
            sep = "\t")

```


It's kind of tricky but not all patients have very elevated ALP and ALT levels relative to healthy individuals. Whether this impacts on downstream analysis we will need to keep and eye on.


## Protein abundance distribution

Anna has done imputation and normalisation and so the number of proteins is equivalent across samples. Here I will look at the abundance distribution to see if any filtering should be done.


```{r abundance distribution, echo=FALSE, message=FALSE}

prots <- read_xlsx("~/archive/proj080/proteomics/NickIllotResults_MSQ2955.xlsx",
                  sheet = 1)
prots <- as.data.frame(prots)

prots <- prots[,c(3,5:47)]

# filter proteins that are na
prots <- prots[!is.na(prots$`T: Genes`),]
rownames(prots) <- prots$`T: Genes`
prots <- prots[,2:ncol(prots)]

colnames(prots) <- gsub("Sample", "NI_PS_", colnames(prots))

metadata_no_na <- prot_meta[!is.na(prot_meta$ALP),]

# 162 is 1.5x the upper limit of the HC group
metadata_no_na$ALPbin <- ifelse(metadata_no_na$ALP >= 162, "High", "Low")

metadata_no_na$diagnosis <- paste0(metadata_no_na$diagnosis, metadata_no_na$ALPbin)

# get combined samples
samples <- intersect(rownames(prot_meta), colnames(prots))

prots <- prots[,samples]
prot_meta <- prot_meta[samples,]

abundance_distribution <- data.frame(gene = rownames(prots),
                                     ave_abund = rowMeans(prots, na.rm = TRUE))


ggplot(abundance_distribution, aes(ave_abund)) +
  geom_histogram() +
  theme_classic() +
  scale_x_log10()

```

There is a bit of a tail but nothing too extreme. Will keep everything for now.


## Protein abundance distribution

This is the protein abundance distribution per sample it lets us know whether the normalisation between samples has been effective.


```{r sample distributions, echo=FALSE, message = FALSE, warning = FALSE}
sample_dist <- tidyr::pivot_longer(prots, cols = everything())
sample_dist$diagnosis <- prot_meta[sample_dist$name,]$diagnosis
ggplot(sample_dist, aes(x = name, y=value, fill = diagnosis)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = group_colors)

```

Looks pretty good.


# Comparing RNA to protein

We have RNA-seq data on the same samples. We didn't see a whole lot of differences at the transcriptional level but do want to compare proteins and transcripts with the potential to perform a multi-omics integration analysis. Here I start to explore the relationship between RNA and protein.

```{r RNA vsa. protein, echo = FALSE, message = FALSE, warning = FALSE}


# get the RNA data
vst_agg <- read.csv("/well/powrie/users/utu691/work/proj080/analysis/rnaseq/differential_expression/export.dir/transformed_counts_symbols.tsv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   sep = "\t",
                   row.names = 1)

# make RNA and Protein the same samples
# protein has fewer samples
vst_agg <- vst_agg[,colnames(prots)]

# compare all - even if they aren't present in both
# datasets
rna <- data.frame(gene = rownames(vst_agg),
                  mean_rna = rowMeans(vst_agg))
prt <- data.frame(gene = rownames(prots),
                  mean_rna = rowMeans(prots))


rna_prt <- merge(rna, prt, all.x = TRUE, all.y = TRUE, by = "gene")

colnames(rna_prt) <- c("gene", "RNA_mean", "Protein_mean")
#rna_prt$RNA_mean <- ifelse(is.na(rna_prt$RNA_mean), 0, rna_prt$RNA_mean)
#rna_prt$Protein_mean <- ifelse(is.na(rna_prt$Protein_mean), -6, rna_prt$Protein_mean)


# Genes of interest in the context of GSTA1 are the GSTA1 module genes
# So create the gois list here
gsta1_genes <- list(GSTA1_Module = c("GSTA1", "MT1H", "MT1M", "HLA-B", "MT1G", "MT1F", "ACTB", "CCL25", "TMSB10", "FABP1", "FABP6", "KRT8", "LCN2", "REG1A", "FAM151A", "PIGR", "RBP2"))

# annotate with some genes of interest
rna_prt$anno <- ifelse(rna_prt$gene %in% gsta1_genes[["GSTA1_Module"]],
                       "GOI",
                       "Not_GOI")
rna_prt$gene_anno <- ifelse(rna_prt$anno == "GOI",  rna_prt$gene, NA)
                                           

rna_prt_plot <- ggplot(rna_prt, aes(x = RNA_mean, y = Protein_mean, color = anno, size = anno)) +
  geom_point(alpha = 0.5) +
  theme_classic() +
  stat_smooth(data = rna_prt, color = "snow4", method = "lm", aes(x = RNA_mean, y = Protein_mean), inherit.aes = FALSE) +
  ggpubr::stat_cor(data = rna_prt, aes(x = RNA_mean, y = Protein_mean), inherit.aes = FALSE) +
  scale_color_manual(values = c("#386CB0", "snow3")) +
  scale_size_manual(values = c(4,1)) +
  ggrepel::geom_text_repel(label = rna_prt$gene_anno) +
  geom_point(data = rna_prt[rna_prt$anno == "GOI",]) +
  theme(legend.position = "none")

rna_prt_plot
ggsave(rna_prt_plot, file = "export.dir/protein_vs_rna.pdf", height = 5, width = 5)


```


This looks pretty standard for a correlation (mean) between RNA levels and protein levels. These transcript levels are not length corrected but things may not change that much. I should use RPKM or the like in future analysis.

## Correlations across individuals

It is also of interest to look at how these genes are correlated across individuals i.e do people with higher mRNA levels have higher transcript levels.


```{r cors individuals, echo = FALSE, message = FALSE, warning = FALSE}

gois <- gsta1_genes[["GSTA1_Module"]]
gois <- gois[gois != "MT1M" & gois != "ACTB"]

to_cor <- list()
for (goi in gois){
  protein <- unlist(prots[goi, ])
  rna <- unlist(vst_agg[goi,])
  df <- data.frame(Prot = protein, RNA = rna, goi = goi)
  to_cor[[goi]] <- df
}
to_cor <- dplyr::bind_rows(to_cor)

cors_plot <- ggplot(to_cor, aes(x = RNA, y = Prot)) +
  geom_point(size = 0.5) +
  theme_classic() +
  facet_wrap(~goi, scales = "free", nrow = 3) +
  ggpubr::stat_cor(size = 1) +
  stat_smooth(method = "lm", color = "black")

ggsave(cors_plot, file = "export.dir/gois_protein_rna.pdf", height = 5, width = 6.5)
cors_plot



```

For the most part, for our genes of interest the protein levels are pretty well predicted by the RNA levels.



```{r metadata alp, echo = FALSE, message = FALSE, warning = FALSE}

# set groups based on alp - this is the same as in the RNA-seq analysis
prot_meta$ALPbin <- ifelse(prot_meta$ALP >= 162, "High", "Low")
prot_meta$diagnosis <- paste0(prot_meta$diagnosis, prot_meta$ALPbin)
metadata_no_na <- prot_meta[!is.na(prot_meta$ALP),]

```


## PCA of proteomics data

This is the PCA using proteomics data. 

```{r pca proteomics, echo = FALSE, message = FALSE, warning = FALSE}

# new group colors
new_group_colors <- group_colors
new_group_colors <- append(new_group_colors, blues9[8])
names(new_group_colors) <- c("PSCUCLow", "UCLow", "HCLow", "PSCUCHigh")

# make agegroup variable
# Make age group variable
metadata_no_na$agegroup <- ifelse(metadata_no_na$age >= 17 & metadata_no_na$age <= 25, "group1", NA)
metadata_no_na$agegroup <- ifelse(metadata_no_na$age >= 26 & metadata_no_na$age <= 35, "group2", metadata_no_na$agegroup)
metadata_no_na$agegroup <- ifelse(metadata_no_na$age >= 36 & metadata_no_na$age <= 45, "group3", metadata_no_na$agegroup)
metadata_no_na$agegroup <- ifelse(metadata_no_na$age >= 46 & metadata_no_na$age <= 55, "group4", metadata_no_na$agegroup)
metadata_no_na$agegroup <- ifelse(metadata_no_na$age >= 56, "group5", metadata_no_na$agegroup)

# impute for missing
metadata_no_na$agegroup <- ifelse(is.na(metadata_no_na$agegroup), "group5", metadata_no_na$agegroup)

# Change the sex variable
metadata_no_na$sex <- metadata_no_na$Sex..dropdown..M..F.

metadata_no_na$diagnosis <- gsub("-", "", metadata_no_na$diagnosis)

pc_prot <- runPCA(prots[,rownames(metadata_no_na)], scale = FALSE)

pca_prot <- plotPrincipleComponents(pc_prot,
                        metadata_no_na,
                        colourby = "diagnosis",
                        group = "diagnosis") +
  theme_classic() +
  stat_ellipse() +
  scale_color_manual(values = new_group_colors)

pca_prot
ggsave(pca_prot, file = "export.dir/pca_proteins.pdf", height = 2.5, width = 4)  

set.seed(10)
vegan::adonis2(t(prots[,rownames(metadata_no_na)]) ~ diagnosis + agegroup + sex,
               data = metadata_no_na,
               by = "margin",
               na.action = na.omit,
               method = "euclidean")

```

## Differential abundance

I am going to run linear models controlling for sex and age group as we did for the RNA-seq and see how our genes of interest look in terms of differences in the PSC-UC + High ALP group.  

```{r linear models, echo = FALSE, message = FALSE, warning = FALSE}

# subset the protein abundance data and the 
# metadata for common samples
samples_prots <- intersect(rownames(metadata_no_na), colnames(prots))
prt <- prots[,samples_prots]
met <- metadata_no_na[samples_prots,]

lm_results <- list()
for (i in 1:nrow(prt)){
  
  l <- lm(unlist(prt[i,]) ~ met$age + met$sex + met$diagnosis)

  x <- summary(l)
  
  # P-values
  age_p <- x[[4]][20]
  sex_p <- x[[4]][21]
  psc_high_p <- x[[4]][22]
  psc_p <- x[[4]][23]
  uc_p <- x[[4]][24]
  
  # estimates
  age_e <- x[[4]][2]
  sex_e <- x[[4]][3]
  psc_high_e <- x[[4]][4]
  psc_e <- x[[4]][5]
  uc_e <- x[[4]][6]
  
  df <- data.frame(Gene = rownames(prt)[i],
                   age_p = age_p,
                   psc_high_p = psc_high_p,
                   psc_low_p = psc_p,
                   uc_p = uc_p,
                   age_estimate = age_e,
                   sex_estimate = sex_e,
                   psc_high_estimate = psc_high_e,
                   psc_low_estimate = psc_e,
                   uc_estimate = uc_e)
 lm_results[[i]] <- df
}
lm_results <- dplyr::bind_rows(lm_results)

# table of the gois
knitr::kable(subset(lm_results, subset = Gene %in% gois))


```

## Plot the genes of interest

```{r gois plots, echo = FALSE, message = FALSE, warning = FALSE}

gois <- c("GSTA1",  "MT1H", "MT1F", "FAM151A")

## For plotting genes of interest
ss <- intersect(colnames(prots), rownames(metadata_no_na))
goi_featurebox <- featurebox(prots[,ss],
           metadata_no_na[ss,],
           group_by = "diagnosis",
           features = gois) +
  facet_wrap(~feature, scales = "free") +
  theme_classic() +
  scale_color_manual(values = new_group_colors) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45), legend.position = "none")

goi_featurebox
ggsave(goi_featurebox, file = "export.dir/protein_gois_featurebox.pdf", height = 2.5, width = 5)

```

## Correlating

```{r alp correlations, echo = FALSE, message = FALSE}


to_cor <- metadata_no_na[ss,]
to_cor$GSTA1 <- unlist(prots["GSTA1", ss])
to_cor$RBP2 <- unlist(prots["RBP2", ss])
to_cor$MT1F <- unlist(prots["MT1F", ss])
to_cor$MT1H <- unlist(prots["MT1H", ss])
to_cor$FABP6 <- unlist(prots["FABP6", ss])
to_cor$FABP1 <- unlist(prots["FABP1", ss])


ggplot(to_cor, aes(x = log10(ALP), y = GSTA1)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()


ggplot(to_cor, aes(x = log10(ALP), y = RBP2)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

ggplot(to_cor, aes(x = log10(ALP), y = MT1F)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

ggplot(to_cor, aes(x = log10(ALP), y = MT1H)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

ggplot(to_cor, aes(x = log10(ALP), y = FABP6)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

ggplot(to_cor, aes(x = log10(ALP), y = FABP1)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

```

# output data

```{r write data, echo = FALSE, message = FALSE}

# output a reduced metadata set with - UID, RNA_ID, Protein_ID, diagnosis, ALP, ALT, GGT, sex, agegroup, age, ALPbin
output_meta <- as.data.frame(metadata_no_na[,c("UID",
                                 "RNA_ID",
                                 "Protein_ID",
                                 "diagnosis",
                                 "ALP",
                                 "sex",
                                 "age",
                                 "agegroup",
                                 "ALPbin")])
write.table(output_meta,
            file = "export.dir/proteomics_metadata.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)


write.table(data.frame(Protein = rownames(prt), prt),
            file = "export.dir/proteomics_processed.tsv",
            sep = "\t",
            quote = FALSE, 
            row.names = FALSE)

```







