---
title: "Proteomics QC"
output: html_document
date: "2025-09-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load libraries, echo = FALSE, message = FALSE}

library(tidyverse)
source("~/devel/MIGTranscriptomeExplorer/R/MIGTranscriptome_plotting.R")
source("~/devel/MIGTranscriptomeExplorer/R/MIGTranscriptome_retrieval.R")

```


## Number of proteins

```{r nproteins, echo=FALSE, message=FALSE}

prots <- read.csv("~/archive/proj080/diann_results.txt",
                  header = TRUE,
                  stringsAsFactors = FALSE,
                  sep="\t")
prots <- prots[,c(3,6:ncol(prots))]

colnames(prots) <- gsub("X", "", colnames(prots))
colnames(prots) <- gsub("UC|PSC\\.UC|HC", "", colnames(prots))
colnames(prots) <- paste0("NI_PS_", colnames(prots))
colnames(prots)[1] <- "Gene" 
prots <- prots[prots$Gene != "",]

prots[prots == ""] <- NA

# aggregate by gene
rownames(prots) <- prots$Gene

# read metadata
prot_meta <- read.csv("~/archive/proj080/metadata/proteomics_metadata_sub.txt",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep="\t", 
                      row.names = 2)
colnames(prot_meta)[9] <- "diagnosis"

# get combined samples
samples <- intersect(rownames(prot_meta), colnames(prots))

prots <- prots[,samples]
prot_meta <- prot_meta[samples,]


nproteins <- data.frame(sample = colnames(prots),
                        nprot = colSums(!is.na(prots)))
ggplot(nproteins, aes(x = sample, y = nprot)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  coord_flip()

```

Remove those with < 4000 proteins.

## Prevalence abundance

```{r prev abund, echo = FALSE, message=FALSE}

prots <- prots[,colSums(!is.na(prots)) > 4000]

prev_abund <- data.frame(prevalence = rowSums(!is.na(prots)),
                         abundance = rowMeans(prots, na.rm = TRUE))
ggplot(prev_abund, aes(x = log10(abundance), y = prevalence)) +
  geom_point() +
  theme_classic()
```


## Number of samples per group


```{r abundance distribution, echo = FALSE, message = FALSE}
abundance_distribution <- data.frame(gene = rownames(prots),
                                     ave_abund = rowMeans(prots, na.rm = TRUE))


ggplot(abundance_distribution, aes(ave_abund)) +
  geom_histogram() +
  theme_classic() +
  scale_x_log10()

```


## Protein abundance distribution


```{r sample distributions, echo=FALSE, message = FALSE}

group_colors <- c(blues9[5], "red3", "grey")
names(group_colors) <- c("PSC-UC", "UC", "HC")

sample_dist <- tidyr::pivot_longer(prots, cols = everything())
sample_dist$diagnosis <- prot_meta[sample_dist$name,]$diagnosis
ggplot(sample_dist, aes(x = name, y=log10(value), fill = diagnosis)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = group_colors)


```

They actually look pretty ok with log10(). May quantile normalise as well just to be conservative.

```{r quantile, echo = FALSE, message = FALSE}
library(preprocessCore)
prots_q <- as.data.frame(normalize.quantiles(as.matrix(log10(prots)), keep.names = TRUE))

sample_dist <- tidyr::pivot_longer(prots_q, cols = everything())
sample_dist$diagnosis <- prot_meta[sample_dist$name,]$diagnosis
ggplot(sample_dist, aes(x = name, y=log10(value), fill = diagnosis)) +
  geom_boxplot() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = group_colors)

```


## PCA



```{r pca, echo=FALSE, message=FALSE}

# remove outliers


prot_meta <- prot_meta[colnames(prots_q),]

pc <- runPCA(prots_q)

plotPrincipleComponents(pc,
                        metadata = prot_meta,
                        colourby = "diagnosis",
                        group = "diagnosis",
                        pcs = c("PC1", "PC2")) +
  scale_color_manual(values = group_colors) +
  stat_ellipse()

plotPrincipleComponents(pc,
                        metadata = prot_meta,
                        colourby = "Extraction_batch",
                        group = "Extraction_batch",
                        pcs = c("PC1", "PC2")) +
  stat_ellipse()

plotPrincipleComponents(pc,
                        metadata = prot_meta,
                        colourby = "Sex..dropdown..M..F.",
                        group = "Sex..dropdown..M..F.",
                        pcs = c("PC1", "PC2")) +
  stat_ellipse()

```


## T-tests

Assume that the data are normally distributed

```{r t tests, echo = FALSE, message = FALSE}

calculate_fold_change <- function(abundance, metadata, ref = "HC"){
  # add min value to everything to get out of negatives??
  abundance <- abundance + min(abundance)
  abund_ref <- mean(abundance[metadata$diagnosis == ref])
  abund_test <- mean(abundance[metadata$diagnosis != ref])
  fold <- log2(abund_test/abund_ref)
}

run_ttest <- function(mat, metadata, ref = "HC"){
  results <- list()
  for (i in 1:nrow(mat)){
    abund <- unlist(mat[i,])
    abund_ref <- abund[metadata$diagnosis == ref]
    abund_test <- abund[metadata$diagnosis != ref]
    t_res <- t.test(abund_ref, abund_test)
    p_val <- unlist(t_res[3], use.names = FALSE)
    fold <- calculate_fold_change(abund, metadata, ref = ref)
    results[[i]] <- data.frame("t" = t_res[1],
                               "p_val" = p_val,
                               "log2FoldChange" = fold)
    }
  results <- dplyr::bind_rows(results)
  rownames(results) <- rownames(mat)
  results$padj <- p.adjust(results$p_val, method = "BH")
  results <- results[order(results$padj, decreasing = FALSE),]
  return(results)
}

meta_psc <- prot_meta[prot_meta$diagnosis %in% c("HC", "PSC-UC"),]
prots_psc <- prots[,rownames(meta_psc)]

res_psc <- run_ttest(prots_psc, meta_psc)
res_psc$fcp <- -log10(res_psc$p_val)*res_psc$log2FoldChange
res_psc <- res_psc[order(res_psc$fcp, decreasing = FALSE),]



```

## Pathway enrichment

There isn't anything particularly significant from this analysis Will look for pathway enrichment based on fold change.

```{r pathways, echo=FALSE, message=FALSE}

pathways <- fgsea::gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")

genes <- res_psc$log2FoldChange
names(genes) <- rownames(res_psc)
genes <- genes[is.finite(genes)]

fgsea_res_psc <- fgsea::fgsea(pathways, genes, minSize = 50, maxSize = 500)
fgsea_res_psc <- as.data.frame(fgsea_res_psc[order(fgsea_res_psc$padj, decreasing = FALSE),])

genes_sig_down <- rownames(res_psc[res_psc$p_val < 0.05 & res_psc$log2FoldChange > 0,])

fgsea_fora_res <- fgsea::fora(pathways, genes_sig_down, minSize = 50, maxSize = 500, universe = rownames(res_psc))
fgsea_fora_res <- as.data.frame(fgsea_fora_res[order(fgsea_fora_res$padj, decreasing = FALSE),])

```





## Linear models

Using maaslin2 because it is easy to implement...


```{r linear models, echo = FALSE, message - FALSE}

library(Maaslin2)

Maaslin2(prots,
                prot_meta,
                min_abundance = -Inf,
                min_prevalence = 0,
                normalization = "NONE",
                transform = "NONE",
                analysis_method = "LM",
                output = "maaslin2",
                fixed_effects = "diagnosis",
                reference = "diagnosis,HC")

res <- read.csv("maaslin2/all_results.tsv",
                header = TRUE, 
                stringsAsFactors = FALSE,
                sep="\t")

res_psc <- res[res$value == "PSC-UC",]


genes <- res_psc$coef
names(genes) <- res_psc$feature
genes <- genes[is.finite(genes)]

oxphos <- pathways[grep("GO_OXIDATIVE_PHOSPHORYLATION", names(pathways))]

fgsea_res_psc <- fgsea::fgsea(oxphos, genes, minSize = 50, maxSize = 500)
fgsea_res_psc <- as.data.frame(fgsea_res_psc[order(fgsea_res_psc$padj, decreasing = FALSE),])


genes_sig_down <- res_psc$feature[res$pval < 0.05 & res$coef < 0]

fgsea_fora_res <- fgsea::fora(pathways, genes_sig_down, minSize = 50, maxSize = 500, universe = res_psc$feature)
fgsea_fora_res <- as.data.frame(fgsea_fora_res[order(fgsea_fora_res$padj, decreasing = FALSE),])


res <- res[order(res$pval, decreasing = FALSE),]
res[res$feature == "GSTA1",]

library(OCMSutility)

featurebox(prots, prot_meta, features = c("AK4","COX7A1","COX7C","NDUFA3","DMAC2L","SDHD","SNCA","NDUFA1"), group_by = "diagnosis") +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3)), method = "t.test") +
  facet_wrap(~feature, scales = "free")


```
