---
title: "This is a bulk RNA-seq analysis of PSC-UC, UC and Healthy controls"
output: 
  html_document:
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(bitmapType = "cairo")
```

```{r load libraries, echo=FALSE, message=FALSE, warning=FALSE}

library(DESeq2)
library(ComplexHeatmap)
library(ggplot2)
library(tidyr)
library(fgsea)
library(tximport)
source("/well/powrie/shared/proj064/MIGTranscriptome_retrieval.R")
source("/well/powrie/shared/proj064/MIGTranscriptome_plotting.R")
library(OCMSutility)
library(RColorBrewer)
#source("convenience_functions.R")
source("~/devel/NGSKit/R/kmeans_helper.R")

if (!(dir.exists("export.dir"))){
  dir.create("export.dir")
}

# metadata is same as proteomics data
metadata <- read.csv("../../qc/export.dir/metadata_clean.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep = "\t")
rownames(metadata) <- metadata$UID
                 
```

## Overview

These are RNA-seq data that are part of a set of ileum samples that also had proteomics and metabolomics performed on them.


```{r read in data, echo=FALSE, message=FALSE, warning=FALSE}

# transcript2gene mapping
tx2gene <- read.csv("~/work/proj080/analysis/rnaseq/kallisto/transcripts2genes.dir/transcripts2genes.tsv", header=TRUE, stringsAsFactors=FALSE, sep="\t")

files <- list.files(".")[grep("*abundance.tsv", list.files("."))]
names(files) <- gsub("_abundance.tsv", "", files)

txi <- tximport(paste0(".", "/", files), type = "kallisto", tx2gene = tx2gene)

# create column names that match the metadata
counts <- as.data.frame(txi$counts)
colnames(counts) <- names(files)

# just make sure everything is in everything
samples <- intersect(rownames(metadata), colnames(counts))
counts <- counts[,samples]
metadata <- metadata[samples,]

# map colors - try and make the diagnosis names without
# hyphens in
metadata$diagnosis <- gsub("-", "", metadata$diagnosis)
group_colors <- c("red3", blues9[5], "grey")
names(group_colors) <- c("UC", "PSCUC", "HC")

```



```{r gene id to gene names, echo=FALSE, message = FALSE, warning=FALSE}

geneid2genename <- read.csv("/well/powrie/users/utu691/work/proj080/analysis/rnaseq/gene_names/geneid2genenames2.tsv",
                            header=TRUE,
                            stringsAsFactors=FALSE,
                            sep="\t")
rownames(geneid2genename) <- geneid2genename$gene_id

```

## Counts per group

The first thing to check is just that the number of reads coming from each condition is equivalent. From the kallisto pseudoalignment we saw ~80% of reads could be pseudoaligned and so the data should be of high quality for differential expression analysis.

```{r read counts, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}

counts_to_plot <- data.frame(sample=rownames(metadata),
                             group=metadata$diagnosis,
                            count=colSums(counts))

ggplot(counts_to_plot, aes(x=group, y=count, color=group)) +
  geom_boxplot() +
  geom_jitter(height=0, width=0.1) +
  theme_classic() +
  ggpubr::stat_compare_means() +
  scale_color_manual(values = group_colors)

```

Although there looks by eye to be some difference in counts between groups this is not significant and is unlikely to affect downstream analysis. 

```{r aggregate counts, echo = FALSE, message = FALSE}

counts_agg <- aggregate(counts, by = list(geneid2genename[rownames(counts),]$gene_name),
                        FUN = "mean")

rownames(counts_agg) <- counts_agg$Group.1
counts_agg <- counts_agg[,2:ncol(counts_agg)]
counts_agg <- round(counts_agg, 0)

# cpm for DSSC
counts_cpm <- edgeR::cpm(counts_agg)

# write out the cpm
write.table(data.frame(gene = rownames(counts_cpm),
                       counts_cpm),
            file = "export.dir/cpm.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)

# write out the cpm
write.table(data.frame(gene = rownames(counts_agg),
                       counts_agg),
            file = "export.dir/counts.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)

```


## PCA

This is a PCA of everything to begin with.

```{r differential expression, echo=FALSE, message=FALSE, warning=FALSE}

# Make age group variable
metadata$agegroup <- ifelse(metadata$age >= 17 & metadata$age <= 25, "group1", NA)
metadata$agegroup <- ifelse(metadata$age >= 26 & metadata$age <= 35, "group2", metadata$agegroup)
metadata$agegroup <- ifelse(metadata$age >= 36 & metadata$age <= 45, "group3", metadata$agegroup)
metadata$agegroup <- ifelse(metadata$age >= 46 & metadata$age <= 55, "group4", metadata$agegroup)
metadata$agegroup <- ifelse(metadata$age >= 56, "group5", metadata$agegroup)

# impute for missing
metadata$agegroup <- ifelse(is.na(metadata$agegroup), "group5", metadata$agegroup)


# change the disease so it doesn't have odd characters in it
metadata$diagnosis <- ifelse(metadata$diagnosis == "PSC-UC", "PSCUC", metadata$diagnosis)

dds <- DESeqDataSetFromTximport(txi, colData = metadata[names(files),], design=~diagnosis)
dds <- dds[,rownames(metadata)]

# and the age object
#dds_age <- DESeqDataSetFromTximport(txi, colData = metadata[names(files),], design=~agegroup)
#dds_age <- dds_age[,rownames(metadata)]


if (!(file.exists("export.dir/transformed_counts.tsv"))){
  
  # transformed count data
  transformed_counts <- as.data.frame(assay(vst(dds)))
  to_write <- data.frame(gene_id = rownames(transformed_counts), transformed_counts)
  
  write.table(to_write,
              file="export.dir/transformed_counts.tsv",
              sep="\t",
              quote=FALSE,
              row.names=FALSE)
}

transformed_counts <- read.csv("export.dir/transformed_counts.tsv",
                               sep="\t",
                               row.names=1)

```

### Expression distribution and PCA

This gives us a general sense of where to filter genes based on expression level - usually ~7.

```{r distribution, echo=FALSE, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}

names(group_colors) <- c("UC", "PSCUC", "HC")

transformed_counts <- as.data.frame(transformed_counts)
t_m <- reshape2::melt(transformed_counts)
ggplot(t_m, aes(x=value)) + 
  geom_histogram() + 
  theme_classic() + 
  geom_vline(xintercept = 7, linetype="dashed")
```

```{r pca, echo=FALSE, message=FALSE, fig.height=5, fig.width=12, warning=FALSE}
transformed_counts <- transformed_counts[rowSums(transformed_counts > 7) >= 5, rownames(metadata)]
pc <- runPCA(transformed_counts, scale=FALSE)

p1 <- plotPrincipleComponents(pc, metadata, group="diagnosis", colourby="diagnosis", pcs=c("PC1", "PC2")) +
  scale_color_manual(values = group_colors) +
  theme_classic() +
  stat_ellipse() +
  ggtitle("Disease status")

p2 <- plotPrincipleComponents(pc, metadata, group="agegroup", colourby="agegroup", pcs=c("PC1", "PC2")) +
  theme_classic() +
  stat_ellipse() +
  ggtitle("Age group")

p3 <- plotPrincipleComponents(pc, metadata, group="Sex..dropdown..M..F.", colourby="Sex..dropdown..M..F.", pcs=c("PC1", "PC2")) +
  theme_classic() +
  stat_ellipse() +
  ggtitle("Sex")

p4 <- plotPrincipleComponents(pc, metadata, colourby="ALP", pcs=c("PC1", "PC2"), continuous = TRUE) +
  theme_classic() +
  ggtitle("ALP") +
  scale_color_gradient2()

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)

metadata$sex <- metadata$Sex..dropdown..M..F.

set.seed(10)
vegan::adonis2(t(transformed_counts[,metadata$UID]) ~ diagnosis + agegroup + sex + ALP,
               data = metadata, by = "margin", na.action = na.omit)

```

There is some hints that ALP is associated with transcriptional profiles. This is potentially of interest as we see that some PSC patients have normal ALP levels where a small subset are high. Next we look at the PCA but instead use a separate grouping of PSC who are ALP high vs. low. Taken as High >= 162 (1.5x the max of ALP of healthy controls).


## PCA with ALP groupings


```{r pca with ALP groupings, echo = FALSE, message = FALSE}

metadata_no_na <- metadata[!is.na(metadata$ALP),]

#metadata_no_na$ALP <- scale(metadata_no_na$ALP)
metadata_no_na$ALPbin <- ifelse(metadata_no_na$ALP >= 162, "High", "Low")

metadata_no_na$diagnosis <- paste0(metadata_no_na$diagnosis, metadata_no_na$ALPbin)

# new group colors
new_group_colors <- group_colors
new_group_colors <- append(new_group_colors, blues9[8])
names(new_group_colors) <- c("UCLow", "PSCUCLow", "HCLow", "PSCUCHigh")

pc <- runPCA(transformed_counts[,metadata_no_na$UID], scale=FALSE)
p_newgroups <- plotPrincipleComponents(pc, metadata_no_na, group="diagnosis", colourby="diagnosis", pcs=c("PC1", "PC2")) +
  scale_color_manual(values = new_group_colors) +
  theme_classic() +
  stat_ellipse()

p_newgroups
ggsave(p_newgroups, file = "export.dir/pca_rna.pdf", height = 2.5, width = 4)  


vegan::adonis2(t(transformed_counts[,metadata_no_na$UID]) ~ diagnosis + ALP + agegroup + sex, data = metadata_no_na, by = "margin")

```

Doens't seem to have much of an overall effect on the associations seen.


## Differential expression

Here I am keeping the subgrouping of patients with PSC.


```{r each vs. healthy, echo=FALSE, message=FALSE, warning=FALSE}

counts <- round(txi$counts, 0)
colnames(counts) <- names(files)

dds <- DESeqDataSetFromMatrix(counts[,metadata_no_na$UID], colData = metadata_no_na, design=~agegroup + sex + diagnosis)

dds <- dds[,rownames(metadata_no_na)]
colData(dds)$diagnosis <- relevel(colData(dds)$diagnosis, ref = "HCLow")


# filter for genes
dds <- dds[rownames(transformed_counts),] 
dds <- DESeq(dds)

# get the contrasts
contrasts <- resultsNames(dds)

# get all of the results into a single data frame
results <- list()
for (i in contrasts){
  if (i == "Intercept"){
    next}
  if (i == "ALP"){
    next}
  else{
  contrast <- unlist(strsplit(i, "_"))
  contrast <- c(contrast[1], contrast[2], contrast[4])}
  print(contrast)
  res <- lfcShrink(dds,
                   contrast=contrast,
                   res=results(dds, contrast=contrast),
                   type="normal")
  res <- as.data.frame(res)

  res$gene_id <- rownames(res)
  res$contrast <- contrast[2]
  results[[contrast[2]]] <- res
  
}
results <- dplyr::bind_rows(results)

# add gene name to results
results$gene_name <- geneid2genename[results$gene_id,]$gene_name 

# write the results
write.table(results, file = "export.dir/results_all.tsv",
            row.names=FALSE,
            quote=FALSE,
            sep="\t")


# Aggregate expression over genes
genes_to_agg <- geneid2genename[rownames(transformed_counts),]$gene_name
transformed_counts_sym <- aggregate(transformed_counts, by = list(genes_to_agg), FUN = "mean")

rownames(transformed_counts_sym) <- transformed_counts_sym$Group.1
transformed_counts_sym <- transformed_counts_sym[,2:ncol(transformed_counts_sym)]

# write out the transformed counts - will be useful for
# comparing to the proteomics data
write.table(data.frame(gene = rownames(transformed_counts_sym),
                       transformed_counts_sym),
            file = "export.dir/transformed_counts_symbols.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)


```

## Fgsea of GSTA1-associated genes

```{r fgsea gsta1, echo = FALSE, message = FALSE}

gsta1_genes <- list(GSTA1_Module = c("GSTA1", "MT1H", "MT1M", "HLA-B", "MT1G", "MT1F", "ACTB", "CCL25", "TMSB10", "FABP1", "FABP6", "KRT8", "LCN2", "REG1A", "FAM151A", "PIGR", "RBP2"))

sub_dat <- subset(results, subset = contrast == "PSCUCHigh")
input_genes <- sub_dat$log2FoldChange
names(input_genes) <- sub_dat$gene_name 

fgsea_res_gsta <- fgsea(gsta1_genes, input_genes)


```



## Results for genes of interest

These are the results for the genes of interest i.e. GSTA-associated genes that are downregulated in PSC enterocytes.

```{r gois results, echo = FALSE, message = FALSE}

gois_results <- as.data.frame(subset(results, subset = gene_name %in% gsta1_genes[["GSTA1_Module"]]))

knitr::kable(subset(gois_results, subset = contrast == "PSCUCHigh"))

```

GSTA1 and GGT1 show decreased expression. Only focus on these - these are more explicitly detoxification genes so are of interest.

## Plotting some genes of interest

```{r gois, echo = FALSE, message = FALSE}

gois <- c("GSTA1",  "MT1H", "MT1F")

detox_genes <- featurebox(transformed_counts_sym[,metadata_no_na$UID],
               metadata_no_na,
               group_by = "diagnosis",
               features = gois) +
  facet_wrap(~feature, scales = "free") +
  theme_classic() +
  scale_color_manual(values = new_group_colors) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45)) +
  theme(legend.position = "none")

detox_genes
ggsave(detox_genes, file = "export.dir/rna_gois_featurebox.pdf", height = 2.5, width = 5)
```


## Gene set enrichment analysis

This is a fgsea of ALP associated genes - ranking by fold change i.e. the differential expression analysis of the PSC-UC; ALP High group compared to Healthy controls.

```{r fgsea, echo = FALSE, message = FALSE}

res_alp <- subset(results, subset = contrast == "PSCUCHigh")
genes <- res_alp$log2FoldChange
names(genes) <- res_alp$gene_name

genes <- genes[!is.na(names(genes))]
genes <- genes[is.finite(genes)]

pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")

#detox <- pathways[names(pathways) == "GO_DETOXIFICATION"]

fgsea_res <- fgsea(pathways, genes, minSize = 20, maxSize = 200)
fgsea_res <- fgsea_res[order(fgsea_res$padj, decreasing = FALSE),]

ggplot(fgsea_res[1:10,], aes(x = pathway, y = NES)) +
  geom_bar(stat = "identity", color = "black", fill = "snow2") +
  theme_classic() +
  coord_flip()

```

Mainly Ig related genes.

## Have a look at some of the interesting pathways

```{r fgsea results genes, echo = FALSE, message = FALSE}

OCMSutility::featurebox(transformed_counts_sym[,metadata_no_na$UID],
                        metadata_no_na,
                        features = c("IGHV3-43", "IGHV1-24", "IGHV2-26", "IGHV1-69D"),
                        group_by = "diagnosis") +
  facet_wrap(~feature, scales = "free") +
  scale_color_manual(values = new_group_colors) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## PCA with the new diagnosis split

Here we look at the PCA with the ALP high and Low groupings to see if there is any additional insight from that.

```{r pca new, echo = FALSE, message = FALSE}

pc_new <- runPCA(transformed_counts_sym[,metadata_no_na$UID])

plotPrincipleComponents(pc_new,
                        metadata_no_na,
                        colourby = "diagnosis",
                        group = "diagnosis") +
  theme_classic() +
  stat_ellipse() +
  scale_color_manual(values = group_colors)
  
set.seed(10)
vegan::adonis2(t(transformed_counts_sym[,metadata_no_na$UID]) ~ diagnosis + agegroup + sex,
               data = metadata_no_na,
               by = "margin",
               na.action = na.omit,
               method = "euclidean")

# write out the metadata - can use for the metabolomics analysis
write.table(metadata_no_na,
            file = "export.dir/metadata_alp_combined.tsv",
            sep = "\t",
            row.names = FALSE,
            quote = FALSE)

```







