---
title: "Caecum epithelial analysis: Forced cells 10000"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}


# temporary solution to libpaths
.libPaths("~/R/x86_64-pc-linux-gnu-library/4.3/")

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args=list(png=list(type="cairo")))
options(bitmapType = "cairo")

# reticulate stuff
reticulate::use_virtualenv('~/devel/venv/Python-3.10.8-GCCcore-12.2.0/skylake/', required=TRUE)
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")
```

```{r load libraries, echo=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(speckle)
library(scWGCNA)
library(circlize)
library(ggraph)
library(igraph)
library(DESeq2)


# create Clustering analysis dir
if (!(dir.exists("export.dir/Epithelium"))){
  dir.create("export.dir/Epithelium")
}
```


```{r set colors, echo=FALSE, message=FALSE, warning=FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```

# Overview

We will need to organise data analysis by cell types. There are some that will not have enough cells to perform pseudobulk for example but we can determine this as we go along. In this doc I am going to look at the epithelial compartment.

```{r read data merged caecum, echo=FALSE, message=FALSE}

merged_caecum <- readRDS("export.dir/Clustering/caecum_harmonised_force_cells_annotated.Rds")

# read in metadata
metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)
```

```{r subset for epithelial cells, echo=FALSE, message=FALSE}

epithelium <- subset(merged_caecum, idents = c("Colonocyte",
                                              "Colonocyte_2",
                                              "Colonocyte_BEST4_OTOP2",
                                              "Tuft_Cell",
                                              "Goblet_Contaminated[B_Cell]"))




```

## Re-clustering of epithelial cells {.tabset .tabset-pills}

Here we recluster just the epithelial cells.

```{r recluster epithelium, echo=FALSE, message=FALSE}

set.seed(10)

# recluster etc
epithelium <- ScaleData(epithelium, features = rownames(epithelium))
epithelium <- FindVariableFeatures(epithelium, selection.method = "vst", nfeatures = 2000)
epithelium <- RunPCA(epithelium, features = VariableFeatures(object = epithelium))

# harmony integration: by individual = orig.ident
epithelium <- epithelium %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
epithelium <- FindNeighbors(epithelium, reduction="harmony", dims = 1:30)
epithelium <- FindClusters(epithelium, resolution = 0.5, algorithm = 4)
epithelium <- RunUMAP(epithelium, dims = 1:30, reduction = "harmony")
```

### UMAP

```{r epitheilum umap, echo=FALSE, message=FALSE}

DimPlot(epithelium, label=TRUE, reduction = "umap")

```

### UMAP by sample

```{r epithelium umamp by sample, echo = FALSE, message = FALSE}

DimPlot(epithelium, label=TRUE, reduction = "umap", split.by = "orig.ident")

```

### G-SCI7T094-SCI5T094 - caecum that looks like ileum

One caecum sample looks a bit like ileum in that it has enterocytes i.e. high SLC10A2, FABP6 cells. Here we investigate this sample in a bit more detail.


```{r G-SCI7T094-SCI5T094, echo = FALSE, message = FALSE}

# read in the ileum data as well
ileum_epi <- readRDS("export.dir/Epithelium/ileum_epithelium_clusters_annotated.Rds")

SCI5T082 <- subset(ileum_epi, subset = orig.ident == "G-SCI7T082-SCI5T082")
Idents(SCI5T082) <- SCI5T082@meta.data$orig.ident

SCI5T094 <- subset(epithelium, subset = orig.ident == "G-SCI7T094-SCI5T094")
Idents(SCI5T094) <- SCI5T094@meta.data$orig.ident

# an actual caecum
SCI5T046 <- subset(epithelium, subset = orig.ident == "G-SCI7T046-SCI5T046")
Idents(SCI5T046) <- SCI5T046@meta.data$orig.ident

gi9593 <- merge(merge(SCI5T082, SCI5T094), SCI5T046)

set.seed(10)

# recluster etc
gi9593 <- ScaleData(gi9593, features = rownames(gi9593))
gi9593 <- FindVariableFeatures(gi9593, selection.method = "vst", nfeatures = 2000)
gi9593 <- RunPCA(gi9593, features = VariableFeatures(object = gi9593))

gi9593 <- gi9593 %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)

gi9593 <- FindNeighbors(gi9593, reduction="harmony", dims = 1:30)
gi9593 <- FindClusters(gi9593, resolution = 0.5, algorithm = 4)
gi9593 <- RunUMAP(gi9593, dims = 1:30, reduction = "harmony")

gi9593_dplot <- DimPlot(gi9593, label = TRUE, split.by = "orig.ident")

FeaturePlot(gi9593, features = c("SLC10A2", "FABP6", "CA1"), split.by = "orig.ident")

```


### Dotplot of markers

```{r dotplot epithelium, echo=FALSE, message=FALSE, width = 11}

# find markers for exploration
epithelium <- JoinLayers(epithelium)
epi_markers <- FindAllMarkers(epithelium, only.pos = TRUE)

epithelial_cell_genes <- c("EPCAM",
                           "VIL1",  
                          "FABP6",
                          "FABP1",
                          "CEACAM7",
                          "CEACAM1",
                          "AQP8",
                          "KRT19",
                          "KRT8",
                          "APOA1",
                          "APOB",
                          "DEFA5",
                          "DEFA6",
                          "REG3A",
                          "SLC10A2",
                          "GSTA2",
                          "GSTA1",
                          "DUOX2",
                          "DUOXA2",
                          "TPT1",
                          "LGR5",
                          "OLFM4",
                          "RGMB",
                          "SMOC2",
                          "ASCL2",
                          "MKI67",
                          "PCNA",
                          "TOP2A",
                          "CCNA2",
                          "POU2F3",
                          "GFI1B",
                          "TRPM5",
                          "DCLK1",
                          "CLCA1",
                          "MUC5AC",
                          "SPINK4",
                          "TFF3",
                          "AGR2",
                          "SPARC",
                          "CPE",
                          "NEUROD1",
                          "BEST4",
                          "OTOP2")

dplot_epi <- DotPlot(epithelium,
             cols = colorRampPalette(c("blue", "red"))(2),
             features=c(epithelial_cell_genes),
             cluster.idents = TRUE) +
             theme(axis.text.x = element_text(angle=45, hjust=1)) 

dplot_epi

```

## Score cell cycle

```{r cell cycle, echo=FALSE, message=FALSE}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

epithelium <- CellCycleScoring(epithelium, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

DimPlot(epithelium, label=TRUE, split.by="disease_type")
Idents(epithelium) <- epithelium@meta.data$seurat_clusters
```

Proliferating cells will be the TA so will not regress out actually.


## Annotation

The main thing here is to pick out the most relevant populations. If there are some that seem sketchy then they will be removed and we will only keep those that have robust markers associated with them. There are no really good markers that distinguish cluster 3.

```{r annotation epithelium, echo=FALSE, message=FALSE}

# annotations
annotations_epi <- c("Colonocyte",
                     "Colonocyte",
                     "Colonocyte_Enterocyte",
                     "Colonocyte_AQP8",
                     "Tuft_Cell",
                     "Colonocyte_BEST4_OTOP2",
                     "Goblet_Cell",
                     "Colonocyte_DUOX2",
                     "Contaminated",
                     "Contaminated")
                                          

names(annotations_epi) <- levels(epithelium)
epithelium <- RenameIdents(epithelium, annotations_epi)


# remove contamination clusters
epithelium <- subset(epithelium, idents = c("Contaminated"),
                                            invert=TRUE)

# remove dodgy sample that has the enterocyte high proportion. Should remove this from the
# immune cell analysis as well just in case it is a weirdness thing.
epithelium <- subset(epithelium, subset = orig.ident == "G-SCI7T094-SCI5T094", invert = TRUE)

# Remove remaining contams
epithelium <- subset(epithelium, idents = c("Colonocyte_Enterocyte"),
                                            invert=TRUE)

annotation_colors_epithelium <- OCMSutility::get_palette(n = 50)[1:7]
#annotation_colors_epithelium <- annotation_colors_epithelium[c(2, 1, 3:20)]

names(annotation_colors_epithelium) <- levels(Idents(epithelium))

# save the colors
saveRDS(annotation_colors_epithelium, file="export.dir/Clustering/epithelium_cell_colors_caecum.Rds")

dimplot_epi <- DimPlot(epithelium, label = TRUE, repel = TRUE) + 
  scale_color_manual(values = annotation_colors_epithelium) +
  ggtitle(paste0("No. cells = ", nrow(epithelium@meta.data)))

dimplot_epi

# write out epithelial annotated clusters

# Figure 1D
ggsave(dimplot_epi + theme(legend.position = "none"), file = "export.dir/Figures/Epithelium_umap_caecum.pdf", height=5, width=5)

saveRDS(epithelium, file = "export.dir/Clustering/caecum_epithelium_clusters_annotated.Rds")

```


### Profile of epithelium subsets across individuals

This is just to show the profiles of these cell types across individuals.

```{epithelial proportions, echo=FALSE, message=FALSE}


ind_props_epi <- plot_props(epithelium, metadata, annotation_colors_epithelium) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

ggsave(ind_props_epi, file = "export.dir/Figures/epithelial_cells_profile_caecum.pdf", height=3, width = 6)

ind_props_epi

```

## Number of cells per cell type

```{r epithelium ncells, echo = FALSE, message=FALSE}

epithelium <- readRDS("export.dir/Clustering/caecum_epithelium_clusters_annotated.Rds")

epithelium@meta.data$annotation <- Idents(epithelium)

ncells_per_sample_epithelium <- as.data.frame(table(epithelium@meta.data$annotation,
                                                epithelium@meta.data$orig.ident))
knitr::kable(ncells_per_sample_epithelium)

# only analyse when there are at least 20 cells per indivudual
ncells_per_sample_epithelium$disease_type <- metadata[as.character(ncells_per_sample_epithelium$Var2), ]$disease_type
ggplot(ncells_per_sample_epithelium, aes(x=disease_type, y=Freq, color=disease_type)) +
  geom_point() +
  theme_classic() +
  facet_wrap(~Var1, scales = "free") +
  geom_hline(yintercept = 20,  linetype="dashed") +
  theme(axis.text.x = element_blank()) +
  scale_color_manual(values = group_colors) +
 # ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  ggrepel::geom_text_repel(label = ncells_per_sample_epithelium$Freq)

```


## Total proportion of each cell type

```{r totals, echo=FALSE, message=FALSE}

total <- nrow(epithelium@meta.data)

enterocyte_prop <- (nrow(epithelium@meta.data[grep("Colonocyte", Idents(epithelium)),])/total)*100

enterocyte_prop
```


### Testing for siginficant differences

Here I use propeller to look for significant differences across disease states in terms of proportions of cell types.

```{r significant differences, echo=FALSE, message=FALSE}

# significant testing using propeller
prop_list_epi <- getTransformedProps(clusters = Idents(epithelium),
                                 sample = epithelium@meta.data$orig.ident)

metadata_caecum <- metadata[metadata$tissue == "Caecum",]
metadata_caecum <- metadata_caecum[colnames(prop_list_epi[[1]]),]

design <- model.matrix(~0+disease_type, data = metadata_caecum)

res_comp <- propeller.anova(prop_list_epi, design=design, coef=c(1,2,3), robust=TRUE,
  trend=FALSE, sort=TRUE)

knitr::kable(res_comp)
```


There are no signficant differences found between diseases.

## Differences between Colonocyte subsets




### Marker expression across cells

The DUOX2 colonocytes are odd ones. Ignore those for now

```{r featureplots markers, echo=FALSE, message=FALSE}

colonocytes <- subset(epithelium, idents = c("Colonocyte_AQP8", "Colonocyte"))

colonocytes <- ScaleData(colonocytes)
colonocytes <- RunPCA(colonocytes)
```

### Dimplot of colonocytes alone

```{r colonocyte dimplot, echo=FALSE, message=FALSE}

dplot_colonocytes <- DimPlot(colonocytes, reduction = "pca", label=TRUE) +
  scale_color_manual(values = annotation_colors_epithelium) +
  theme(legend.position = "none")

dplot_colonocytes

ggsave(dplot_enterocytes, file="export.dir/Figures/Dimplot_colonocytes_caecum.pdf", height=5, width=5)

```


### Differences between the major colonocyte subsets

Do they represent different cell "types" or cell "states"?

```{r diff in enterocytes, echo=FALSE, message=FALSE}

colonocyte_diffs <- FindMarkers(colonocytes, ident.2="Colonocyte_AQP8",
                                ident.1 = "Colonocyte")

# Annotate the results
colonocyte_diffs$sig <- ifelse(colonocyte_diffs$p_val_adj < 0.05 &
                              colonocyte_diffs$avg_log2FC > 1, 
                              "Colonocyte", 
                               NA)

colonocyte_diffs$sig <- ifelse(colonocyte_diffs$p_val_adj < 0.05 &
                              colonocyte_diffs$avg_log2FC < 1, 
                               "AQP8", 
                               colonocyte_diffs$sig)

colonocyte_diffs$label <- ifelse(rownames(colonocyte_diffs) %in% 
                                   c("Colonocyte", "AQP8"),
                                 rownames(colonocyte_diffs), NA) 

ggplot(colonocyte_diffs, aes(x=avg_log2FC, y = -log10(p_val_adj), color=sig)) +
  geom_point(alpha=0.5, size=0.5) +
  theme_classic() +
  scale_color_manual(values = c(annotation_colors_epithelium[[1]],
                                annotation_colors_epithelium[[2]])) +
  geom_vline(xintercept = c(-1, 1), linetype="dashed") +
  geom_vline(xintercept = 0) +
  ggrepel::geom_text_repel(label = colonocyte_diffs$label)

```


```{r plotting features, echo=FALSE, message=FALSE}

faqp8 <- feature_plot(colonocytes, feature = "AQP8", high=annotation_colors_epithelium[[2]]) 
  
faqp8

```



## Differential expression - pseudobulk

Here we perform differential expression analysis on pseudobulk expression across cells in cell types that have enough cells (typically >20/individual).

### Colonocytes


```{r pseudobulk colonocytes, echo=FALSE, message=FALSE}

# table the number of cells per invididual
colonocytes <- subset(epithelium, idents = c("Colonocyte", "Colonocyte_AQP8"))
colonocytes@meta.data$annotation <- as.character(Idents(colonocytes))
Idents(colonocytes) <- colonocytes@meta.data$orig.ident
ncells_colono <- as.data.frame(table(Idents(colonocytes)))
knitr::kable(ncells_colono)

# remove troublesome samples
colonocytes <- subset(colonocytes, idents = c("G-SCI7T022-SCI5T022"), invert=TRUE)

Idents(colonocytes) <- colonocytes@meta.data$annotation

# Write out the colonocytes object so we can compare
# with ileal enterocytes
saveRDS(colonocytes, file = "export.dir/caecum_colonocytes.Rds")

# aggregate for pseudobulk analysis
agg_all <- AggregateExpression(colonocytes,
                               group.by = c("orig.ident"))                                                             

agg_all <- as.data.frame(agg_all)
colnames(agg_all) <- gsub("RNA\\.", "", colnames(agg_all))
rownames(metadata) <- gsub("-", "\\.", rownames(metadata))
coldata <- metadata[colnames(agg_all),]
dds <- DESeqDataSetFromMatrix(agg_all, colData = coldata, design=~disease_type)
dds$disease_type <- relevel(dds$disease_type, ref = "HC")
dds <- dds[rowSums(counts(dds) >= 10) >= 3,]
dds <- DESeq2::DESeq(dds)

# get results for each contrast
# get results for each contrast
# lfc shrink
res1 <- lfcShrink(dds,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2 <- lfcShrink(dds,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds, name = "disease_type_UC_vs_HC"),
                  type = "normal")

```

#### Volcano plot

```{psc enterocytes volcano, echo=FALSE, message=FALSE}

#res1_toplot <- res1
#res1$log2FoldChange[res1$log2FoldChange < -5] <- -5

psc_colonocyte_volc <- volcano_plot(res1, use="padj") +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = "blue3") +
  theme(legend.position = "none")

psc_colonocyte_volc

# Figure
ggsave(psc_colonocyte_volc, file = "export.dir/Figures/colonocyte_caecum_psc_volcano.pdf", height = 3, width=3)

uc_colonocyte_volc <- volcano_plot(res2, use="padj") +
  ggtitle("UC vs. Healthy") +
  scale_color_manual(values = "blue3") +
  theme(legend.position = "none")

ggsave(uc_colonocyte_volc, file = "export.dir/Figures/colonocyte_caecum_uc_volcano.pdf", height = 5, width=5)

uc_colonocyte_volc

# save the results to compare with ileal enterocytes
write.table(data.frame(gene = rownames(res1), as.data.frame(res1)),
            file = "export.dir/caecum_colonocytes_pscuc_vs_healthy.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)




```

```{r table of differences enterocytes, echo=FALSE, message=FALSE}

nup_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange > 0 & !(is.na(res1$padj)),])
nup_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange > 0 & !(is.na(res2$padj)),])
ndown_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange < 0 & !(is.na(res1$padj)),])
ndown_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange < 0 & !(is.na(res2$padj)),])

results_ent <- data.frame(comparison = resultsNames(dds)[2:3],
                          no.up = c(nup_psc, nup_uc),
                          no.down = c(ndown_psc, ndown_uc))

knitr::kable(results_ent)
```

## Gene set enrichment analysis


```{r colonocyte fgsea, echo=FALSE, message=FALSE}

set.seed(10)

colonocyte_genes <- res1$log2FoldChange
names(colonocyte_genes) <- rownames(res1)

# gene set enrichments
colonocyte_fgsea <- fgsea::fgsea(pathways, colonocyte_genes, minSize=50)
colonocyte_fgsea <- colonocyte_fgsea[order(colonocyte_fgsea$padj, decreasing = FALSE),]

colonocyte_pathways_sig <- colonocyte_fgsea[colonocyte_fgsea$padj < 0.05,]

colonocyte_top10 <- colonocyte_pathways_sig[1:10,]
colonocyte_top10$pathway <- factor(colonocyte_top10$pathway, levels = colonocyte_top10$pathway)
colonocyte_top10_p <- ggplot(colonocyte_top10, aes(x = pathway, y=NES, fill=-log10(padj))) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low="snow2", high=immune_cell_anno_colors[[2]]) +
  theme_classic() +
  coord_flip()
  
ggsave(colonocyte_top10_p, file = "export.dir/Figures/colonocyte_pathway_barplot_caecum.pdf", height=2.5, width=10)

colonocyte_top10_p

```

## Leading edge heatmap

```{r protein metabolism heat, echo=FALSE, message=FALSE}

vst_all <- as.data.frame(assay(vst(dds)))

peptide_le <- unlist(colonocyte_fgsea[colonocyte_fgsea$pathway == "GO_PEPTIDE_METABOLIC_PROCESS",]$leadingEdge)

# Heatmap these genes in different T-cells
vst_colonocyte_peptide <- ScaleData(vst_all[peptide_le,])

# sort out annotations for the disease states
disease_anno <- HeatmapAnnotation("Disease type" = coldata$disease_type,
                                  col = list("Disease type" = c("HC" = "grey",
                                                                "PSCUC" = blues9[6],
                                                                "UC" = "red3")))

heat_colors <- colorRamp2(breaks = c(-2,0,2), colors = c("blue3", "white", "red3"))
h1 <- Heatmap(vst_colonocyte_peptide,
              column_split = coldata$disease_type,
              col = heat_colors,
              cluster_columns = FALSE,
              top_annotation = disease_anno,
              show_column_names = FALSE,
              rect_gp = gpar(col="white", lwd=0.5),
              row_names_gp = gpar(fontsize=4))

h1

```




