---
title: "Enterocytes/colonocytes Cells Compare Differences"
output: html_document
date: "2026-02-17"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")
```

```{r load libraries, echo = FALSE, message = FALSE}

library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(speckle)
library(scWGCNA)
library(circlize)
library(ggraph)
library(igraph)
library(DESeq2)
library(OCMSutility)

```

# Overview

We have seen some interesting results particularly in the ileum. We also ahve data from the caecum and so we need to compare what we see across the two tissues. Given the tissue-specific functions of mature absorptive epithelial cells in these two tissues it is unlikely that the same features are observed but here we characterise specific pathways to see if we can say something about GSTA1-associated genes etc.

## Comparing cell types

```{r compare cell types, echo = FALSE, message = FALSE}

ileum_entero <- readRDS("export.dir/ileum_enterocytes.Rds")
Idents(ileum_entero) <- "Ileum"
caecum_colono <- readRDS("export.dir/caecum_colonocytes.Rds")
Idents(caecum_colono) <- "Caecum"

all_colono_entero <- merge(ileum_entero, caecum_colono)

# recluster etc
all_colono_entero <- ScaleData(all_colono_entero,
                               features = rownames(all_colono_entero))

all_colono_entero <- FindVariableFeatures(all_colono_entero,
                                          selection.method = "vst",
                                          nfeatures = 2000)

all_colono_entero <- RunPCA(all_colono_entero,
                            features = VariableFeatures(object = all_colono_entero))

# harmony integration: by individual = orig.ident
all_colono_entero <- all_colono_entero %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
all_colono_entero<- FindNeighbors(all_colono_entero,
                                  reduction="harmony",
                                  dims = 1:30)
all_colono_entero <- RunUMAP(all_colono_entero,
                             dims = 1:30,
                             reduction = "harmony")

tissue_colors <- c(c(blues9[4], blues9[8]))
names(tissue_colors) <- c("Ileum", "Caecum")

dplot_compare <- DimPlot(all_colono_entero, reduction = "pca") +
  scale_color_manual(values = tissue_colors) 
  
dplot_compare

ggsave(dplot_compare + theme(legend.position = "none"),
       file = "export.dir/Comparison/epithelium_tissue_compare_pca.pdf",
       height = 5, width = 5)


```



## GSTA1 and APOA1 genes

This shows the differences in expression of GSTA1-associated genes - we assess whether these are tissue-sepcific genes in some respects.

```{r tissue specific genes, echo = FALSE, message = FALSE}

agg_ents <- as.data.frame(AggregateExpression(all_colono_entero, group.by = "orig.ident"))

colnames(agg_ents) <- gsub("RNA\\.", "", colnames(agg_ents))

all_colono_entero$tissue <- Idents(all_colono_entero)

agg_meta <- distinct(data.frame(sample = all_colono_entero@meta.data$orig.ident,
                                tissue = all_colono_entero@meta.data$tissue,
                                disease_type = all_colono_entero@meta.data$disease_type))


rownames(agg_meta) <- gsub("-", "\\.", agg_meta$sample)
agg_meta <- agg_meta[colnames(agg_ents),]

## DESeq analysis between tissues
agg_meta <- agg_meta[colnames(agg_ents),]
dds_agg <- DESeqDataSetFromMatrix(agg_ents,
                                  colData = agg_meta,
                                  design = ~tissue)
dds_agg <- DESeq(dds_agg)
agg_res <- lfcShrink(dds_agg,
                     contrast = c("tissue", "Ileum", "Caecum"),
                     res = results(dds_agg),
                     type = "normal")

gsta1_genes <- readRDS("export.dir/gsta1_cluster.Rds")
gsta1_genes <- rownames(gsta1_genes[gsta1_genes$Membership > 0.6, ])

apoa1_genes <- readRDS("export.dir/apoa1_cluster.Rds")
apoa1_genes <- rownames(apoa1_genes[apoa1_genes$Membership > 0.6, ])


agg_vst <- as.data.frame(vst(as.matrix(agg_ents)))

agg_res <- as.data.frame(agg_res)
agg_res <- agg_res[order(agg_res$padj, decreasing = FALSE),]

sig_up_ileum <- rownames(agg_res[agg_res$padj < 0.05 & agg_res$log2FoldChange > 1,])


# change volcano plot function
volcano_plot_2 <- function(x, use="padj", value=0.05, labels = NULL){
  
  if (is.null(labels)){
    x$label <- ifelse(x[,use] < value &!(is.na(x[,use])), rownames(x), NA)
  }else{
    x$label <- ifelse(rownames(x) %in% labels, rownames(x), NA)
  }
  
  x$color <- ifelse(x[,use] < value &!(is.na(x[,use])) & x$log2FoldChange > 1, "Ileum", NA)
  x$color <- ifelse(x[,use] < value &!(is.na(x[,use])) & x$log2FoldChange < -1, "Caecum", x$color)
  ggplot(x, aes(x=log2FoldChange, y = -log10(pvalue), color=color)) +
    geom_point(size = 1, alpha = 0.5) +
    geom_vline(xintercept=c(-1,1), linetype = "dashed") +
    ggrepel::geom_text_repel(label = x$label, max.overlaps=20, color = "black") +
    theme_classic() +
    scale_color_manual(values = c(tissue_colors)) 
}



gsta1_genes_sig <- gsta1_genes[gsta1_genes %in% sig_up_ileum]
apoa1_genes_sig <- apoa1_genes[apoa1_genes %in% sig_up_ileum]

vplot <- volcano_plot_2(agg_res, labels = gsta1_genes_sig) +
  theme(legend.position = "none")

ggsave(vplot, file = "export.dir/Comparison/epithelium_volcano_caecum_ileum.pdf", height = 5, width = 5)

vplot

featurebox(agg_vst, agg_meta, features = gsta1_genes_sig, group_by = "tissue") +
  theme_classic() +
  scale_color_manual(values = tissue_colors) +
  facet_wrap(~feature, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_compare_means(comparisons = list(c(1,2)), size = 2)

featurebox(agg_vst, agg_meta, features = apoa1_genes_sig, group_by = "tissue") +
  theme_classic() +
  scale_color_manual(values = tissue_colors) +
  facet_wrap(~feature, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_compare_means(comparisons = list(c(1,2)), size = 2)
```


Basically I think this is quite revealing and shows that `r length(gsta1_genes_sig)/length(gsta1_genes)` GSTA1-associated genes and `r length(apoa1_genes_sig)/length(apoa1_genes)` APOA1-associated genes are more highly expressed in the ileum than the caecum. This suggests that differences seen in PSC-UC in GSTA1-associated genes in the ileum are indeed to do with more ileum-associated functions i.e. detoxification in ileal enterocytes.


## Feature plots of GSTA1 and associated genes

I might include these in a figure.


```{r features, echo = FALSE, message = FALSE}

gsta1_featureplots <- FeaturePlot(all_colono_entero, features = gsta1_genes_sig, reduction = "pca") 
ggsave(file = "export.dir/Comparison/gsta1_across_tissues.pdf", width = 10, height = 15)

```


## Adonis test

Here I just want to see if there is a statistical association with tissue site ie. so cells cluster by tissue. The adonis test can do this for me.

```{r adonis test, echo = FALSE, message = FALSE}

vegan::adonis2(t(agg_ents) ~ tissue, method = "euclidean", data = agg_meta)


```



## Metagene signatures

To summarise the differences we can look at a metagene signature i.e. just a mean across all genes for each sample. This will enable a more reduced plotting landscape.

```{r metagenes, echo = FALSE, message = FALSE, warning = FALSE}

metagene_gsta1 <- agg_vst[gsta1_genes,]
metagene_gsta1 <- colMeans(metagene_gsta1)

metagene_apoa1 <- agg_vst[apoa1_genes,]
metagene_apoa1 <- na.omit(metagene_apoa1)
metagene_apoa1 <- colMeans(metagene_apoa1)

both_metagenes <- as.data.frame(t(dplyr::bind_cols(as.data.frame(metagene_gsta1),
                                   as.data.frame(metagene_apoa1))))


featurebox(agg_vst, agg_meta, features = c("ADIRF"), group_by = "disease_type") +
  theme_classic() +
  scale_color_manual(values = tissue_colors) +
  facet_wrap(~feature, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggpubr::stat_compare_means(comparisons = list(c(1,2)), size = 2) 


```

For sure these are more ileum functions.


## Comparison differential expression results

There is only one gene, ADIRF, which is signficantly lower in expression in PSC-UC in the caecum in colonocytes. Below we also plot this in the ileum enterocytes to see if it is at least similarly down-regulated.


### ADIRF

```{r adirf expression, echo = FALSE, message = FALSE}

# add ADIRF to agg_meta
agg_meta$ADIRF <- unlist(agg_vst["ADIRF",rownames(agg_meta)])
ggplot(agg_meta, aes(x = disease_type, y = ADIRF, color = disease_type)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  theme_classic() +
  facet_wrap(~tissue) +
  scale_color_manual(values = unlist(group_colors)) +
  theme(legend.position = "none")
ggsave("export.dir/ADIRF.pdf", height = 2.5, width = 4)


```



```{r comparison, echo = FALSE, message = FALSE}

pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")


ileum_res <- read.csv("export.dir/ileum_enterocytes_pscuc_vs_healthy.tsv",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep = "\t",
                      row.names = 1)

caecum_res <- read.csv("export.dir/caecum_colonocytes_pscuc_vs_healthy.tsv",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep = "\t",
                      row.names = 1)

subset(caecum_res, subset = padj < 0.05)

# merge the results tables
ileum_fold <- data.frame(gene = rownames(ileum_res),
                         fold = ileum_res$log2FoldChange)

caecum_fold <- data.frame(gene = rownames(caecum_res),
                         fold = caecum_res$log2FoldChange)

merged_ents <- merge(ileum_fold,
                     caecum_fold,
                     by.x = "gene",
                     by.y = "gene",
                     all.x = TRUE,
                     all.y = TRUE)
merged_ents[is.na(merged_ents)] <- 0
colnames(merged_ents) <- c("gene", "Ileum", "Caecum")

merged_ents$label <- ifelse(merged_ents$gene %in% gsta1_genes_sig, merged_ents$gene, NA)
merged_ents$GSTA1 <- ifelse(merged_ents$gene %in% gsta1_genes_sig, "GSTA1", NA)


colors <- readRDS("export.dir/Clustering/epithelium_cell_colors.Rds")

ggplot(merged_ents, aes(x = Ileum, y = Caecum, color = GSTA1)) +
  geom_point(size = 0.3) +
  theme_classic() +
  geom_hline(yintercept = c(-0.5,0.5, 0), linetype = "dashed") +
  geom_vline(xintercept = c(-0.5,0.5, 0), linetype = "dashed") +
  ylim(c(-3,3)) +
  ggrepel::geom_text_repel(label = merged_ents$label) +
  stat_smooth(method = "lm") +
  scale_color_manual(values = c(colors[[1]]))


```
