---
title: "Ileum epithelial analysis: Forced cells 10000"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}


# temporary solution to libpaths
.libPaths("~/R/x86_64-pc-linux-gnu-library/4.3/")

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args=list(png=list(type="cairo")))
options(bitmapType = "cairo")
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")


# reticulate stuff
reticulate::use_virtualenv('~/devel/venv/Python-3.10.8-GCCcore-12.2.0/skylake/', required=TRUE)
```

```{r load libraries, echo=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(speckle)
library(scWGCNA)
library(circlize)
library(ggraph)
library(igraph)
library(DESeq2)
library(viridis)

# create Clustering analysis dir
if (!(dir.exists("export.dir/Epithelium"))){
  dir.create("export.dir/Epithelium")
}
```


```{r set colors, echo=FALSE, message=FALSE, warning=FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```

# Overview

We will need to organise data analysis by cell types. There are some that will not have enough cells to perform pseudobulk for example but we can determine this as we go along. In this doc I am going to look at the epithelial compartment. Part of the analysis will be to understand a bit more about genes that we saw in bulk RNA-seq so am very interested in glutathione metabolism and oxidative stress responses.

```{r read data merged ileum, echo=FALSE, message=FALSE}

merged_ileum <- readRDS("export.dir/Clustering/ileum_harmonised_force_cells_annotated.Rds")

# read in metadata
metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)
```

```{r subset for epithelial cells, echo=FALSE, message=FALSE}

epithelium <- subset(merged_ileum, idents = c("Enterocyte",
                                              "Enterocyte_2",
                                              "Enterocyte_3",
                                              "Epithelial_Stem",
                                              "Tuft_Cell",
                                              "Goblet_Cell",
                                              "Paneth_Cell",
                                              "Enterocyte_BEST4_OTOP2"))


```

## Re-clustering of epithelial cells {.tabset .tabset-pills}

Here we recluster just the epithelial cells.

```{r recluster epithelium, echo=FALSE, message=FALSE}

set.seed(10)

# recluster etc
epithelium <- ScaleData(epithelium, features = rownames(epithelium))
epithelium <- FindVariableFeatures(epithelium, selection.method = "vst", nfeatures = 2000)
epithelium <- RunPCA(epithelium, features = VariableFeatures(object = epithelium))

# harmony integration: by individual = orig.ident
epithelium <- epithelium %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
epithelium <- FindNeighbors(epithelium, reduction="harmony", dims = 1:30)
epithelium <- FindClusters(epithelium, resolution = 0.5, algorithm = 4)
epithelium <- RunUMAP(epithelium, dims = 1:30, reduction = "harmony")
```

### UMAP

```{r epitheilum umap, echo=FALSE, message=FALSE}

DimPlot(epithelium, label=TRUE, reduction = "umap")

```

### Dotplot of markers

```{r dotplot epithelium, echo=FALSE, message=FALSE, width = 11}

# find markers for exploration
epithelium <- JoinLayers(epithelium)
epi_markers <- FindAllMarkers(epithelium, only.pos = TRUE)

epithelial_cell_genes <- c("VIL1",  
                          "FABP6",
                          "FABP1",
                          "KRT19",
                          "KRT8",
                          "APOA1",
                          "APOB",
                          "DEFA5",
                          "DEFA6",
                          "REG3A",
                          "SLC10A2",
                          "GSTA2",
                          "GSTA1",
                          "DUOX2",
                          "DUOXA2",
                          "TPT1",
                          "LGR5",
                          "OLFM4",
                          "RGMB",
                          "SMOC2",
                          "ASCL2",
                          "MKI67",
                          "PCNA",
                          "TOP2A",
                          "CCNA2",
                          "POU2F3",
                          "GFI1B",
                          "TRPM5",
                          "DCLK1",
                          "CLCA1",
                          "MUC5AC",
                          "SPINK4",
                          "TFF3",
                          "AGR2",
                          "SPARC",
                          "CPE",
                          "NEUROD1",
                          "BEST4",
                          "OTOP2")

dplot_epi <- DotPlot(epithelium,
             cols = colorRampPalette(c("blue", "red"))(2),
             features=c(epithelial_cell_genes),
             cluster.idents = TRUE) +
             theme(axis.text.x = element_text(angle=45, hjust=1)) 

dplot_epi

```

## Score cell cycle

```{r cell cycle, echo=FALSE, message=FALSE}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

epithelium <- CellCycleScoring(epithelium, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

DimPlot(epithelium, label=TRUE, split.by="disease_type")
Idents(epithelium) <- epithelium@meta.data$seurat_clusters
```

Proliferating cells will be the TA so will not regress out actually.


## Annotation

The main thing here is to pick out the most relevant populations. If there are some that seem sketchy then they will be removed and we will only keep those that have robust markers associated with them. There are no really good markers that distinguish cluster 3.

```{r annotation epithelium, echo=FALSE, message=FALSE}

# annotations
annotations_epi <- c("Enterocyte_GSTA1",
                     "Enterocyte_APOA1",
                     "Epithelium_OLFM4",
                     "Enterocyte_APOA1",
                     "Enterocyte_Mt",
                     "Epithelium_Flagged_CD45",
                     "Undetermined",
                     "Enterocyte_GSTA1",
                     "Tuft_Cell",
                     "Enterocyte_Flagged_Ig",
                     "Goblet_Cell",
                     "Enterocyte_BEST4_OTOP2",
                     "Undetermined",
                     "Undetermined",
                     "Paneth_Cell",
                     "Enteroendocrine")

names(annotations_epi) <- levels(epithelium)
epithelium <- RenameIdents(epithelium, annotations_epi)


# remove contamination clusters
epithelium <- subset(epithelium, idents = c("Epithelium_Flagged_CD45",
                                            "Enterocyte_Flagged_Ig",
                                            "Enterocyte_Mt",
                                            "Undetermined"),
                                            invert=TRUE)


annotation_colors_epithelium <- OCMSutility::get_palette(n = 50)[10:20]
annotation_colors_epithelium <- annotation_colors_epithelium[c(2, 1, 3:20)]

names(annotation_colors_epithelium) <- levels(Idents(epithelium))

# save the colors
saveRDS(annotation_colors_epithelium, file="export.dir/Clustering/epithelium_cell_colors.Rds")

dimplot_epi <- DimPlot(epithelium, label = TRUE, repel = TRUE) + 
  scale_color_manual(values = annotation_colors_epithelium) +
  ggtitle(paste0("No. cells = ", nrow(epithelium@meta.data)))

dimplot_epi

# For Figure 1
ggsave(dimplot_epi + theme(legend.position = "none"), file = "export.dir/Figures/Epithelium_umap.pdf", height=5, width=5)

# write out epithelial annotated clusters

saveRDS(epithelium, file = "export.dir/Epithelium/ileum_epithelium_clusters_annotated.Rds")


dplot_epi <- DotPlot(epithelium,
             cols = colorRampPalette(c("blue", "red"))(2),
             features=c(epithelial_cell_genes),
             cluster.idents = TRUE) +
             theme(axis.text.x = element_text(angle=45, hjust=1)) 

dplot_epi

ggsave(file = "export.dir/Epithelium/epithelium_dotplot.pdf", height = 3, width = 15)

```


### Profile of epithelium subsets across individuals

This is just to show the profiles of these cell types across individuals.

```{epithelial proportions, echo=FALSE, message=FALSE}

epithelium <- readRDS("export.dir/Epithelium/ileum_epithelium_clusters_annotated.Rds")
annotation_colors_epithelium <- readRDS("export.dir/Clustering/epithelium_cell_colors.Rds")

ind_props_epi <- plot_props(epithelium, metadata, annotation_colors_epithelium) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

ggsave(ind_props_epi, file = "export.dir/Figures/epithelial_cells_profile.pdf", height=3, width = 6)

ind_props_epi

```

## Number of cells per cell type

```{r epithelium ncells, echo = FALSE, message=FALSE}


epithelium@meta.data$annotation <- Idents(epithelium)

ncells_per_sample_epithelium <- as.data.frame(table(epithelium@meta.data$annotation,
                                                epithelium@meta.data$orig.ident))
knitr::kable(ncells_per_sample_epithelium)

# only analyse when there are at least 20 cells per indivudual
ncells_per_sample_epithelium$disease_type <- metadata[as.character(ncells_per_sample_epithelium$Var2), ]$disease_type
ggplot(ncells_per_sample_epithelium, aes(x=disease_type, y=Freq, color=disease_type)) +
  geom_point() +
  theme_classic() +
  facet_wrap(~Var1, scales = "free") +
  geom_hline(yintercept = 20,  linetype="dashed") +
  theme(axis.text.x = element_blank()) +
  scale_color_manual(values = group_colors) +
 # ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  ggrepel::geom_text_repel(label = ncells_per_sample_epithelium$Freq)

```


## Total proportion of each cell type

```{r totals, echo=FALSE, message=FALSE}

total <- nrow(epithelium@meta.data)

enterocyte_prop <- (nrow(epithelium@meta.data[grep("Enterocyte", Idents(epithelium)),])/total)*100

enterocyte_prop
```


### Testing for siginficant differences

Here I use propeller to look for significant differences across disease states in terms of proportions of cell types.

```{r significant differences, echo=FALSE, message=FALSE}

# significant testing using propeller
prop_list_epi <- getTransformedProps(clusters = Idents(epithelium),
                                 sample = epithelium@meta.data$orig.ident)

metadata_ileum <- metadata[metadata$tissue == "Ileum",]
metadata_ileum <- metadata_ileum[colnames(prop_list_epi[[1]]),]

design <- model.matrix(~0+disease_type, data = metadata_ileum)

res_comp <- propeller.anova(prop_list_epi, design=design, coef=c(1,2,3), robust=TRUE,
  trend=FALSE, sort=TRUE)

knitr::kable(res_comp)
```


There are no signficant differences found between diseases.

## Differences between enterocyte subsets

I'm very interested in the APOA1 vs GSTA1 high enterocyte subsets because it looks a bit like they are the same cells in different states i.e. the GSTA1 high are potentially actively removing toxins etc. We did see these different clusters in our other dataset so I am interested in knowing how widespread it is i.e. in other datasets as well as whether there i evidence of transition from one state to another. This may also be important as I have a feeling about glutathione homeostasis being important in PSC.


### Marker expression across cells

Restricting to the APOA1 and GSTA1 enterocyte subsets.

```{r featureplots markers, echo=FALSE, message=FALSE}

enterocytes <- subset(epithelium, idents = c("Enterocyte_APOA1", "Enterocyte_GSTA1"))

enterocytes <- ScaleData(enterocytes)
enterocytes <- RunPCA(enterocytes)
```

### Dimplot of enterocytes alone

```{r enterocyte dimplot, echo=FALSE, message=FALSE}

dplot_enterocytes <- DimPlot(enterocytes, reduction = "pca", label=TRUE) +
  scale_color_manual(values = annotation_colors_epithelium) +
  theme(legend.position = "none")

dplot_enterocytes

ggsave(dplot_enterocytes, file="export.dir/Figures/Dimplot_enterocytes.pdf", height=5, width=5)

```


### Differences between the major enterocyte subsets

I'm very interested in the GSTA1 hi enterocytes - ew can get a first handle on how different they are from APOA1 enterocytes by simply looking at differential gene expression between the groups. Do they represent different cell "types" or cell "states"?

```{r diff in enterocytes, echo=FALSE, message=FALSE}

enterocyte_diffs <- FindMarkers(enterocytes, ident.2="Enterocyte_APOA1",
                                ident.1 = "Enterocyte_GSTA1")

# Annotate the results
enterocyte_diffs$sig <- ifelse(enterocyte_diffs$p_val_adj < 0.05 &
                                 enterocyte_diffs$avg_log2FC > 1, 
                               "GSTA1", 
                               NA)

enterocyte_diffs$sig <- ifelse(enterocyte_diffs$p_val_adj < 0.05 &
                                 enterocyte_diffs$avg_log2FC < 1, 
                               "APOA1", 
                               enterocyte_diffs$sig)

enterocyte_diffs$label <- ifelse(rownames(enterocyte_diffs) %in% 
                                   c("GSTA1", "APOA1"),
                                 rownames(enterocyte_diffs), NA) 

ggplot(enterocyte_diffs, aes(x=avg_log2FC, y = -log10(p_val_adj), color=sig)) +
  geom_point(alpha=0.5, size=0.5) +
  theme_classic() +
  scale_color_manual(values = c(annotation_colors_epithelium[[2]],
                                annotation_colors_epithelium[[1]])) +
  geom_vline(xintercept = c(-1, 1), linetype="dashed") +
  geom_vline(xintercept = 0) +
  ggrepel::geom_text_repel(label = enterocyte_diffs$label)

```


```{r plotting features, echo=FALSE, message=FALSE}

fgsta1 <- feature_plot(enterocytes, feature = "GSTA1", high=annotation_colors_epithelium[[1]]) 
  
fapoa1 <- feature_plot(enterocytes, feature = "APOA1", high=annotation_colors_epithelium[[2]]) 

gridExtra::grid.arrange(fgsta1, fapoa1, nrow=1)


# Figure 2
ggsave(fgsta1, file="export.dir/Figures/gsta1_featureplot.pdf", height=2, width=3)
ggsave(fapoa1, file="export.dir/Figures/apoa1_featureplot.pdf", height=2, width=3)

```



### Public datasets

To check that this isn't a quirk of our dataset I look at the paper that was published on healthy ileal epithelial cells (GSE125970).


```{r public ileum data, echo=FALSE, message = FALSE}

set.seed(10)

public_data <- read.csv("~/work/proj039/analysis/GSE125970/GSE125970_raw_UMIcounts.txt",
                        sep="\t",
                        row.names=1)

# create metadata from sample names
tissue <- unlist(strsplit(colnames(public_data), "_"))
tissue <- tissue[seq(2,length(tissue),3)]
tissuex <- gsub("\\.[0-9]", "", tissue)

#individual <- unlist(strsplit(colnames(public_data), "_"))
#$individual <- individual[seq(2,length(cell.type),3)]
#individual <- gsub("(.*\\.)", "", individual) 
  
cell.type <- unlist(strsplit(colnames(public_data), "_"))
cell.type <- cell.type[seq(3,length(cell.type),3)]

meta_pub <- data.frame(tissue=tissuex, cell.type=cell.type, donor = tissue)
rownames(meta_pub) <- colnames(public_data)

ileum_data <- public_data[,meta_pub$tissue == "Ileum"]
meta_pub <- meta_pub[meta_pub$tissue == "Ileum",]
meta_pub <- meta_pub[colnames(ileum_data),]

# make seurat object
public_sobj <- CreateSeuratObject(ileum_data, meta.data = meta_pub)
public_sobj <- NormalizeData(public_sobj)

Idents(public_sobj) <- public_sobj@meta.data$cell.type

# jsut take enterocytes
public_sobj <- subset(public_sobj, idents = "Enterocyte")

# recluster etc
public_sobj <- ScaleData(public_sobj, features = rownames(public_sobj))
public_sobj <- FindVariableFeatures(public_sobj, selection.method = "vst", nfeatures = 2000)
public_sobj <- RunPCA(public_sobj, features = VariableFeatures(object = public_sobj))

public_sobj <- public_sobj %>% 
RunHarmony("donor", plot_convergence = TRUE)

public_sobj <- RunUMAP(public_sobj, dims = 1:30, reduction="harmony")
DimPlot(public_sobj, reduction = "umap", label = TRUE) 
  
fgsta1_pub <- feature_plot(public_sobj, feature = "GSTA1", high=annotation_colors_epithelium[[1]]) 
fapoa1_pub <- feature_plot(public_sobj, feature = "APOA1", high=annotation_colors_epithelium[[2]]) 

gridExtra::grid.arrange(fgsta1_pub, fapoa1_pub, nrow=1)

# Figure 2
ggsave(fgsta1_pub, file="export.dir/Figures/gsta1_pub_featureplot.pdf", height=2, width=3)
ggsave(fapoa1_pub, file="export.dir/Figures/apoa1_pub_featureplot.pdf", height=2, width=3)

gridExtra::grid.arrange(fgsta1_pub, fapoa1_pub, nrow = 1)

```

## scWGCNA

There aren't really many markers that separate the APOA1 cs GSTA1 - I want to see if there are co-expression patterns that might highlight any transcriptional networks that implicate control by TFs etc. Am doing this on the public dataset as they have more cells than we do.

```{r scWGCNA, echo=FALSE, message=FALSE}

if (!dir.exists("export.dir/scWGCNA")){
  dir.create("export.dir/scWGCNA")
}

if (!file.exists("export.dir/scWGCNA/scwgcna_enterocytes.Rds")){

  # have to convert the seurat object
  scwgcna_epithelium <- public_sobj
  scwgcna_epithelium[["RNA"]] <- as(object = scwgcna_epithelium[["RNA"]], Class = "Assay") 
  scwgcna_epithelium <- ScaleData(scwgcna_epithelium)
  scwgcna_epithelium <- FindVariableFeatures(scwgcna_epithelium)
  scwgcna_epithelium <- RunPCA(scwgcna_epithelium)

  scwgcna.pcells = calculate.pseudocells(s.cells = scwgcna_epithelium,
                                          seeds=0.2,
                                          nn = 10,
                                          reduction = "pca",
                                          dims = 1:10)

  scwgcna <- run.scWGCNA(p.cells = scwgcna.pcells,
              s.cells = scwgcna_epithelium,
              is.pseudocell = F,
              features = VariableFeatures(scwgcna_epithelium))

  saveRDS(scwgcna, file = "export.dir/scWGCNA/scwgcna_enterocytes.Rds")
}

scwgcna <- readRDS("export.dir/scWGCNA/scwgcna_enterocytes.Rds")
scwgcna_epithelium <- public_sobj
scwgcna_epithelium[["RNA"]] <- as(object = scwgcna_epithelium[["RNA"]], Class = "Assay") 
scwgcna_epithelium <- ScaleData(scwgcna_epithelium)
scwgcna_epithelium <- FindVariableFeatures(scwgcna_epithelium)
scwgcna_epithelium <- RunPCA(scwgcna_epithelium)

# Plot the expression of all modules at once
apoa1_metagene_p <- scW.p.expression(s.cells = scwgcna_epithelium,
                             scWGCNA.data = scwgcna,
                             modules = 1, 
                             reduction = "pca", 
                             ncol=3) +
    ggplot2::scale_colour_gradientn(colours = c("gray90", 
            "gray90", annotation_colors_epithelium[2], annotation_colors_epithelium[2])) +
    ggtitle("APOA1 module [1]")

gsta1_metagene_p <- scW.p.expression(s.cells = scwgcna_epithelium,
                             scWGCNA.data = scwgcna,
                             modules = 5, 
                             reduction = "pca", 
                             ncol=3)  +
    ggplot2::scale_colour_gradientn(colours = c("gray90", 
            "gray90", annotation_colors_epithelium[1], annotation_colors_epithelium[1])) +
    ggtitle("GSTA1 module [5]")

sc_modules <- scwgcna[["modules"]]
apoa1_cluster <- sc_modules[[1]]
gsta1_cluster <- sc_modules[[5]]

saveRDS(apoa1_cluster, file = "export.dir/apoa1_cluster.Rds")
saveRDS(gsta1_cluster, file = "export.dir/gsta1_cluster.Rds")

gridExtra::grid.arrange(gsta1_metagene_p, apoa1_metagene_p, nrow=1)

```


## Plot the module eigengenes across cells

Here we plot the heatmap of the MEs across cells. This should hopefully enable us to indicate which module contains GSTA1 and which one contains APOA1.

```{r MES, echo=FALSE, message=FALSE}

mes <- t(as.data.frame(scwgcna$MEs))

# rename the modules to numbers
rownames(mes) <- paste0("M", seq(1,6, 1))

# Column annotation
annotation <- ifelse(rownames(mes) == "M1", "Contains APOA1", NA)
annotation <- ifelse(rownames(mes) == "M5", "Contains GSTA1", annotation)

row_anno <- data.frame(annotation = annotation)
row_anno <- rowAnnotation(annotation = row_anno$annotation,
                          col = list(annotation = c("Contains APOA1" = annotation_colors_epithelium[[2]],
                                                    "Contains GSTA1" = annotation_colors_epithelium[[1]])))

cols <- viridis(n = 10)
heat_colors <- colorRamp2(breaks = c(-0.04, 0, 0.04), colors = c(cols[1], cols[5], cols[10]))
h_modules <- Heatmap(mes, show_column_names = FALSE,
        use_raster = FALSE,
        right_annotation = row_anno,
        row_split = rownames(mes),
        na_col = "white",
        col=heat_colors)

pdf("export.dir/Figures/heatmap_modules.pdf", width = 10, height = 4)
h_modules
dev.off()



```

## Metagene of modules 1 (APOA1) and 5 (GSTA1) in the datasets

```{r metagene expression, echo=FALSE, message=FALSE}

apoa1_genes <- rownames(apoa1_cluster[apoa1_cluster$Membership >= 0.60,])
gsta1_genes <- rownames(gsta1_cluster[gsta1_cluster$Membership >= 0.60,])

# write out genes in eah of these modules
write.table(data.frame(gene = gsta1_genes),
            file = "export.dir/scWGCNA/gsta1_genes.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")

write.table(data.frame(gene = apoa1_genes),
            file = "export.dir/scWGCNA/apoa1_genes.tsv",
            row.names = FALSE,
            quote=FALSE,
            sep="\t")

Idents(enterocytes) <- "Enterocyte"
Idents(public_sobj) <- "Enterocyte"

counts <- Seurat::GetAssayData(enterocytes, assay="RNA", layer="data")
#counts_2 <- Seurat::GetAssayData(public_sobj, assay="RNA", layer="data")

apoa1 <- intersect(rownames(counts), apoa1_genes)
gsta1 <- intersect(rownames(counts), gsta1_genes)
  
colors <- colorRamp2(breaks = c(-4, 0, 4), colors = c("blue4", "white", "red4"))
     
#pub_counts <- ScaleData(as.data.frame(counts_2)[c(apoa1, gsta1),])
our_counts <- ScaleData(as.data.frame(counts)[c(apoa1, gsta1),])
         
our_heat <- Heatmap(our_counts, show_column_names = FALSE, cluster_rows = FALSE,row_names_gp = gpar(fontsize=6), row_split = c(rep("APOA1", length(apoa1)), rep("GSTA1", length(gsta1))), clustering_distance_columns = "pearson", use_raster=FALSE, col = colors)

#pub_heat <- Heatmap(pub_counts, show_column_names = FALSE, row_split = c(rep("APOA1", length(apoa1_genes)), rep("GSTA1", length(gsta1_genes))), clustering_distance_columns = "pearson", use_raster=FALSE, col=colors, cluster_rows = FALSE)

our_heat

```

## Build correlated gene graphs

Here I take the genes that are >0.6 module membership in those modules that contain GSTA1 and APOA1 to display the most highly correlated genes in these relevant modules.

```{r graphs, echo=FALSE, message = FALSE}

counts <- as.data.frame(our_counts)
genes_apoa1 <- intersect(rownames(counts), apoa1_genes)
apoa1_cor <- cor(t(counts[genes_apoa1,]))

library(igraph)
g1 <- igraph::graph_from_adjacency_matrix(apoa1_cor, weighted = TRUE, mode = "undirected", diag = FALSE)
g1 <- delete.edges(g1, E(g1)[weight < 0 ])

graph_apoa1 <- ggraph(g1, layout = "fr") +
   geom_edge_link(color=annotation_colors_epithelium[2], aes(width=weight)) +
   #geom_edge_arc(color="red4", aes(width=weight)) +
   geom_node_point(size = 25, color="black") +
   geom_node_text(aes(label = name), size=4, color="white") +
   coord_fixed() +
   theme_void()


genes_gsta1 <- intersect(rownames(counts), gsta1_genes)
gsta1_cor <- cor(t(counts[genes_gsta1,]))
g2 <- igraph::graph_from_adjacency_matrix(gsta1_cor, weighted = TRUE, mode = "undirected", diag = FALSE)
g2 <- delete.edges(g2, E(g2)[weight < 0 ])


graph_gsta1 <- ggraph(g2, layout = "fr") +
   geom_edge_link(color=annotation_colors_epithelium[1], aes(width=weight)) +
   #geom_edge_arc(color="red4", aes(width=weight)) +
   geom_node_point(size = 25, color="black") +
   geom_node_text(aes(label = name), size=4, color="white") +
   coord_fixed() +
   theme_void()



# Figure 2
pdf("export.dir/Figures/gsta1_network_graph.pdf", height=6, width=6)
graph_gsta1
dev.off()

pdf("export.dir/Figures/apoa1_network_graph.pdf", height=6, width=6)
graph_apoa1
dev.off()

graph_apoa1
graph_gsta1

```

## Plot a couple of examples

I'm particularly interested in the observation that there are bile acid related genes in the GSTA1 module. I want to look at a couple of examples where bile acid genes correlate with detox genes.

```{r gsta1 example scatters, echo=FALSE, message = FALSE}


to_cor_gsta1 <- data.frame(GSTA1 = unlist(counts["GSTA1",]),
                           FABP6 = unlist(counts["FABP6",]))

ggplot(to_cor_gsta1, aes(x = FABP6, y = GSTA1)) +
  geom_point(color = annotation_colors_epithelium[[1]], size=0.2) +
  theme_classic() +
  stat_smooth(method="lm")

```


```{r fgsea module, echo=FALSE, message=FALSE}

pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")
apoa1_fgsea <- fgsea::fora(pathways, apoa1_genes, unlist(scwgcna$module.genes))
apoa1_fgsea <- apoa1_fgsea[order(apoa1_fgsea$padj, decreasing = FALSE),]

gsta1_fgsea <- fgsea::fora(pathways, gsta1_genes, unlist(scwgcna$module.genes))
gsta1_fgsea <- gsta1_fgsea[order(gsta1_fgsea$padj, decreasing = FALSE),]

# add enrichment ratio
apoa1_fgsea <- pathway_fold_enrichment(as.data.frame(apoa1_fgsea),
                                       length(apoa1_genes),
                                       pathways,
                                       rownames(counts))


gsta1_fgsea <- pathway_fold_enrichment(as.data.frame(gsta1_fgsea),
                                       length(gsta1_genes),
                                       pathways,
                                       rownames(counts))


# Plot the results - top 5

## GSTA1
gsta1_fgsea_top <- gsta1_fgsea[1:5,]
gsta1_fgsea_top <- gsta1_fgsea_top[order(gsta1_fgsea_top$foldEnrichment, decreasing = FALSE),]
gsta1_fgsea_top$pathway <- factor(gsta1_fgsea_top$pathway, levels = gsta1_fgsea_top$pathway)


gsta_fgsea_plt <- ggplot(gsta1_fgsea_top, aes(x=pathway, y=foldEnrichment)) +
  geom_bar(stat="identity", fill = annotation_colors_epithelium[[1]]) +
  theme_classic() +
  coord_flip()


## APOA1
apoa1_fgsea_top <- apoa1_fgsea[1:5,]
aopa1_fgsea_top <- apoa1_fgsea_top[order(apoa1_fgsea_top$foldEnrichment, decreasing = TRUE),]
apoa1_fgsea_top$pathway <- factor(apoa1_fgsea_top$pathway, levels = apoa1_fgsea_top$pathway)

apoa1_fgsea_plt <- ggplot(apoa1_fgsea_top, aes(x=pathway, y=foldEnrichment)) +
  geom_bar(stat="identity", fill = annotation_colors_epithelium[[2]]) +
  theme_classic() +
  coord_flip()

# save plots
ggsave(gsta_fgsea_plt, file = "export.dir/Figures/gsta1_pathways.pdf", height=2, width=7)
ggsave(apoa1_fgsea_plt, file = "export.dir/Figures/apoa1_pathways.pdf", height=2, width=7)

gsta_fgsea_plt
apoa1_fgsea_plt

```


## Differential expression - pseudobulk

Here we perform differential expression analysis on pseudobulk expression across cells in cell types that have enough cells (typically >20/individual).

### Enterocytes

We take all enterocytes together i.e. the APOA1 and the GSTA1 high subsets. Given that the DUOX2 ones look so different we don't include these in this initial analysis. The two samples that have very high DUOX2 enterocytes are removed from this analysis as they only have 1 or 3 cells of these major enterocyte subsets.

```{r pseudobulk enterocytes, echo=FALSE, message=FALSE}

# table the number of cells per invididual
enterocytes <- subset(epithelium, idents = c("Enterocyte_APOA1", "Enterocyte_GSTA1"))
enterocytes@meta.data$annotation <- as.character(Idents(enterocytes))
Idents(enterocytes) <- enterocytes@meta.data$orig.ident
ncells_entero <- as.data.frame(table(Idents(enterocytes)))
knitr::kable(ncells_entero)

# remove troublesome samples
enterocytes <- subset(enterocytes, idents = c("G-SCI7T080-SCI5T080", "G-SCI7T033-SCI5T033"), invert=TRUE)

Idents(enterocytes) <- enterocytes@meta.data$annotation

# Write our enterocytes to compare with caecum colonocytes
saveRDS(enterocytes, file = "export.dir/ileum_enterocytes.Rds")

# aggregate for pseudobulk analysis
agg_all <- AggregateExpression(enterocytes,
                               group.by = c("orig.ident"))                                                             

agg_all <- as.data.frame(agg_all)
colnames(agg_all) <- gsub("RNA\\.", "", colnames(agg_all))
rownames(metadata) <- gsub("-", "\\.", rownames(metadata))
coldata <- metadata[colnames(agg_all),]
dds <- DESeqDataSetFromMatrix(agg_all, colData = coldata, design=~disease_type)
dds$disease_type <- relevel(dds$disease_type, ref = "HC")
dds <- dds[rowSums(counts(dds) >= 10) >= 3,]
dds <- DESeq2::DESeq(dds)

# get results for each contrast
# lfc shrink
res1 <- lfcShrink(dds,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2 <- lfcShrink(dds,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds, name = "disease_type_UC_vs_HC"),
                  type = "normal")

# save the results to compare with caecum colonocytes
write.table(data.frame(gene = rownames(res1), as.data.frame(res1)),
            file = "export.dir/ileum_enterocytes_pscuc_vs_healthy.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)

```

#### Volcano plot

```{psc enterocytes volcano, echo=FALSE, message=FALSE}

res1_toplot <- res1
res1$log2FoldChange[res1$log2FoldChange < -5] <- -5

psc_enterocyte_volc <- volcano_plot(res1_toplot, use="padj", labels = c("SPINK1", "GSTA1", "RBP2", "FABP6")) +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = "blue3") +
  theme(legend.position = "none")

psc_enterocyte_volc

# Figure
ggsave(psc_enterocyte_volc, file = "export.dir/Figures/enterocyte_psc_volcano.pdf", height = 3, width=3)

uc_enterocyte_volc <- volcano_plot(res2, use="padj") +
  ggtitle("UC vs. Healthy") +
  scale_color_manual(values = "blue3") +
  theme(legend.position = "none")

ggsave(uc_enterocyte_volc, file = "export.dir/Figures/enterocyte_uc_volcano.pdf", height = 5, width=5)

uc_enterocyte_volc
```

```{r table of differences enterocytes, echo=FALSE, message=FALSE}

nup_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange > 0 & !(is.na(res1$padj)),])
nup_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange > 0 & !(is.na(res2$padj)),])
ndown_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange < 0 & !(is.na(res1$padj)),])
ndown_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange < 0 & !(is.na(res2$padj)),])

results_ent <- data.frame(comparison = resultsNames(dds)[2:3],
                          no.up = c(nup_psc, nup_uc),
                          no.down = c(ndown_psc, ndown_uc))

knitr::kable(results_ent)
```

### Enrichments

```{r enrichments, echo=FALSE, message=FALSE}

genes <- res1$log2FoldChange
names(genes) <- rownames(res1)
pathways_gsta1 <- list("GSTA1_Module" = gsta1_genes,
                       "APOA1_Module" = apoa1_genes)

set.seed(100)
fgsea_enterocyte_psc <- fgseaMultilevel(pathways_gsta1, genes)

enrich_gsta1 <- plotEnrichment(pathway = pathways_gsta1[["GSTA1_Module"]], genes) +
  theme_classic() +
  geom_line(color=annotation_colors_epithelium[[1]], aes(x = rank, y = ES), linewidth = 2) +
  ggtitle(paste0("fgsea p = ", round(fgsea_enterocyte_psc[2,]$padj, 2))) +
  ylim(c(-0.8,0.1))

enrich_apoa1 <- plotEnrichment(pathway = pathways_gsta1[["APOA1_Module"]], genes) +
  theme_classic() +
  geom_line(color=annotation_colors_epithelium[[2]], aes(x = rank, y = ES), linewidth = 2) +
  ggtitle(paste0("fgsea p = ", round(fgsea_enterocyte_psc[1,]$padj, 2))) +
  ylim(c(-0.8,0.1))

ggsave(enrich_gsta1, file = "export.dir/Figures/enterocyte_fgsea_gsta1.pdf", height = 3, width=3)
ggsave(enrich_apoa1, file = "export.dir/Figures/enterocyte_fgsea_apoa1.pdf", height = 3, width=3)

gridExtra::grid.arrange(enrich_gsta1, enrich_apoa1, nrow = 1)
```

## Leading edge genes in GSTA1 module

```{r leading edge, echo=FALSE, message=FALSE}

vst_all <- as.data.frame(assay(vst(dds)))
gois_leadingedge <- unlist(fgsea_enterocyte_psc[2,]$leadingEdge)
leading_gsta1 <- OCMSutility::featurebox(vst_all, coldata, features=gois_leadingedge, group_by = "disease_type") +
  theme_classic() +
  facet_wrap(~feature, scales = "free") +
  scale_color_manual(values = rep(annotation_colors_epithelium[[1]], 3)) +
  theme(legend.position = "none")

ggsave(leading_gsta1, file = "export.dir/Figures/gsta1_leading_edge.pdf", height=2.5, width=4)
gois_leadingedge

leading_gsta1

res1[rownames(res1) %in% gois_leadingedge,]

```

## Bile acid pathway

We expect that genes that are regulated by bile acid signalling would be decreased in PSC because of reduced bbile flow. They don't seem to come out in our analysis but that might be a power issue. Here we look at them to see how they look like


```{r bile acid pathways, echo=FALSE, message=FALSE}

bile_acid <- OCMSutility::featurebox(vst_all, coldata, features=c("SLC51A", "SLC10A2", "SLC51B", "FGF19"), group_by = "disease_type") +
  theme_classic() +
  facet_wrap(~feature, scales = "free") +
  scale_color_manual(values = unlist(group_colors)) +
  theme(legend.position = "none")

ggsave(bile_acid, file = "export.dir/Figures/bile_acid_genes.pdf", height=5, width=7)
bile_acid

res1[c("SLC51A", "SLC10A2", "SLC51B", "FGF19"),]

```

It's interesting and could be a power issue...

## GSTA genes

There is a lot of homology in the GST alpha family. Here we just see if there is any evidence for changes in other GSTAs

```{r gstas, echo=FALSE, message=FALSE}
gsta_genes <- OCMSutility::featurebox(vst_all, coldata,
                                      features=c("GSTA1",
                                                 "GSTA2"),
                                      group_by = "disease_type") +
  theme_classic() +
  facet_wrap(~feature, scales = "free") +
  scale_color_manual(values = rep(annotation_colors_epithelium[[1]], 3)) +
  theme(legend.position = "none")
gsta_genes

```

## Mouse GSTAs

```{r mouse gstas, echo=FALSE, message = FALSE}

FeaturePlot(merged_ileum, features = c("GSTA1"))

mouse_dat <- readRDS("/well/powrie/shared/scrnaseq_shen/whole_final_annotation.rds")

mouse_dat <- subset(mouse_dat, subset = tissue == "ileum")

spf <- subset(mouse_dat, subset = mice == "SPF")
gf <- subset(mouse_dat, subset = mice == "GF")

Idents(spf) <- spf@meta.data$celltypes
Idents(gf) <- gf@meta.data$celltypes

FeaturePlot(spf, features = c("Gsta2"))
FeaturePlot(gf, features = c("Gsta1", "Gsta2"))

```


### Heatmap

```{r psc heatmap, echo=FALSE, message=FALSE}

vst_all <- as.data.frame(assay(vst(dds)))

res1$label <- ifelse(rownames(res1) %in% gsta1_genes, rownames(res1), NA)
res1$label <- ifelse(rownames(res1) %in% gsta1_genes, rownames(res1), NA)

psc_genes <- rownames(res1[res1$padj < 0.05 & !(is.na(res1$padj)),])
uc_genes <- rownames(res2[res2$padj < 0.05 & !(is.na(res2$padj)),])
psc_genes <- setdiff(psc_genes, uc_genes)

col_annotation <- HeatmapAnnotation(disease_type = coldata$disease_type,
                                    col = list(disease_type = unlist(group_colors)))

heatmap_psc_genes <- Heatmap(ScaleData(vst_all[psc_genes,]), column_gap = unit(2, "mm"),
        cluster_columns = FALSE,
        column_split = coldata$disease_type,
        show_column_names = FALSE,
        top_annotation = col_annotation,
        col = colorRamp2(breaks = c(-2,0, 2), c("blue4", "white", "red4")),
        rect_gp = gpar(col = "white", lwd=1))  

pdf("export.dir/Figures/heatmap_enterocytes_psc.pdf", height = 7, width =7)
heatmap_psc_genes
dev.off()

# draw the heatmap
heatmap_psc_genes

```

### Gene expression on PCA of enterocytes

```{r featureplots of psc gois}


FeaturePlot(enterocytes, features=c("MT1X"), split.by="disease_type", reduction = "pca", pt.size=1)


```

