---
title: "Ileum immune cell analysis: Forced cells 10000"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}

# temporary solution to libpaths
.libPaths("~/R/x86_64-pc-linux-gnu-library/4.3/")

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args=list(png=list(type="cairo")))
options(bitmapType = "cairo")
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")

# reticulate stuff
reticulate::use_virtualenv('~/devel/venv/Python-3.10.8-GCCcore-12.2.0/skylake/', required=TRUE)

```

```{r load libraries, echo=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(speckle)
library(scWGCNA)
library(circlize)
library(ggraph)
library(igraph)
library(DESeq2)
library(nVennR)


# create Clustering analysis dir
if (!(dir.exists("export.dir/Immune"))){
  dir.create("export.dir/Immune")
}
```


```{r set colors, echo=FALSE, message=FALSE, warning=FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```


```{r read in pathways, echo=FALSE, message=FALSE}

pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")

```


## Cluster all immune cells

Here we re-cluster all the immune cells. 

```{r cluster immune cells, echo=FALSE, message=FALSE}

# setting seed for all clustering/umappy things
set.seed(10)

# Read in data
merged_ileum <- readRDS("export.dir/Clustering/ileum_harmonised_force_cells_annotated.Rds")

unique(Idents(merged_ileum))

immune_cells <- subset(merged_ileum, idents = c("CD4_T_Cell",
                                                "CD8_T_cell",
                                                "B_cell",
                                                "B_Cell_2",
                                                "Gamma_Delta_T_Cell",
                                                "Mast_Cell",
                                                "Mono_Mac",
                                                "Mono_Mac_Inf",
                                                "B_Cell_IgA",
                                                "B_Cell_IgG",
                                                "B_Cell_Activated"))

# take out high mt genes now
immune_cells <- subset(immune_cells, subset = percent.mt < 10)

# recluster etc
immune_cells <- ScaleData(immune_cells, features = rownames(immune_cells))
immune_cells <- FindVariableFeatures(immune_cells, selection.method = "vst", nfeatures = 2000)
immune_cells <- RunPCA(immune_cells, features = VariableFeatures(object = immune_cells))

# harmony integration: by individual = orig.ident
immune_cells <- immune_cells %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
immune_cells <- FindNeighbors(immune_cells, reduction="harmony", dims = 1:30)
immune_cells <- FindClusters(immune_cells, resolution = 0.6, algorithm = 4)
immune_cells <- RunUMAP(immune_cells, dims = 1:30, reduction = "harmony")

```

## Cell cycle markers

```{r cell cycle markers, echo=FALSE, message =FALSE}

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
immune_cells <- JoinLayers(immune_cells)

immune_cells <- CellCycleScoring(immune_cells, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

DimPlot(immune_cells, label = TRUE)
Idents(immune_cells) <- immune_cells@meta.data$seurat_clusters

```


## Find markers and annotate

### Dimplot

```{r find markers immune, echo=FALSE, message=FALSE}
markers_immune_cells <- FindAllMarkers(immune_cells)

DimPlot(immune_cells, label=TRUE, reduction = "umap") +
  ggtitle(paste0("No. cells = ", nrow(immune_cells@meta.data)))

```

```{r immune cell marker genes, echo=FALSE, message=FALSE, fig.width=11}

t_cell_genes <- c("PTPRC",
                  "CD3D",
                  "CD4",
                  "TRDC",
                  "CD8A",
                  "RORC",
                  "SELL",
                  "IL7R",
                  "CD2",
                  "S100A4",
                  "CCR7",
                  "Il22",
                  "IL23R",
                  "KLF2",
                  "FOXP3",
                  "CD69",
                  "CTLA4",
                  "IL2RA",
                  "IKZF2",
                  "TNFRSF18",
                  "LAG3",
                  "IL17A",
                  "IFNG",
                  "TNF",
                  "CXCR3",
                  "CCR5",
                  "GZMA",
                  "GZMK",
                  "TWIST1")

b_cell_genes <- c("CD19",
                  "CD79A",
                  "CD27",
                  "MZB1",
                  "IGHA1",
                  "IGHA2",
                  "IGHG1",
                  "IGHG2",
                  "IGHD",
                  "CCL3",
                  "CCL4",
                  "MEF2B",
                  "CXCR4",
                  "CD83")


mono_mac_genes <- c("CD68",
                    "CD14",
                    "CD74",
                    "HLA-DRA",
                    "S100A8",
                    "S100A9",
                    "IL1B",
                    "TCF4",
                    "C1QA",
                    "LAMP3")

mast_cell_genes <- c( "KIT",
                     "FCER1A")


DotPlot(immune_cells, features = c(t_cell_genes, b_cell_genes, mono_mac_genes, mast_cell_genes, "MKI67", "ITGAX"),
        cols = colorRampPalette(c("blue", "red"))(2),
        cluster.idents = TRUE) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) 
```


```{r annotated immune clusters, echo=FALSE, message=FALSE}

immune_cell_annotations <- c("CD4_Memory",
                             "Plasma_Cell_IgA",
                             "CD8_Memory",
                             "B_Cell_memory_IgD",
                             "B_Cell_Memory",
                             "Circulating_Monocyte",
                             "Treg",
                             "B_Cell_Cycling",
                             "CD4_Activated",
                             "Gamma_Delta_T_Cell",
                             "B_Cell_MEF2B",
                             "CD8_Memory_GZMK",
                             "cDC",
                             "CD4_CD8_Memory",
                             "Mast_Cell",
                             "Plasma_Cell_IgG")
                             
names(immune_cell_annotations) <- levels(Idents(immune_cells))
immune_cells <- RenameIdents(immune_cells, immune_cell_annotations)

immune_cells <- subset(immune_cells, idents = c("CD4_CD8_Memory"), invert=TRUE)
immune_cell_anno_colors <- OCMSutility::get_palette(n = 50)[17:32]
names(immune_cell_anno_colors) <- as.character(levels(Idents(immune_cells)))

# save the colors
saveRDS(immune_cell_anno_colors, file="export.dir/Immune/immune_cell_colors.Rds")

dimplot_immune <- DimPlot(immune_cells, label = TRUE, repel = TRUE) +
  theme(legend.position = "none") +
  scale_color_manual(values = immune_cell_anno_colors) +
  ggtitle(paste0("No. cells = ", nrow(immune_cells@meta.data)))

dimplot_immune 
  
# Figure 1
ggsave(dimplot_immune, file = "export.dir/Figures/Immune_umap.pdf", height=5, width=5)

saveRDS(immune_cells, file = "export.dir/Clustering/ileum_immune_clusters_annotated.Rds")

```

We remove the ones that are CD4_CD8_memory as the look odd.

## DotPlot with annotations

This is to show how we annotated the immune cell clusters.

```{r dotplot with annotations, echo=FALSE, message=FALSE}

dotplot_immune <- DotPlot(immune_cells, features = c(t_cell_genes, b_cell_genes, mono_mac_genes, mast_cell_genes),
        cols = colorRampPalette(c("blue", "red"))(2),
        cluster.idents = TRUE) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) 

dotplot_immune
ggsave(file = "export.dir/Immune/immune_dotplot.pdf", height = 3, width = 15)
```

## Proportion of immune cells across samples

```{r proportion immune cells, echo = FALSE, message = FALSE}

# read in metadata
metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)


rownames(metadata) <- gsub("\\.", "-", rownames(metadata))
immune_props_plt <- plot_props(immune_cells, metadata, immune_cell_anno_colors) +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

# Figure 3
ggsave(immune_props_plt, file = "export.dir/Figures/immune_cells_profile.pdf", height=3, width = 5)

immune_props_plt

immune_cells@meta.data$annotation <- Idents(immune_cells)
```

```{r ncells immune, echo=FALSE, message=FALSE}

clusters <- unique(Idents(immune_cells))
ncells_cluster = list()
ninds_cluster = list()

for (i in 1:length(clusters)){
  flog.info(paste0("Counting cells for cluster ", clusters[i]))
  dat <- subset(immune_cells, idents = clusters[i])
  n <- ncol(dat)
  nind <- length(unique(as.character(dat@meta.data$orig.ident)))
  ncells_cluster[[i]] <- n
  ninds_cluster[[i]] <- nind
}
names(ncells_cluster) <- clusters

ncells_cluster <- as.data.frame(t(as.data.frame(ncells_cluster)))

ncells_cluster$cluster <- gsub("X", "", rownames(ncells_cluster))
colnames(ncells_cluster) <- c("ncells", "cluster")

# order by cluster size
ncells_cluster <- ncells_cluster[order(ncells_cluster$ncells, decreasing = FALSE),]
ncells_cluster$cluster <- factor(ncells_cluster$cluster, levels = ncells_cluster$cluster)

ncells_cluster$prop <- (ncells_cluster$ncells/8395)*100
knitr::kable(ncells_cluster)

write.table(ncells_cluster,
            file = "export.dir/Immune/ileum_proportions.tsv",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

## Compare proportions of immune cell clusters

Here we use propeller to see of there are any significant differences in immune cell popoulations between the disease groups.


```{r props immune cells, echo=FALSE, message=FALSE}

annos_immune <- immune_cells@meta.data
annos_immune$annotation <- Idents(immune_cells)

annos_immune <- annos_immune %>% 
  group_by(orig.ident, annotation) %>%
  summarize(n = n())

annos_immune$disease_type <- metadata[annos_immune$orig.ident,]$disease_type

annos_immune_prop <- annos_immune %>% 
  mutate(prop = (n/sum(n))*100)

comparisons <- list(c(1,2), c(1,3), c(2,3))
annos_immune_prop <- annos_immune_prop[!is.na(annos_immune_prop$annotation),]
ggplot(annos_immune_prop, aes(x=disease_type, y=prop, color=annotation)) +
  geom_boxplot(outlier.alpha = 0, fill = "white") +
  geom_jitter(width=0.1, height=0) +
  theme_classic() +
  facet_wrap(~annotation, scales="free", nrow = 4) +
  ggpubr::stat_compare_means(comparisons = comparisons, size=3) +
  scale_color_manual(values = immune_cell_anno_colors) +
  theme(legend.position = "none")

ggsave(file = "export.dir/Immune/immune_cell_composition_differences.pdf", height = 10, width = 10)

prop_list_immune <- getTransformedProps(clusters = Idents(immune_cells),
                                 sample = immune_cells@meta.data$orig.ident)

metadata_immune <- metadata[colnames(prop_list_immune[[1]]),]

design_immune <- model.matrix(~0+disease_type, data = metadata_immune)

res_comp_immune <- propeller.anova(prop_list_immune, design=design_immune, coef=c(1,2,3), robust=FALSE,
  trend=FALSE, sort=TRUE)

knitr::kable(res_comp_immune)

```

There are no statisitically significant differences between disease groups.


## Number of cells per person 

Here we look at the number of cells in each individual to determine how well we can do pseudobulk.

```{r n per individual immune cells, echo=FALSE, message=FALSE}

ncells_per_sample_immune <- as.data.frame(table(immune_cells@meta.data$annotation,
                                                immune_cells@meta.data$orig.ident))
knitr::kable(ncells_per_sample_immune)

# only analyse when there are at least 20 cells per indivudual
ncells_per_sample_immune$disease_type <- metadata[as.character(ncells_per_sample_immune$Var2), ]$disease_type
ggplot(ncells_per_sample_immune, aes(x=disease_type, y=Freq, color=disease_type)) +
  geom_point() +
  theme_classic() +
  facet_wrap(~Var1, scales = "free") +
  geom_hline(yintercept = 20,  linetype="dashed") +
  theme(axis.text.x = element_blank()) +
  scale_color_manual(values = group_colors) +
 # ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  ggrepel::geom_text_repel(label = ncells_per_sample_immune$Freq)


```


## CD4 memory pseudobulk

```{r cd4 memory pseudobulk, echo=FALSE, message = FALSE}

set.seed(10)

cd4 <- subset(immune_cells, idents = "CD4_Memory")

# Some oddities with CD8 and Ig in there so
# recluster and think about filtering choices
cd4 <- ScaleData(cd4, features = rownames(cd4))
cd4 <- FindVariableFeatures(cd4, selection.method = "vst", nfeatures = 2000)
cd4 <- RunPCA(cd4, features = VariableFeatures(object = cd4))

# harmony integration: by individual = orig.ident
cd4 <- cd4 %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
cd4 <- FindNeighbors(cd4, reduction="harmony", dims = 1:30)
cd4 <- FindClusters(cd4, resolution = 1, algorithm = 4)
cd4 <- RunUMAP(cd4, dims = 1:30, reduction = "harmony")

cd4_markers <- FindAllMarkers(cd4, only.pos = TRUE)

DimPlot(cd4, label=TRUE, reduction = "umap")
FeaturePlot(cd4, label=TRUE, reduction = "umap", features = c("CD8A", "CD4", "JCHAIN", "MS4A1", "CD79A"))
DotPlot(cd4, features = c("CD8A", "CD4", "JCHAIN", "MS4A1", "CD79A"))

# Remove clusters that are high for CD8
cd4 <- subset(cd4, idents = c(1,3,4,5))

# aggregate and diff expression
agg_cd4 <- AggregateExpression(cd4, group.by = c("orig.ident"))
agg_cd4 <- as.data.frame(agg_cd4)
colnames(agg_cd4) <- gsub("RNA\\.", "", colnames(agg_cd4))

meta_cd4 <- metadata
rownames(meta_cd4) <- gsub("-", "\\.", rownames(meta_cd4))
coldata <- meta_cd4[colnames(agg_cd4),]

dds_cd4 <- DESeqDataSetFromMatrix(agg_cd4, colData = coldata, design=~disease_type)
dds_cd4$disease_type <- relevel(dds_cd4$disease_type, ref = "HC")
dds_cd4 <- dds_cd4[rowSums(counts(dds_cd4) >= 10) >= 4,]
dds_cd4 <- DESeq2::DESeq(dds_cd4)

# lfc shrink
res1 <- lfcShrink(dds_cd4,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds_cd4, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2 <- lfcShrink(dds_cd4,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds_cd4, name = "disease_type_UC_vs_HC"),
                  type = "normal")

genes_psc <- rownames(res1[res1$padj < 0.05 & !is.na(res1$padj),])
genes_uc <- rownames(res2[res2$padj < 0.05 & !is.na(res2$padj),])

cd4_memory_volc <- volcano_plot(res1, use="padj") +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = immune_cell_anno_colors[[1]]) +
  theme(legend.position = "none")

ggsave(cd4_memory_volc, file = "export.dir/Figures/cd4_memory_volcano.pdf", height = 3, width=3)

cd4_memory_volc

# Write out the results to compare with the caecum
write.table(data.frame(gene = rownames(res1), as.data.frame(res1)),
            file = "export.dir/ileum_cd4_mem_pscuc_vs_healthy.tsv",
            row.names = FALSE,
            quote = FALSE,
            sep = "\t")

# Write the cd4
saveRDS(cd4, file = "export.dir/Immune/ileum_cd4_memory.Rds")

```

```{r table of diffences, echo=FALSE, message=FALSE}

nup_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange > 0 & !(is.na(res1$padj)),])
nup_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange > 0 & !(is.na(res2$padj)),])
ndown_psc <- nrow(res1[res1$padj < 0.05 & res1$log2FoldChange < 0 & !(is.na(res1$padj)),])
ndown_uc <- nrow(res2[res2$padj < 0.05 & res2$log2FoldChange < 0 & !(is.na(res2$padj)),])

results_cd4 <- data.frame(comparison = resultsNames(dds_cd4)[2:3],
                          no.up = c(nup_psc, nup_uc),
                          no.down = c(ndown_psc, ndown_uc))

knitr::kable(results_cd4)
```


## Where are these genes expressed

```{r expression of genes across cell types, echo=FALSE, message=FALSE}
cd4_gois <- rownames(res1[res1$padj < 0.05 & res1$log2FoldChange < 0 & !(is.na(res1$padj)),])

FeaturePlot(immune_cells, features = cd4_gois, label=TRUE)

# Violin plot each goi separately as it is a bit annoying
# to do the default seurat one
cd4_gois_p <- list()
for (i in cd4_gois){
  p <- VlnPlot(immune_cells, features = i, pt.size = 0) +
  scale_fill_manual(values = immune_cell_anno_colors) +
  theme(legend.position = "none") 
  cd4_gois_p[[i]] <- p
  ggsave(p, file = paste0("export.dir/Figures/", "cd4_", i, "_celltypes.pdf"), height=4, width=4)
}

gridExtra::grid.arrange(grobs = cd4_gois_p, nrow=3)

```



## Feature boxplots of the changed genes in CD4 Memory cells

```{r featurebox gois cd4 memory, echo=FALSE, message=FALSE}

vst_cd4 <- as.data.frame(assay(vst(dds_cd4)))

# write out for compass
write.table(data.frame("Gene Symbol" = rownames(vst_cd4), vst_cd4),
            file = "export.dir/ileum_cd4_vst.tsv",
            sep = "\t",
            row.names = FALSE,
            quote = FALSE)

gois_cd4 <- OCMSutility::featurebox(vst_cd4, coldata, features=genes_psc, group_by = "disease_type") +
              theme_classic() +
              facet_wrap(~feature, scales="free") +
              scale_color_manual(values = rep(immune_cell_anno_colors[[1]], 3)) +
  theme(legend.position = "none")


ggsave(gois_cd4, file = "export.dir/Figures/cd4_memory_gois.pdf", height=4, width=6)

gois_cd4

# Write out cd4 vst data
write.table(data.frame(gene = rownames(vst_cd4), vst_cd4),
            file = "export.dir/Immune/ileum_cd4_mem_vst.tsv",
            row.names = FALSE,
            quote = FALSE,
            sep = "\t")

```


## CXCR3+ cells

```{r cxcr3, echo=FALSE, message=FALSE}

ppos_cxcr3 <- get_ppos(cd4, "CXCR3")
ppos_cxcr3$individual <- gsub("-", "\\.", ppos_cxcr3$individual)
ppos_cxcr3$disease_type <- meta_cd4[ppos_cxcr3$individual,]$disease_type

cxcr3_pos_p <- ggplot(ppos_cxcr3, aes(x=disease_type, y=ppos, color=disease_type)) +
  geom_boxplot(outlier.alpha=0) +
  geom_jitter(height=0, width=0.1) +
  scale_color_manual(values = unlist(group_colors)) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  theme(legend.position = "none")

ggsave(cxcr3_pos_p, file = "export.dir/Figures/CXCR3_positivity.pdf", height=2.5, width=2.5)
cxcr3_pos_p

```


## CCR9+ cells

Of interest, Kate Williamson has found reduced CCR9+ T cells in the ileum by FACs. Also increased circulating CCR9+ cells. We see here if this is also true in our scRNA-seq data.It would aslo fit with the epithelial signature of reduced CCL25.

```{r ccr9, echo=FALSE, message=FALSE}

ppos_ccr9 <- get_ppos(cd4, "CCR9")
ppos_ccr9$individual <- gsub("-", "\\.", ppos_ccr9$individual)
ppos_ccr9$disease_type <- meta_cd4[ppos_ccr9$individual,]$disease_type

ccr9_pos_p <- ggplot(ppos_ccr9, aes(x=disease_type, y=ppos, color=disease_type)) +
  geom_boxplot(outlier.alpha=0) +
  geom_jitter(height=0, width=0.1) +
  scale_color_manual(values = unlist(group_colors)) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  theme(legend.position = "none")

ggsave(ccr9_pos_p, file = "export.dir/Figures/CCR9_positivity.pdf", height=2.5, width=2.5)
ccr9_pos_p

```


## CD4 FGSEA


```{r cd4 fgsea, echo=FALSE, message=FALSE}

set.seed(10)

cd4_genes <- res1$log2FoldChange
names(cd4_genes) <- rownames(res1)

# gene set enrichments
cd4_fgsea <- fgsea::fgsea(pathways, cd4_genes, minSize=50)
cd4_fgsea <- cd4_fgsea[order(cd4_fgsea$padj, decreasing = FALSE),]

cd4_pathways_sig <- cd4_fgsea[cd4_fgsea$padj < 0.05,]

cd4_top10 <- cd4_pathways_sig[1:10,]
cd4_top10$pathway <- factor(cd4_top10$pathway, levels = cd4_top10$pathway)
cd4_top10_p <- ggplot(cd4_top10, aes(x = pathway, y=NES, fill=-log10(padj))) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low="snow2", high=immune_cell_anno_colors[[1]]) +
  theme_classic() +
  coord_flip()
  
ggsave(cd4_top10_p, file = "export.dir/Figures/cd4_memory_pathway_barplot.pdf", height=2.5, width=10)

cd4_top10_p

```

```{r cd4 in uc, echo=FALSE, message=FALSE}

cd4_genes_uc <- res2$log2FoldChange
names(cd4_genes_uc) <- rownames(res2)

# gene set enrichments
cd4_fgsea_uc <- fgsea::fgsea(pathways, cd4_genes_uc, minSize=50)
cd4_fgsea_uc <- cd4_fgsea_uc[order(cd4_fgsea_uc$padj, decreasing = FALSE),]

cd4_pathways_sig_uc <- cd4_fgsea_uc[cd4_fgsea_uc$padj < 0.05,]

cd4_top10_uc <- cd4_pathways_sig_uc[1:10,]
cd4_top10_uc$pathway <- factor(cd4_top10_uc$pathway, levels = cd4_top10_uc$pathway)
cd4_top10_p_uc <- ggplot(cd4_top10_uc, aes(x = pathway, y=NES, fill=-log10(padj))) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low="snow2", high=immune_cell_anno_colors[[2]]) +
  theme_classic() +
  coord_flip()
  
ggsave(cd4_top10_p, file = "export.dir/Figures/cd4_memory_pathway_barplot.pdf", height=2.5, width=10)

cd4_top10_p_uc



```


## CD8 memory pseudobulk

```{r cd8 memory pseudobulk, echo=FALSE, message = FALSE}

set.seed(10)

cd8 <- subset(immune_cells, idents = "CD8_Memory")

cd8 <- ScaleData(cd8, features = rownames(cd8))
cd8 <- FindVariableFeatures(cd8, selection.method = "vst", nfeatures = 2000)
cd8 <- RunPCA(cd8, features = VariableFeatures(object = cd8))

# harmony integration: by individual = orig.ident
cd8 <- cd8 %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
cd8 <- FindNeighbors(cd8, reduction="harmony", dims = 1:30)
cd8 <- FindClusters(cd8, resolution = 0.8, algorithm = 4)
cd8 <- RunUMAP(cd8, dims = 1:30, reduction = "harmony")


```



```{r cd8 pseudobulk, echo=FALSE, message=FALSE}

agg_cd8 <- AggregateExpression(cd8, group.by = c("orig.ident"))
agg_cd8 <- as.data.frame(agg_cd8)
colnames(agg_cd8) <- gsub("RNA\\.", "", colnames(agg_cd8))

meta_cd8 <- metadata
rownames(meta_cd8) <- gsub("-", "\\.", rownames(meta_cd8))
coldata_cd8 <- meta_cd8[colnames(agg_cd8),]

dds_cd8 <- DESeqDataSetFromMatrix(agg_cd8, colData = coldata_cd8, design=~disease_type)
dds_cd8$disease_type <- relevel(dds_cd8$disease_type, ref = "HC")
dds_cd8 <- dds_cd8[rowSums(counts(dds_cd8) >= 10) >= 4,]
dds_cd8 <- DESeq2::DESeq(dds_cd8)

res1_cd8 <- lfcShrink(dds_cd8,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds_cd8, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2_cd8 <- lfcShrink(dds_cd8,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds_cd8, name = "disease_type_UC_vs_HC"),
                  type = "normal")

genes_psc_cd8 <- rownames(res1_cd8[res1_cd8$padj < 0.05 & !is.na(res1_cd8$padj),])
genes_uc_cd8 <- rownames(res2_cd8[res2_cd8$padj < 0.05 & !is.na(res2_cd8$padj),])

cd8_memory_volc <- volcano_plot(res1_cd8, use="padj") +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = immune_cell_anno_colors[[3]]) +
  theme(legend.position = "none")

ggsave(cd8_memory_volc, file = "export.dir/Figures/cd8_memory_volcano.pdf", height = 3, width=3)

cd8_memory_volc

```

```{r table of differences cd8, echo=FALSE, message=FALSE}

nup_psc <- nrow(res1_cd8[res1_cd8$padj < 0.05 & res1_cd8$log2FoldChange > 0 & !(is.na(res1_cd8$padj)),])
nup_uc <- nrow(res2_cd8[res2_cd8$padj < 0.05 & res2_cd8$log2FoldChange > 0 & !(is.na(res2_cd8$padj)),])
ndown_psc <- nrow(res1_cd8[res1_cd8$padj < 0.05 & res1_cd8$log2FoldChange < 0 & !(is.na(res1_cd8$padj)),])
ndown_uc <- nrow(res2_cd8[res2_cd8$padj < 0.05 & res2_cd8$log2FoldChange < 0 & !(is.na(res2_cd8$padj)),])

results_cd8 <- data.frame(comparison = resultsNames(dds_cd4)[2:3],
                          no.up = c(nup_psc, nup_uc),
                          no.down = c(ndown_psc, ndown_uc))

knitr::kable(results_cd8)
```

## Where are these genes expressed

```{r expression of cd8 genes across cell types, echo=FALSE, message=FALSE}
cd8_gois <- rownames(res1_cd8[res1_cd8$padj < 0.05 & !(is.na(res1_cd8$padj)),])


# Violin plot each goi separately as it is a bit annoying
# to do the default seurat one
cd8_gois_p <- list()
for (i in cd8_gois){
  p <- VlnPlot(immune_cells, features = i, pt.size = 0) +
  scale_fill_manual(values = immune_cell_anno_colors) +
  theme(legend.position = "none") 
  cd8_gois_p[[i]] <- p
  ggsave(p, file = paste0("export.dir/Figures/cd8_", i, "_celltypes.pdf"), height=4, width=4)
}

gridExtra::grid.arrange(grobs = cd8_gois_p, nrow=3)


```



## Feature boxplots of the changed genes in cd8 memory cells

```{r featurebox gois cd8 memory, echo=FALSE, message=FALSE}

vst_cd8 <- as.data.frame(assay(vst(dds_cd8)))

gois_cd8 <- OCMSutility::featurebox(vst_cd8, coldata_cd8, features=genes_psc_cd8, group_by = "disease_type") +
              theme_classic() +
              facet_wrap(~feature, scales="free") +
              scale_color_manual(values = rep(immune_cell_anno_colors[[3]], 3)) +
  theme(legend.position = "none")


ggsave(gois_cd8, file = "export.dir/Figures/cd4_memory_gois.pdf", height=6, width=6)

gois_cd8
```


## XCL1+ cells

```{r xcl1, echo=FALSE, message=FALSE}

ppos_xcl1 <- get_ppos(cd8, "XCL1")
ppos_xcl1$individual <- gsub("-", "\\.", ppos_xcl1$individual)
ppos_xcl1$disease_type <- meta_cd8[ppos_xcl1$individual,]$disease_type

xcl1_pos_p <- ggplot(ppos_xcl1, aes(x=disease_type, y=ppos, color=disease_type)) +
  geom_boxplot(outlier.alpha=0) +
  geom_jitter(height=0, width=0.1) +
  scale_color_manual(values = unlist(group_colors)) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2), c(1,3), c(2,3))) +
  theme(legend.position = "none")

ggsave(xcl1_pos_p, file = "export.dir/Figures/XCL1_positivity.pdf", height=2.5, width=2.5)
xcl1_pos_p
```

## CD8 FGSEA


```{r cd8 fgsea, echo=FALSE, message=FALSE}

set.seed(10)

cd8_genes <- res1_cd8$log2FoldChange
names(cd8_genes) <- rownames(res1_cd8)

# gene set enrichments
cd8_fgsea <- fgsea::fgsea(pathways, cd8_genes, minSize=50)
cd8_fgsea <- cd8_fgsea[order(cd8_fgsea$padj, decreasing = FALSE),]

cd8_pathways_sig <- cd8_fgsea[cd8_fgsea$padj < 0.05,]

cd8_top10 <- cd8_pathways_sig[1:10,]
cd8_top10$pathway <- factor(cd8_top10$pathway, levels = cd8_top10$pathway)
cd8_top10_p <- ggplot(cd8_top10, aes(x = pathway, y=NES, fill=-log10(padj))) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low="snow2", high=immune_cell_anno_colors[[3]]) +
  theme_classic() +
  coord_flip()
  
ggsave(cd8_top10_p, file = "export.dir/Figures/cd8_memory_pathway_barplot.pdf", height=2.5, width=10)

cd8_top10_p

```


## Compare significant pathways between CD4 and CD8 T cells

```{r compare pathways T cells, echo=FALSE, message=FALSE}

set.seed(10)

pathways_cd4 <- cd4_fgsea[cd4_fgsea$padj < 0.05,]$pathway
pathways_cd8 <- cd8_fgsea[cd8_fgsea$padj < 0.05,]$pathway

library(VennDiagram)

to_venn_t_cells <- list("CD4 pathways" = pathways_cd4,
                        "CD8_pathways" = pathways_cd8)

t_cell_pathway_overlap <- VennDiagram::venn.diagram(to_venn_t_cells,
                                                    filename = NULL,
                                                    fill = c(immune_cell_anno_colors[[1]],
                                                             immune_cell_anno_colors[[3]]),
                                                    alpha = 0.3,
                                                    sub.fontfamily = "sans",
                                                    )
pdf("export.dir/Figures/overlap_t_cell_pathways.pdf", height=2, width=2.5)
grid.draw(t_cell_pathway_overlap)
dev.off()                          

grid.draw(t_cell_pathway_overlap)

```


## Shared leading edge genes - GO_OXIDATIVE_PHOSPHORYLATION

Interestingly a number of pathways overlap between CD8 and CD4 T cells and they seem to be linked to mitochondria and metabolism, particularly OXPHOS. Here we see if the same genes are changed in these cell types.

```{r leading edge oxphos, echo=FALSE, message=FALSE}

cd4_le <- unlist(cd4_fgsea[cd4_fgsea$pathway == "GO_OXIDATIVE_PHOSPHORYLATION",]$leadingEdge)
cd8_le <- unlist(cd8_fgsea[cd8_fgsea$pathway == "GO_OXIDATIVE_PHOSPHORYLATION",]$leadingEdge)

# very shared genes
to_venn_oxphos <- list("CD4 OXPHOS" = cd4_le,
                       "CD8 OXPHOS" = cd8_le)

t_cell_oxphos_overlap <- VennDiagram::venn.diagram(to_venn_oxphos,
                                                    filename = NULL,
                                                    fill = c(immune_cell_anno_colors[[1]],
                                                             immune_cell_anno_colors[[3]]),
                                                    alpha = 0.3,
                                                    sub.fontfamily = "sans",
                                                    )
pdf("export.dir/Figures/overlap_t_cell_oxphos.pdf", height=2, width=2.5)
grid.draw(t_cell_oxphos_overlap)
dev.off()                          

grid.draw(t_cell_oxphos_overlap)


# Heatmap these genes in different T-cells
gois_oxphos <- intersect(cd4_le, cd8_le)
vst_cd4_oxphos <- ScaleData(vst_cd4[gois_oxphos,])
vst_cd8_oxphos <- ScaleData(vst_cd8[gois_oxphos,])

# sort out annotations for the disease states
disease_anno <- HeatmapAnnotation("Disease type" = coldata$disease_type,
                                  col = list("Disease type" = c("HC" = "grey",
                                                                "PSCUC" = blues9[6],
                                                                "UC" = "red3")))

heat_colors <- colorRamp2(breaks = c(-2,0,2), colors = c("blue3", "white", "red3"))
h1 <- Heatmap(vst_cd4_oxphos,
              column_split = coldata$disease_type,
              col = heat_colors,
              cluster_columns = FALSE,
              top_annotation = disease_anno,
              show_column_names = FALSE,
              rect_gp = gpar(col="white", lwd=0.5),
              row_names_gp = gpar(fontsize=4))

h2 <- Heatmap(vst_cd8_oxphos,
              column_split = coldata$disease_type,
              col = heat_colors,
              cluster_columns = FALSE,
              top_annotation = disease_anno,
              show_column_names = FALSE,
              rect_gp = gpar(col="white", lwd=0.5),
              row_names_gp = gpar(fontsize=4))
              
# write out the gois oxphos
write.table(data.frame(oxphos_leading = gois_oxphos),
            file = "export.dir/oxphos_leading_edge.tsv",
            sep = "\t",
            row.names = FALSE,
            quote = FALSE)

pdf("export.dir/Figures/heatmap_oxphos_t_cells.pdf", height=5, width=5)
h1 + h2
dev.off()
h1 + h2

```


## B (plasma) IgA producing cell pseudobulk

```{r iga cell clustering pseudobulk, echo=FALSE, message = FALSE}

set.seed(10)

iga <- subset(immune_cells, idents = "Plasma_Cell_IgA")

iga <- ScaleData(iga, features = rownames(iga))
iga <- FindVariableFeatures(iga, selection.method = "vst", nfeatures = 2000)
iga <- RunPCA(iga, features = VariableFeatures(object = iga))

# harmony integration: by individual = orig.ident
iga <- iga %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
iga <- FindNeighbors(iga, reduction="harmony", dims = 1:30)
iga <- FindClusters(iga, resolution = 0.8, algorithm = 4)
iga <- RunUMAP(iga, dims = 1:30, reduction = "harmony")

DimPlot(iga, reduction = "umap")
```

Cluster 6 looks a bit different to the others - check for marker genes that define cluster 6.


```{r iga markers, echo = FALSE, message = FALSE}

iga_markers <- FindAllMarkers(iga, only.pos = TRUE)

knitr::kable(head(iga_markers[iga_markers$cluster == 6,]))

# T cells!
dplot_iga_cells <- DotPlot(iga, features = c(b_cell_genes, "CD3E", "CD3D"))

dplot_iga_cells
```

Looks like T Cell contamination so will remove this cluster.

```{r iga filtering, echo = FALSE, message = FALSE}

iga <- subset(iga, ident = 6, invert = TRUE)

# Write out the iga object
saveRDS(iga, file = "export.dir/Immune/ileum_iga.Rds")

```


```{r iga pseudobulk, echo=FALSE, message=FALSE}

# get number of cells per sample
knitr::kable(as.data.frame(table(iga@meta.data$orig.ident, iga@meta.data$annotation)))

# remove G-SCI7T081-SCI5T081
iga <- subset(iga, subset = orig.ident == "G-SCI7T081-SCI5T081", invert = TRUE)


agg_iga <- AggregateExpression(iga, group.by = c("orig.ident"))
agg_iga <- as.data.frame(agg_iga)
colnames(agg_iga) <- gsub("RNA\\.", "", colnames(agg_iga))

# remove immmunoglobulin genes
agg_iga <- agg_iga[grep("^IG[K|H|L].*-.*", rownames(agg_iga), invert=TRUE),]


meta_iga <- metadata
rownames(meta_iga) <- gsub("-", "\\.", rownames(meta_iga))
coldata_iga <- meta_iga[colnames(agg_iga),]

dds_iga <- DESeqDataSetFromMatrix(agg_iga, colData = coldata_iga, design=~disease_type)
dds_iga$disease_type <- relevel(dds_iga$disease_type, ref = "HC")
dds_iga <- dds_iga[rowSums(counts(dds_iga) >= 10) >= 4,]
dds_iga <- DESeq2::DESeq(dds_iga)


res1_iga <- lfcShrink(dds_iga,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds_iga, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2_iga <- lfcShrink(dds_iga,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds_iga, name = "disease_type_UC_vs_HC"),
                  type = "normal")

genes_psc_iga <- rownames(res1_iga[res1_iga$padj < 0.05 & !is.na(res1_iga$padj),])
genes_uc_iga <- rownames(res2_iga[res2_iga$padj < 0.05 & !is.na(res2_iga$padj),])

iga_volc <- volcano_plot(res1_iga, use="padj") +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = immune_cell_anno_colors[[2]]) +
  theme(legend.position = "none")

iga_volc_uc <- volcano_plot(res2_iga, use="padj") +
  ggtitle("UC vs. Healthy") +
  scale_color_manual(values = immune_cell_anno_colors[[2]]) +
  theme(legend.position = "none")

ggsave(iga_volc, file = "export.dir/Figures/iga_volcano_psc.pdf", height = 3, width=3)

iga_volc


```

## Feature plots of most differentially expressed genes

```{r featureplots iga, echo = FALSE, message = FALSE}

FeaturePlot(immune_cells, features = c("AMPD1", "CADM1", "CRTAM", "NPAS3", "P2RX4"))

```

The two most prominent genes are restricted to the plasma B cells.


```{r table of differences iga, echo=FALSE, message=FALSE}

nup_psc <- nrow(res1_iga[res1_iga$padj < 0.05 & res1_iga$log2FoldChange > 0 & !(is.na(res1_iga$padj)),])
nup_uc <- nrow(res2_iga[res2_iga$padj < 0.05 & res2_iga$log2FoldChange > 0 & !(is.na(res2_iga$padj)),])
ndown_psc <- nrow(res1_iga[res1_iga$padj < 0.05 & res1_iga$log2FoldChange < 0 & !(is.na(res1_iga$padj)),])
ndown_uc <- nrow(res2_iga[res2_iga$padj < 0.05 & res2_iga$log2FoldChange < 0 & !(is.na(res2_iga$padj)),])

results_iga <- data.frame(comparison = resultsNames(dds_iga)[2:3],
                          no.up = c(nup_psc, nup_uc),
                          no.down = c(ndown_psc, ndown_uc))

knitr::kable(results_iga)
```

## Feature boxplots of the changed genes in IgA plasma cells

```{r featurebox gois iga, echo=FALSE, message=FALSE}

vst_iga <- as.data.frame(assay(vst(dds_iga)))

gois_iga <- OCMSutility::featurebox(vst_iga, coldata_iga, features=genes_psc_iga, group_by = "disease_type") +
              theme_classic() +
              facet_wrap(~feature, scales="free") +
              scale_color_manual(values = rep(immune_cell_anno_colors[[2]], 3)) +
  theme(legend.position = "none")

ggsave(gois_iga, file = "export.dir/Figures/iga_gois_uc.pdf", height=6, width=6)

gois_iga


```

## Plasma Cell IgA FGSEA


```{r iga fgsea, echo=FALSE, message=FALSE}

set.seed(10)

iga_genes <- res1_iga$log2FoldChange
names(iga_genes) <- rownames(res1_iga)

# gene set enrichments
iga_fgsea <- fgsea::fgsea(pathways, iga_genes, minSize=50)
iga_fgsea <- iga_fgsea[order(iga_fgsea$padj, decreasing = FALSE),]
iga_pathways_sig <- iga_fgsea[iga_fgsea$padj < 0.05,]

iga_top10 <- iga_pathways_sig[1:10,]
iga_top10$pathway <- factor(iga_top10$pathway, levels = iga_top10$pathway)
iga_top10_p <- ggplot(iga_top10, aes(x = pathway, y=NES, fill=-log10(padj))) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low="snow2", high=immune_cell_anno_colors[[2]]) +
  theme_classic() +
  coord_flip()
  
ggsave(iga_top10_p, file = "export.dir/Figures/iga_memory_pathway_barplot.pdf", height=2.5, width=10)

iga_top10_p

iga_top10[1,]$leadingEdge
```

Fgsea is showing ribosomal proteins being downregulated in PSC-UC. Would be worth checking for quality etc between groups to see if this is an artificact of data quality etc.

## QC metrics

```{r qc because of ribosomal, echo = FALSE, message = FALSE}

VlnPlot(iga, features = "percent.mt", group.by = "disease_type")
VlnPlot(iga, features = "nCount_RNA", group.by = "disease_type")

# There are a lot more cells in the PSC-UC group. Maybe
# we shoudl rarefy the cells
iga_small <- iga
Idents(iga_small) <- iga_small@meta.data$orig.ident
iga_small <- subset(iga_small, downsample = 20)
VlnPlot(iga_small, features = "percent.mt", group.by = "disease_type")

# get number of cells per sample
knitr::kable(as.data.frame(table(iga_small@meta.data$orig.ident, iga_small@meta.data$annotation)))

agg_iga_small <- AggregateExpression(iga_small, group.by = c("orig.ident"))
agg_iga_small <- as.data.frame(agg_iga_small)
colnames(agg_iga_small) <- gsub("RNA\\.", "", colnames(agg_iga_small))

# remove immmunoglobulin genes
agg_iga <- agg_iga[grep("^IG[K|H|L].*-.*", rownames(agg_iga), invert=TRUE),]


meta_iga_small <- metadata
rownames(meta_iga_small) <- gsub("-", "\\.", rownames(meta_iga_small))
coldata_iga_small <- meta_iga_small[colnames(agg_iga_small),]

dds_iga_small <- DESeqDataSetFromMatrix(agg_iga_small, colData = coldata_iga_small, design=~disease_type)
dds_iga_small$disease_type <- relevel(dds_iga_small$disease_type, ref = "HC")
dds_iga_small <- dds_iga_small[rowSums(counts(dds_iga_small) >= 10) >= 4,]
dds_iga_small <- DESeq2::DESeq(dds_iga_small)


res1_iga_small <- lfcShrink(dds_iga_small,
                  contrast = c("disease_type", "PSCUC", "HC"),
                  res = results(dds_iga_small, name = "disease_type_PSCUC_vs_HC"),
                  type = "normal")

res2_iga_small <- lfcShrink(dds_iga_small,
                  contrast = c("disease_type", "UC", "HC"),
                  res = results(dds_iga_small, name = "disease_type_UC_vs_HC"),
                  type = "normal")

genes_psc_iga_small <- rownames(res1_iga_small[res1_iga_small$padj < 0.05 & !is.na(res1_iga_small$padj),])
genes_uc_iga_small <- rownames(res2_iga_small[res2_iga_small$padj < 0.05 & !is.na(res2_iga_small$padj),])

iga_volc_small <- volcano_plot(res1_iga_small, use="padj") +
  ggtitle("PSC-UC vs. Healthy") +
  scale_color_manual(values = immune_cell_anno_colors[[2]]) +
  theme(legend.position = "none")

iga_volc_small

# Correlate fold changes from both analyses
common_genes <- intersect(rownames(res1_iga_small), rownames(res1_iga))
to_cor_iga <- data.frame(all_data = res1_iga[common_genes,]$log2FoldChange,
                         small_data = res1_iga_small[common_genes,]$log2FoldChange,
                         gene = common_genes)

# show ribosomal protein genes
to_cor_iga$rp <- ifelse(to_cor_iga$gene %in% unlist(iga_top10[1,]$leadingEdge), "RP", NA)

ggplot(to_cor_iga, aes(x = all_data, y = small_data, color = rp)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor()

```

They look about the same in terms of robosomal proteins. Could be a signal or could potentially be an artifact still.

## Where are these genes expressed

```{r expression of iga genes across cell types, echo=FALSE, message=FALSE}
iga_gois <- rownames(res1_iga[res1_iga$padj < 0.05 & !(is.na(res1_iga$padj)),])


# Violin plot each goi separately as it is a bit annoying
# to do the default seurat one
iga_gois_p <- list()
for (i in iga_gois){
  p <- VlnPlot(immune_cells, features = i, pt.size = 0) +
  scale_fill_manual(values = immune_cell_anno_colors) +
  theme(legend.position = "none") 
  iga_gois_p[[i]] <- p
  ggsave(p, file = paste0("export.dir/Figures/", i, "_celltypes.pdf"), height=4, width=4)
}

gridExtra::grid.arrange(grobs = iga_gois_p, nrow=3)


```


## Featureplots of AMPD1 and CADM1

```{r iga featureplots, echo = FALSE, message = FALSE}

FeaturePlot(immune_cells, feature = c("AMPD1", "CADM1"), label = TRUE, split.by = "disease_type")

FeatureScatter(iga, feature1 = "CADM1", feature2 = "AMPD1", split.by = "disease_type")

```


## Write out h5ad file for cellphoneDB

```{r write h5ad, echo=FALSE, message=FALSE}

# convert to h5ad
library(SeuratDisk)

immune_cells3 <- immune_cells

immune_cells3[["RNA"]] <- as(object = immune_cells3[["RNA"]], Class = "Assay")
immune_cells3 <- NormalizeData(immune_cells3)


# save the whole thing
SaveH5Seurat(immune_cells3, filename = "export.dir/immune_cells.h5Seurat", overwrite = TRUE)
Convert("export.dir/immune_cells.h5Seurat", dest = "h5ad", overwrite = TRUE)

# output metadata file for cellphonedb
meta_barcodes <- data.frame("barcode_sample" = rownames(immune_cells3@meta.data),
                            "cell_type" = immune_cells3@meta.data$annotation)

write.table(meta_barcodes, file = "export.dir/meta_barcodes.txt",
            sep="\t",
            row.names = FALSE,
            quote=FALSE)

```


