---
title: "Clustering analysis: Forced cells 10000"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args=list(png=list(type="cairo")))
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")
options(bitmapType = "cairo")

# reticulate stuff
reticulate::use_virtualenv('~/devel/venv/Python-3.10.8-GCCcore-12.2.0/skylake/', required=TRUE)
```

```{r load libraries, echo=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(decontX)
library(flexmix)
library(scriabin)
library(OCMSutility)
library(circlize)

# create Clustering analysis dir
if (!(dir.exists("export.dir/Clustering"))){
  dir.create("export.dir/Clustering")
}
```


```{r set colors, echo=FALSE, message=FALSE, warning=FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```


# Overview

This document describes clustering and futher anlysis of the scRNA-seq data that have already had cells filtered out based on basic QC metrics (nfeatures, ncount, %mt etc, in QC_Force_Cells_Refactor.Rmd). I am initially starting with data from the ileum which will be clustered separately from data that we have from the caecum.

## Ileum

```{r get ileum sobjs, echo=FALSE, message=FALSE}

# read in metadata
metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)


# read in the filtered sobj
sobjs_filtered <- readRDS("export.dir/QC/all_tissues_filtered.Rds")

# just get the ileum data
sobjs_ileum <- list()
for (i in 1:length(sobjs_filtered)){
  sobj <- sobjs_filtered[[i]]
  id <- as.character(sobj@meta.data$orig.ident)[1]
  tissue <- metadata[id,]$tissue[1]
   if (tissue == "Ileum"){
      sobjs_ileum[[i]] <- sobj
  }
}
sobjs_ileum <- Filter(Negate(is.null), sobjs_ileum)
```


### Overview of ileum data {.tabset .tabset-pills}

#### Feature scatter plots

```{r ileum overview, echo=FALSE, message=FALSE, fig.height=20, fig.width=20}

feature_scatter_plots <- list()
for (i in 1:length(sobjs_ileum)){
  p <-FeatureScatter(sobjs_ileum[[i]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
  feature_scatter_plots[[i]] <- p
}

cowplot::plot_grid(plotlist = feature_scatter_plots, nrow=4)

```


```{r merge sobjs across the ileum , echo=FALSE, message=FALSE}

merged_ileum <- merge(sobjs_ileum[[1]],
                      y = c(sobjs_ileum[[2]],
                      sobjs_ileum[[3]],
                      sobjs_ileum[[4]],
                      sobjs_ileum[[5]],
                      sobjs_ileum[[6]],
                      sobjs_ileum[[7]],
                      sobjs_ileum[[8]],
                      sobjs_ileum[[9]],
                      sobjs_ileum[[10]],
                      sobjs_ileum[[11]],
                      sobjs_ileum[[12]]))

```

## Median nummber of read counts per cell

```{r median read count, echo = FALSE, message=FALSE}

med_count <- merged_ileum@meta.data %>%
  group_by(orig.ident) %>%
  summarise(median_nCount_RNA = median(nCount_RNA)) %>%
  as.data.frame()

knitr::kable(med_count)

```

#### Variable features

We are using 2000 variable features and show the top 10 labelled in the plot.

```{r variable features, echo=FALSE, message=FALSE, fig.width=7, fig.height=7}

# find variable features
merged_ileum <- NormalizeData(merged_ileum)
merged_ileum <- FindVariableFeatures(merged_ileum, selection.method = "vst", nfeatures = 2000)

top10 <- head(VariableFeatures(merged_ileum), 10)

plot1 <- VariableFeaturePlot(merged_ileum)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)

gridExtra::grid.arrange(plot1, plot2, nrow=2)
```

### Dimensionality reduction - PCA

```{r dim reduction ileum, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}

set.seed(10)

if (!file.exists("export.dir/Clustering/ileum_harmonised_force_cells.rds")){
  all.genes <- rownames(merged_ileum)
  merged_ileum <- ScaleData(merged_ileum, features = all.genes)
  merged_ileum <- FindVariableFeatures(merged_ileum, selection.method = "vst", nfeatures = 2000)
  merged_ileum <- RunPCA(merged_ileum, features = VariableFeatures(object = merged_ileum))
 
  # harmony integration: by individual = orig.ident
  merged_ileum <- merged_ileum %>% 
  RunHarmony("orig.ident", plot_convergence = TRUE)
  saveRDS(merged_ileum, file = "export.dir/Clustering/ileum_harmonised_force_cells.rds")
}

merged_ileum <- readRDS("export.dir/Clustering/ileum_harmonised_force_cells.rds")

p_pca <- DimPlot(merged_ileum, reduction = "harmony", group.by = "orig.ident") + NoLegend()
p_elbow <- ElbowPlot(merged_ileum, ndims = 40)

gridExtra::grid.arrange(p_pca, p_elbow, nrow=1)
```

### UMAP 

We are using fairly low resolution clustering (0.5) with leiden clustering. We don't have a huge number of cells so I don't want truly fine grained clusters at this stage particularly - especially as there are lots of different types of cells in these samples.

```{r umap ileum, echo=FALSE, message=FALSE, warning=FALSE, fig.height=12, fig.width=12}

set.seed(10)

merged_ileum <- FindNeighbors(merged_ileum, reduction="harmony", dims = 1:30)
merged_ileum <- FindClusters(merged_ileum, resolution = 0.5, algorithm = 4)
merged_ileum <- RunUMAP(merged_ileum, dims = 1:30, reduction = "harmony")

# add disease into metadata
merged_ileum@meta.data$disease_type <- metadata[as.character(merged_ileum@meta.data$orig.ident),]$disease_type

clusters_plot <- DimPlot(merged_ileum, reduction = "umap", label = TRUE)
clusters_by_sample <- DimPlot(merged_ileum, reduction = "umap", group.by = "orig.ident")
clusters_by_disease <- DimPlot(merged_ileum, reduction = "umap", group.by = "disease_type")

gridExtra::grid.arrange(clusters_plot,
                        clusters_by_sample,
                        clusters_by_disease + scale_color_manual(values = group_colors),
                        nrow=2)
```

### Number of cells in the clusters {.tabset .tabset-pills}

Here we look at how many cells there are in each cluster and how many individuals contribute to each cluster. In an ideal world every donor would contribute cells to each cluster but in reality there are different numbers of cells for each donor and so some may be missing from some clusters.

```{r ncells per cluster, echo=FALSE, message=FALSE}

Idents(merged_ileum) <- merged_ileum@meta.data$seurat_clusters
clusters <- unique(merged_ileum@meta.data$seurat_clusters)
ncells_cluster = list()
ninds_cluster = list()

for (i in 1:length(clusters)){
  flog.info(paste0("Counting cells for cluster ", clusters[i]))
  dat <- subset(merged_ileum, idents = clusters[i])
  n <- ncol(dat)
  nind <- length(unique(as.character(dat@meta.data$orig.ident)))
  ncells_cluster[[i]] <- n
  ninds_cluster[[i]] <- nind
}
names(ncells_cluster) <- clusters
names(ninds_cluster) <- clusters

ncells_cluster <- as.data.frame(t(as.data.frame(ncells_cluster)))
ninds_cluster <- as.data.frame(t(as.data.frame(ninds_cluster)))

ncells_cluster$cluster <- gsub("X", "", rownames(ncells_cluster))
colnames(ncells_cluster) <- c("ncells", "cluster")
ninds_cluster$cluster <- gsub("X", "", rownames(ninds_cluster))
colnames(ninds_cluster) <- c("ninds", "cluster")

# order by cluster size
ncells_cluster <- ncells_cluster[order(ncells_cluster$ncells, decreasing = FALSE),]
ncells_cluster$cluster <- factor(ncells_cluster$cluster, levels = ncells_cluster$cluster)
ninds_cluster$cluster <- factor(ninds_cluster$cluster, levels = ncells_cluster$cluster)

ncells_p <- ggplot(ncells_cluster, aes(x=cluster, y=ncells/1000, size=ncells)) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0.1, linetype = "dashed") +
  ylab("No. cells (k)") +
  ylim(c(0,3))

# How different individuals make up the clusters i.e. each cluster by
# number of individuals represented. Really want all 11 to be present
# for each cluster
ninds_p <- ggplot(ninds_cluster, aes(x=cluster, y=ninds, size=ninds)) +
  geom_point() +
  theme_classic()

gridExtra::grid.arrange(ncells_p, ninds_p)

```

## Find markers

Here we use FindAllMarkers() to identify the marker genes for each cluster. This will be used downstream in conjunction with manual annotation to annotate the clusters.

```{r find markers, echo=FALSE, message=FALSE, warning=FALSE, fig.height=20, fig.width=20}

# This can take a while so think will write out
# and just read in if required
if (!file.exists("export.dir/Clustering/ileum_harmonised_force_cell_markers.tsv")){
  merged_ileum <- JoinLayers(merged_ileum)
  markers <- FindAllMarkers(merged_ileum, only.pos = TRUE)
  markers <- markers %>%
             group_by(cluster) %>%
             dplyr::filter(avg_log2FC > 1)
  
  write.table(markers,
              file = "export.dir/Clustering/ileum_harmonised_force_cell_markers.tsv",
              sep="\t",
              row.names=FALSE,
              quote=FALSE)
}

markers <- read.csv("export.dir/Clustering/ileum_harmonised_force_cell_markers.tsv",
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    sep="\t")  
top5 <- markers %>%
    group_by(cluster) %>%
    top_n(n = 5, wt = pct.1)

ggplot(top5, aes(x = gene, y=avg_log2FC, fill = pct.1)) +
    geom_bar(stat="identity") +
    theme_classic() +
    facet_wrap(~cluster, scales="free") +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Broad annotation 

Here we just look at some common markers used to define specific cell populations - gives a sense of where we need to go.


```{r broad annotation, echo=FALSE, message=FALSE, warning=FALSE, fig.height=20, fig.width=20}

FeaturePlot(merged_ileum, reduction = "umap", features = c("VIL1",
                                                           "FABP6",
                                                           "PTPRC",
                                                           "CD3D",
                                                           "CD19",
                                                           "CD79A",
                                                           "CD68",
                                                           "PDPN",
                                                           "PDGFRA",
                                                           "CD4")) 


```

### More specific annotation

I am using the definitions from the [TAURUS paper](https://www.nature.com/articles/s41590-024-01994-8/figures/7) to define the different subsets. We will not be as fine-grained as their definitions as we have fewer cells in general but we can make some inferences based on their markers.


```{r specific annotation, echo=FALSE, message=FALSE, fig.width = 15}

# T cell markers
t_cell_genes <- c(
             "PTPRC", 
             "CD3D",
             "CD4",
             "CD8A",
             "FOXP3",
             "IL17A",
             "TNF",
             "IFNG",
             "KLRG1",
             "SELL",
             "IL7R",
             "LEF1",
             "CD69",
             "CTLA4",
             "TCF7",
             "CCR7",
             "TRDC",
             "NKG7",
             "GZMK",
             "GZMA")

b_cell_genes <- c("CD19",
                  "CD79A",
                  "CD27",
                  "MZB1",
                  "IGHA1",
                  "IGHA2",
                  "IGHG1",
                  "IGHG2",
                  "IGHD",
                  "CCL3",
                  "CCL4",
                  "MEF2B",
                  "CXCR4",
                  "CD83")


mono_mac_genes <- c("CD68",
                    "CD14",
                    "CD74",
                    "HLA-DRA",
                    "S100A8",
                    "S100A9",
                    "IL1B",
                    "TCF4",
                    "C1QA",
                    "LAMP3")

stromal_genes <- c("PDPN",
                   "PDGFRA",
                   "PDGFRB",
                   "CALD1",
                   "IGFBP7",
                   "PECAM1",
                   "PHGR1",
                   "THY1",
                   "LYVE1",
                   "ACKR1")
     
epithelial_cell_genes <- c("VIL1",  
                          "FABP6",
                          "FABP1",
                          "KRT19",
                          "KRT8",
                          "APOB",
                          "DEFA5",
                          "DEFA6",
                          "REG3A",
                          "SLC10A2",
                          "GSTA2",
                          "GSTA1",
                          "LGR5",
                          "OLFM4",
                          "RGMB",
                          "SMOC2",
                          "ASCL2",
                          "MKI67",
                          "PCNA",
                          "TOP2A",
                          "CCNA2",
                          "POU2F3",
                          "GFI1B",
                          "TRPM5",
                          "DCLK1",
                          "CLCA1",
                          "MUC5AC",
                          "SPINK4",
                          "TFF3",
                          "AGR2",
                          "CPE",
                          "NEUROD1",
                          "BEST4",
                          "OTOP2")


mast_cell_genes <- c( "KIT",
                     "FCER1A")       
DotPlot(merged_ileum, 
        cols = colorRampPalette(c("blue", "red"))(2),
        features=c(t_cell_genes,
                   b_cell_genes,
                   mono_mac_genes,
                   stromal_genes,
                   epithelial_cell_genes,
                   mast_cell_genes),
        cluster.idents = TRUE) +
  theme(axis.text.x = element_text(angle=45, hjust=1, size = 8), axis.text.y = element_text(size = 8)) 
  
```


### Annotations

Here I am putting the broad annotations on the clusters. Cell types will be split off later on because it is unlikely that we can get enough resolution given the diverse cell types at this point.


```{r annotation, echo=FALSE, message=FALSE, fig.height=7, fig.width = 7}

new_labels <- c("Enterocyte",
                "CD4_T_Cell",
                "B_cell",
                "CD8_T_cell",
                "B_Cell_IgA",
                "Mono_Mac",
                "Ribosomal_enriched",
                "Stroma",
                "Epithelial_Stem",
                "Stroma_2",
                "B_Cell_2",
                "Undetermined",
                "Gamma_Delta_T_Cell",
                "Mono_Mac_Inf",
                "Goblet_Cell",
                "Enterocyte_2",
                "Tuft_Cell",
                "Mast_Cell",
                "Enterocyte_BEST4_OTOP2",
                "Undetermined",
                "Endothelium",
                "Endothelium_Lymphatic",
                "Enterocyte_3",
                "B_Cell_IgG",
                "Paneth_Cell",
                "B_Cell_Activated")
                
                
names(new_labels) <- levels(merged_ileum)
merged_ileum <- RenameIdents(merged_ileum, new_labels)


```

```{r umap annotations, echo=FALSE, message=FALSE, fig.height = 7, fig.width = 7}

library(OCMSutility)

set.seed(6)
annotation_colors <- sample(OCMSutility::get_palette(n = 40), 25)
names(annotation_colors) <- unique(Idents(merged_ileum))
dplot_all <- DimPlot(merged_ileum, label = TRUE, repel = TRUE) +
               scale_color_manual(values = annotation_colors) +
               ggtitle(paste0("No. cells = ", nrow(merged_ileum@meta.data))) +
               theme(legend.position = "none")
dplot_all

# write out
saveRDS(merged_ileum, file = "export.dir/Clustering/ileum_harmonised_force_cells_annotated.Rds")

# Figure 1A
if (!dir.exists("export.dir/Figures")){
  dir.create("export.dir/Figures")
  }

ggsave(dplot_all, file = "export.dir/Figures/Figure_1_umap.pdf",
       height = 5,
       width = 5)


```

## Heatmap the markers we used

```{r heatmap markers, echo = FALSE, message = FALSE}

merged_ileum <- readRDS("export.dir/Clustering/ileum_harmonised_force_cells_annotated.Rds")

# row_annotation
row_annotation = data.frame("gene" = c(t_cell_genes,
                                       mono_mac_genes,
                                       b_cell_genes,
                                       stromal_genes,
                                       epithelial_cell_genes,
                                       mast_cell_genes),
                            "Type" = c(rep("T Cell", length(t_cell_genes)),
                                       rep("Mono-Mac", length(mono_mac_genes)),
                                       rep("B Cell", length(b_cell_genes)),
                                       rep("Stroma", length(stromal_genes)),
                                       rep("Epithelium", length(epithelial_cell_genes)),
                                       rep("Mast Cell", length(mast_cell_genes)))
                            )

# use normalised and scaled counts for matrix
scaled_all <- as.data.frame(AggregateExpression(merged_ileum, group.by = "ident")[[1]])
scaled_all <- ScaleData(NormalizeData(scaled_all))
scaled_all <- scaled_all[row_annotation$gene,]

# row annotation
row_anno_colors <- OCMSutility::get_palette(60)[55:60]
names(row_anno_colors) <- unique(row_annotation[,2])
row_anno <- rowAnnotation("Cell Type" = row_annotation[,2],
                          col = list("Cell Type" = row_anno_colors))

# heatmap color scale
heat_col <- colorRamp2(c(-3, 0, 3), c("blue4", "white", "red3"))

neword <- c("Enterocyte",
            "Enterocyte-2",
            "Enterocyte-3",
            "Enterocyte-BEST4-OTOP2",
            "Epithelial-Stem",
            "Goblet-Cell",
            "Tuft-Cell",
            "Paneth-Cell",
            "Stroma",
            "Stroma-2",
            "Endothelium",
            "Endothelium-Lymphatic",
            "B-cell",
            "B-Cell-2",
            "B-Cell-Activated",
            "B-Cell-IgA",
            "B-Cell-IgG",
            "Mono-Mac",
            "Mono-Mac-Inf",
            "Ribosomal-enriched",
            "CD4-T-Cell",
            "CD8-T-cell",
            "Gamma-Delta-T-Cell",
            "Mast-Cell")

scaled_all <- scaled_all[,neword]

h_markers <- Heatmap(scaled_all,
        right_annotation = row_anno,
        column_split = factor(colnames(scaled_all), levels = colnames(scaled_all)),
        row_split = factor(row_annotation$Type, levels = c("Epithelium",
                                                           "Stroma",
                                                           "Mono-Mac",
                                                           "B Cell",
                                                           "T Cell",
                                                           "Mast Cell")),
        use_raster=FALSE,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_row_names = TRUE,
        row_names_gp = gpar(fontsize = 6),
        column_gap = unit(0.5, "mm"),
        row_gap = unit(2, "mm"),
        col = heat_col)

h_markers

# Figure 1B
pdf("export.dir/Figures/Heatmap_markers.pdf",
    width = 7,
    height = 10)
    h_markers
dev.off()

```

## Number of cells in each cluster

```{r number of cells in each cluster, echo = FALSE, message = FALSE}

clusters <- unique(Idents(merged_ileum))
ncells_cluster = list()
ninds_cluster = list()

for (i in 1:length(clusters)){
  flog.info(paste0("Counting cells for cluster ", clusters[i]))
  dat <- subset(merged_ileum, idents = clusters[i])
  n <- ncol(dat)
  nind <- length(unique(as.character(dat@meta.data$orig.ident)))
  ncells_cluster[[i]] <- n
  ninds_cluster[[i]] <- nind
}
names(ncells_cluster) <- clusters

ncells_cluster <- as.data.frame(t(as.data.frame(ncells_cluster)))

ncells_cluster$cluster <- gsub("X", "", rownames(ncells_cluster))
colnames(ncells_cluster) <- c("ncells", "cluster")

# order by cluster size
ncells_cluster <- ncells_cluster[order(ncells_cluster$ncells, decreasing = FALSE),]
ncells_cluster$cluster <- factor(ncells_cluster$cluster, levels = ncells_cluster$cluster)

ncells_p_annotated <- ggplot(ncells_cluster, aes(x=cluster, y=ncells/1000, size=ncells, color=cluster)) +
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0.1, linetype = "dashed") +
  ylab("No. cells (k)") +
  ylim(c(0,3))

ncells_p_annotated <- ncells_p_annotated +
  scale_color_manual(values = annotation_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

ncells_p_annotated

# Figure 1C
ggsave(ncells_p_annotated, file = "export.dir/Figures/ncells_ileum.pdf", height=3, width = 5)

ncells_cluster$prop <- (ncells_cluster$ncells/nrow(merged_ileum@meta.data))*100

```

## Number and proportion of cells

For reporting purposes, below are the numbers and proportions of cells in each of the annotated clusters.

```{r broad clustering cell numbers, echo=FALSE, message=FALSE}

knitr::kable(ncells_cluster)

```











