---
title: "Ileum T cells compass"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}


# temporary solution to libpaths
.libPaths("~/R/x86_64-pc-linux-gnu-library/4.3/")

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev.args=list(png=list(type="cairo")))
options(bitmapType = "cairo")
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")

```

```{r load libraries, echo=FALSE, message=FALSE}

library(tidyverse)
library(OCMSutility)

```


```{r set colors, echo=FALSE, message=FALSE, warning=FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```


# Overview

We ran Compass which basically runs flux-balance analysis (FBA) using metabolic models from Recon2. It predicts the chance of active metabolic pathways in single cells. Given our observation of decreased OXPHOS in CD4 and CD8 memory T cells in our data we have run this to see if it is backed up in metabolic modelling i.e. orthogonal analysis approach.

```{r read data, echo = FALSE, message = FALSE}

reactions <- read.csv("~/work/proj072/analysis/compass/reactions.tsv",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep = "\t",
                      row.names = 1)

# -log10()
reactions <- -log10(reactions + 1)

# remove those close to constant
variablist <- apply(reactions, 1, function(x) max(x) - min(x) > 1e-3)
reactions <- reactions[variablist,]

rx_anno <- read.csv("~/work/proj072/analysis/compass/data/rxn_md.csv",
                    header = TRUE,
                    stringsAsFactors = FALSE)

metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)
rownames(metadata) <- gsub("-", "\\.", rownames(metadata))

samples <- intersect(rownames(metadata),colnames(reactions))
metadata <- metadata[samples,]

```

## Reactions distribution

```{r reaction distribution, echo = FALSE, message = FALSE}

react_distr <- data.frame(reactions = rowMeans(reactions))
ggplot(react_distr, aes(x = reactions)) +
  geom_histogram() +
  theme_classic()

```

## Differential abundance in CD4 T cell compartment

Here we use a simple Wilcox test to test for differences between PSC-UC and healthy.

```{r da test, echo = FALSE, message = FALSE}

metadata_sub <- metadata[metadata$disease_type != "UC",]
reactions_sub <- reactions[,rownames(metadata_sub)]

wilcox_by_row <- function(mat, group, p_adjust_method = "BH") {
  # Convert to matrix
  mat <- as.matrix(mat)
  
  # Basic checks
  if (length(group) != ncol(mat)) {
    stop("Length of 'group' must match number of columns in 'mat'.")
  }
  
  group <- as.factor(group)
  
  if (length(levels(group)) != 2) {
    stop("'group' must contain exactly 2 groups.")
  }
  
  g1 <- levels(group)[1]
  g2 <- levels(group)[2]
  
  # Run test row by row
  res <- t(apply(mat, 1, function(x) {
    x1 <- x[group == g1]
    x2 <- x[group == g2]
    
    wt <- wilcox.test(x1, x2, exact = FALSE)
    
    c(
      group1_median = median(x1, na.rm = TRUE),
      group2_median = median(x2, na.rm = TRUE),
      statistic = unname(wt$statistic),
      p_value = wt$p.value
    )
  }))
  
  res <- as.data.frame(res)
  res$p_adj <- p.adjust(res$p_value, method = p_adjust_method)
  res$feature <- rownames(mat)
  
  # Reorder columns
  res <- res[, c("feature", "group1_median", "group2_median", "statistic", "p_value", "p_adj")]
  
  rownames(res) <- NULL
  return(res)
}

res <- na.omit(wilcox_by_row(reactions_sub, metadata_sub$disease_type))
res <- res[order(res$p_adj, decreasing = FALSE),]

# Annotate
rownames(rx_anno) <- rx_anno$rxn_code_nodirection
res$feature_nodirection <- gsub("_.*", "", res$feature)
res$annotation <- rx_anno[res$feature_nodirection,]$rxn_name_long
res$subsystem <- rx_anno[res$feature_nodirection,]$subsystem
res[grep("Oxidative phosphorylation", res$subsystem),]
res[grep("Citric acid", res$subsystem),]
res[grep("Glycolysis", res$subsystem),]





```




