---
title: "Immune Cells Compare Differences"
output: html_document
date: "2026-02-17"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")
```

```{r load libraries, echo = FALSE, message = FALSE}

library(Seurat)
library(ggplot2)
library(dplyr)
source("~/devel/proj060/code/R/gex_helper.R")
source("~/devel/proj072/R/scfunctions.R")
library(RColorBrewer)
library(fgsea)
library(futile.logger)
library(harmony)
library(speckle)
library(scWGCNA)
library(circlize)
library(ggraph)
library(igraph)
library(DESeq2)
library(OCMSutility)

```

```{r colors, echo = FALSE, message = FALSE}

group_colors <- list("UC" = "red3", "PSCUC" = blues9[6], "HC" = "grey")

```


```{r create directory, echo = FALSE, message = FALSE}

if (!(dir.exists("export.dir/Comparison"))){
  dir.create("export.dir/Comparison")
}

```

# Overview

This is an analysis of how differences in CD4 T cells look in the ileum relative to the caecum. This will clarify whether there are tissue-specific effects as we have seen previously and whether the metabolic differences are specific to the ileum and/or cytokine responses are specific to the caecum.

## Comparing cell types

```{r compare immune cell types, echo = FALSE, message = FALSE}

ileum_cd4 <- readRDS("export.dir/Immune/ileum_cd4_memory.Rds")
Idents(ileum_cd4) <- "Ileum"
caecum_cd4 <- readRDS("export.dir/Immune/caecum_cd4_memory.Rds")
Idents(caecum_cd4) <- "Caecum"

all_cd4 <- merge(ileum_cd4, caecum_cd4)

# Remove cells that express JCHAIN
all_cd4 <- subset(all_cd4, subset = JCHAIN < 1)

# get strict on mitochondrial reads
#all_cd4 <- subset(all_cd4, subset = percent.mt < 5)

# recluster etc
all_cd4 <- ScaleData(all_cd4,
                     features = rownames(all_cd4))

all_cd4 <- FindVariableFeatures(all_cd4,
                                selection.method = "vst",
                                nfeatures = 2000)

all_cd4 <- RunPCA(all_cd4,
                  features = VariableFeatures(object = all_cd4))

# harmony integration: by individual = orig.ident
all_cd4 <- all_cd4 %>% 
RunHarmony("orig.ident", plot_convergence = TRUE)
all_cd4 <- FindNeighbors(all_cd4,
                         reduction="harmony",
                         dims = 1:30)
all_cd4 <- RunUMAP(all_cd4,
                   dims = 1:30,
                   reduction = "harmony")

tissue_colors <- c(c(blues9[4], blues9[8]))
names(tissue_colors) <- c("Ileum", "Caecum")

dplot_compare <- DimPlot(all_cd4, reduction = "pca") +
  scale_color_manual(values = tissue_colors) 
  
dplot_compare

ggsave(dplot_compare + theme(legend.position = "none"),
       file = "export.dir/Comparison/cd4_tissue_compare_pca.pdf",
       height = 5, width = 5)

```

## OXPHOS genes


```{r tissue specific genes, echo = FALSE, message = FALSE}

agg_cd4 <- as.data.frame(AggregateExpression(all_cd4, group.by = "orig.ident"))

colnames(agg_cd4) <- gsub("RNA\\.", "", colnames(agg_cd4))

all_cd4$tissue <- Idents(all_cd4)

agg_meta <- distinct(data.frame(sample = all_cd4@meta.data$orig.ident,
                                tissue = all_cd4@meta.data$tissue))


rownames(agg_meta) <- gsub("-", "\\.", agg_meta$sample)
agg_meta <- agg_meta[colnames(agg_cd4),]

## DESeq analysis between tissues
agg_meta <- agg_meta[colnames(agg_cd4),]
dds_agg <- DESeqDataSetFromMatrix(agg_cd4,
                                  colData = agg_meta,
                                  design = ~tissue)
colData(dds_agg)$tissue <- relevel(colData(dds_agg)$tissue, ref = "Ileum")

dds_agg <- DESeq(dds_agg)
agg_res <- lfcShrink(dds_agg,
                     contrast = c("tissue", "Ileum", "Caecum"),
                     res = results(dds_agg),
                     type = "normal")

# OXPHOS genes etc
pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")

oxphos <- pathways[["GO_OXIDATIVE_PHOSPHORYLATION"]]

# just get the oxhos genes that are in the leading edge part of the 
# cd4 cells
oxphos_leading <- read.csv("export.dir/oxphos_leading_edge.tsv", sep = "\t", header = TRUE)$oxphos_leading

# and those that are significantly different between tissues
#sig_tissue <- rownames(agg_res[agg_res$padj < 0.05 & !is.na(agg_res$padj) & abs(agg_res$log2FoldChange) > 1,])
#oxphos_leading <- intersect(oxphos_leading, sig_tissue)

volcano_plot_2 <- function(x, use="padj", value=0.05, labels = NULL){
  
  if (is.null(labels)){
    x$label <- ifelse(x[,use] < value &!(is.na(x[,use])), rownames(x), NA)
  }else{
    x$label <- ifelse(rownames(x) %in% labels, rownames(x), NA)
  }
  
  x$color <- ifelse(x[,use] < value &!(is.na(x[,use])) & x$log2FoldChange > 1, "Ileum", NA)
  x$color <- ifelse(x[,use] < value &!(is.na(x[,use])) & x$log2FoldChange < -1, "Caecum", x$color)
  ggplot(x, aes(x=log2FoldChange, y = -log10(pvalue), color=color)) +
    geom_point(size = 1, alpha = 0.5) +
    geom_vline(xintercept=c(-1,1), linetype = "dashed") +
    ggrepel::geom_text_repel(label = x$label, max.overlaps=100, color = "black") +
    theme_classic() +
    scale_color_manual(values = c(tissue_colors)) 
}

vplot <- volcano_plot_2(agg_res, labels = oxphos_leading) +
  theme(legend.position = "none")

ggsave(vplot, file = "export.dir/Comparison/cd4_volcano_caecum_ileum.pdf", height = 5, width = 5)

vplot
```


```{r Ig expression, echo = FALSE, message = FALSE}

FeaturePlot(all_cd4, features = c("JCHAIN", "MS4A1", "CD79A"), split.by = "tissue")


```


## Comparison of results

Here we plot the fold changes in PSC-UC vs HC in each tissue for each gene and annotate those that are to do with OXPHOS and cytokine production.


### Number of significant changes in each tissue

```{r numbers and overlaps, echo = FALSE, message = FALSE}

ileum_res <- read.csv("export.dir/ileum_cd4_mem_pscuc_vs_healthy.tsv",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep = "\t",
                      row.names = 1)

caecum_res <- read.csv("export.dir/caecum_cd4_mem_pscus_vs_healthy.tsv",
                      header = TRUE,
                      stringsAsFactors = FALSE,
                      sep = "\t",
                      row.names = 1)

# what is the overlap of tested genes
tested_overlap <- list("Ileum" = rownames(ileum_res),
                       "Caecum" = rownames(caecum_res))
v_tested <- VennDiagram::venn.diagram(tested_overlap,
                                      filename = NULL,
                                      fill = c(blues9[3], blues9[8]),
                                      alpha = 0.6)
grid.draw(v_tested)
pdf("export.dir/Comparison/immune_compare_tested_venn.pdf", height = 5, width = 5)
grid.draw(v_tested)
dev.off()

# subset for significant - use later
ileum_sig <- subset(ileum_res, subset = padj < 0.05)
caecum_sig <- subset(caecum_res, subset = padj < 0.05)


```


### Functions associated with ileum-only genes

```{r functions in ileum genes, echo = FALSE, message = FALSE}

pathways <- gmtPathways("~/mirror/msigdb/msigdb_v7.1/msigdb_v7.1_GMTs/c5.bp.v7.1.symbols.gmt")

ileum_specific <- setdiff(rownames(ileum_res), rownames(caecum_res))
background <- rownames(ileum_res)

ileum_specific_pathways <- fgsea::fora(pathways, ileum_specific, universe = background)
ileum_specific_pathways <- ileum_specific_pathways[order(ileum_specific_pathways$pad, decreasing = FALSE),]

```

Lipid metabolic pathways and lipprotein assembly etc - to be expected but useful information nevertheless.


## fgsea of OXPHOS in the caecum

As we have seen that CD4 memory T cells are quite similar in the ileum and caecum albeit with more lipid metabolic processes ocurring in the ileum, I want to see if there is also an enrichment for OXPHOS in CD4 memory T cells in the caecum as we saw in the ileum.

```{r fgsea caecum, echo = FALSE, message = FALSE}

oxphos <- list("GO_OXIDATIVE_PHOSPHORYLATION" = pathways[["GO_OXIDATIVE_PHOSPHORYLATION"]])
genes <- caecum_res$log2FoldChange
names(genes) <- rownames(caecum_res)

fgsea_oxphos_caecum <- fgsea(oxphos, genes)

fgsea::plotEnrichment(pathways[["GO_OXIDATIVE_PHOSPHORYLATION"]], genes) +
  geom_line(aes(x = rank, y = ES), color = blues9[8], linewidth = 1) +
  theme_classic()

ggsave("export.dir/Comparison/fgsea_oxphos_caecum.pdf", height = 5, width = 5)

```


```{r fgsea ileum, echo = FALSE, message = FALSE}

ileum_genes <- ileum_res$log2FoldChange
names(ileum_genes) <- rownames(ileum_res)

fgsea_oxphos_ileum <- fgsea(oxphos, ileum_genes)

# Get the shared leading edge genes
shared_leading_edge <- intersect(unlist(fgsea_oxphos_ileum$leadingEdge),
                                 unlist(fgsea_oxphos_caecum$leadingEdge))


# Heatmap the shared leading edge genes

vst_cd4_all <- assay(vst(dds_agg))

row_annotation <- data.frame(gene = shared_leading_edge)
row_annotation$type <- "OXPHOS"


# Should scale and heatmap ileum and caecum 
# separately
metadata <- read.csv("/well/powrie/projects/proj072/analysis/GEX/05_cellranger_pipeline/samples.tsv",
                     header=TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names=2)
rownames(metadata) <- gsub("-", "\\.", rownames(metadata))


ileum_samples <- intersect(rownames(metadata[metadata$tissue == "Ileum",]),
                           colnames(vst_cd4_all))
caecum_samples <- intersect(rownames(metadata[metadata$tissue == "Caecum",]),
                           colnames(vst_cd4_all))

vst_cd4_ileum <- vst_cd4_all[,ileum_samples]
vst_cd4_caecum <- vst_cd4_all[,caecum_samples]

vst_cd4_ileum_s <- ScaleData(vst_cd4_ileum)
vst_cd4_caecum_s <- ScaleData(vst_cd4_caecum)

# separate Heatmaps

meta_caecum <- metadata[colnames(vst_cd4_caecum_s),]
meta_ileum <- metadata[colnames(vst_cd4_ileum_s),]
meta_caecum$disease_type <- factor(meta_caecum$disease_type, levels = c("HC", "UC", "PSCUC"))
meta_ileum$disease_type <- factor(meta_ileum$disease_type, levels = c("HC", "UC", "PSCUC"))


column_anno_il <- HeatmapAnnotation(
  disease = meta_ileum$disease_type,
  col = list(disease = c("UC" = group_colors[[1]],
                         "PSCUC" = group_colors[[2]],
                         "HC" = group_colors[[3]])))

h_ileum <- Heatmap(vst_cd4_ileum_s[shared_leading_edge,],
           column_split = meta_ileum$disease_type,
           cluster_columns = FALSE,
           show_column_names = FALSE,
           top_annotation = column_anno_il,
           row_names_gp = gpar(fontsize = 6),
           rect_gp = gpar(col = "white", lwd = 0.5))

column_anno_cae <- HeatmapAnnotation(
  disease = meta_caecum$disease_type,
  col = list(disease = c("UC" = group_colors[[1]],
                         "PSCUC" = group_colors[[2]],
                         "HC" = group_colors[[3]])))

h_caecum <- Heatmap(vst_cd4_caecum_s[shared_leading_edge,],
            column_split = meta_caecum$disease_type,
            cluster_columns = FALSE,
            show_column_names = FALSE,
            top_annotation = column_anno_cae,
            row_names_gp = gpar(fontsize = 6),
            rect_gp = gpar(col = "white", lwd = 0.5))
pdf("export.dir/Comparison/caecum_ileum_cd4_oxphos_heatmap.pdf", height = 5, width = 5)
h_ileum + h_caecum
dev.off()



# Exploring Nuclear genome encoded components
nuclear_cytochrome <- c("COX5A",
                        "COX5B",
                        "COX6C",
                        "COX7C",
                        "COX4A",
                        "COX6B",
                        "COX7B",
                        "COX6A",
                        "COX7A",
                        "COX8A",
                        "NDUFA4")
nuclear_leading <- intersect(nuclear_cytochrome, shared_leading_edge)

featurebox(as.data.frame(vst_cd4_caecum),
           metadata[colnames(vst_cd4_caecum),],
           features = c("IL17A", "IL23R"),
           group_by = "disease_type") +
  facet_wrap(~feature,scales = "free")


```


```{r comparison, echo = FALSE, message = FALSE}

oxphos <- pathways[["GO_OXIDATIVE_PHOSPHORYLATION"]]
cytokines <- pathways[["GO_CYTOKINE_PRODUCTION"]]


caecum_res <- caecum_res[order(caecum_res$padj, decreasing = FALSE),]

shared_genes <- intersect(rownames(caecum_res), rownames(ileum_res))

comparison_set <- data.frame("Caecum" = caecum_res[shared_genes,]$log2FoldChange,
                             "Ileum" = ileum_res[shared_genes,]$log2FoldChange,
                             "Gene" = shared_genes)

comparison_set$type <- ifelse(comparison_set$Gene %in% shared_leading_edge, "OXPHOS", NA)

comparison_set$label <- ifelse(comparison_set$type == "OXPHOS", comparison_set$Gene, NA)

ggplot(comparison_set, aes(x = Ileum, y = Caecum, color = type)) +
  geom_point(size = 0.3) +
  theme_classic() +
  geom_hline(yintercept = c(-0.5,0.5, 0), linetype = "dashed") +
  geom_vline(xintercept = c(-0.5,0.5, 0), linetype = "dashed") +
  ylim(c(-3,3))  +
  ggrepel::geom_text_repel(label = comparison_set$label) +
  stat_smooth(method = "lm") +
  ggpubr::stat_cor() +
  geom_abline(slope = 1)


```


# IgA plasma cells

Here we just do a quick comparison of IgA plasma cells between the tissues. The main focus of the paper is not on this cell type.

```{r plasma cells compare, echo = FALSE, message = FALSE}

pileum <- read.csv("export.dir/Immune/ileum_proportions.tsv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   sep = "\t")

pileum$tissue <- "Ileum"

pcaecum <- read.csv("export.dir/Immune/caecum_proportions.tsv",
                   header = TRUE,
                   stringsAsFactors = FALSE,
                   sep = "\t")

pcaecum$tissue <- "Caecum"

pcells <- bind_rows(pileum, pcaecum)
pcells <- pcells[pcells$cluster == "Plasma_Cell_IgA",]
ggplot(pcells, aes(x = tissue, y = prop, fill = tissue)) +
  geom_bar(stat = "identity", position = "dodge") + 
  scale_fill_manual(values = tissue_colors) +
  theme_classic()

ggsave("export.dir/Comparison/IgA_pcells.pdf", height = 3, width = 3)

```






