---
title: "Germ-free vs. SPF for GSTA1-associated genes"
output: html_document
date: "2025-05-04"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="/well/powrie/users/utu691/work/proj072/analysis/final_analysis")
```

```{r load libraries, echo=FALSE, message=FALSE}

library(Seurat)
library(DESeq2)
library(tidyverse)
source("~/devel/proj072/R/scfunctions.R")

```

## Overview

I have a hypothesis that GSTA1 enterocytes repond to bile acids and this may be in some way mediating the expression of GSTA1 and detoxification genes. With available scRNA-seq data in mouse ileum we can start to look at microbiome dependence on these pathways - GF mice are know to have increased ileal TCA for example because they are unable to deconjugate primary bile acids.

```{r read data, echo=FALSE, message=FALSE}

mouse_ileum <- readRDS("/well/powrie/shared/scrnaseq_shen/whole_final_annotation.rds")

mouse_ileum <- subset(mouse_ileum, subset = tissue == "ileum")
Idents(mouse_ileum) <- mouse_ileum@meta.data$celltypes

spf <- subset(mouse_ileum, subset = mice == "SPF")

```


## Do mouse enterocytes show a GSTA1 hi and APOA1 hi phenotype

First we investigate if mouse (SPF) have the same "cell states" in enterocytes as we saw in human ileum.

```{r enterocye cell states, echo = FALSE, message = FALSE}

spf_colors <- OCMSutility::get_palette(n = length(unique(Idents(spf))))
names(spf_colors) <- unique(Idents(spf))

dplot_spf <- DimPlot(spf, label = TRUE) +
  scale_color_manual(values = spf_colors)

dplot_spf + theme(legend.position = "none")

ggsave(file = "export.dir/mouse_dplot.pdf", height = 2.5, width = 2.5)

# subset for enterocytes
spf_entero <- subset(spf, ident = "Enterocyte")

DimPlot(spf_entero)

# recluster etc
spf_entero <- ScaleData(spf_entero, features = rownames(spf_entero))
spf_entero <- FindVariableFeatures(spf_entero, selection.method = "vst", nfeatures = 2000)
spf_entero <- RunPCA(spf_entero, features = VariableFeatures(object = spf_entero))

# featureplot the genes of interest
annotation_colors_epithelium <- OCMSutility::get_palette(n = 50)[10:18]
#names(annotation_colors_epithelium) <- unique(Idents(epithelium))

gsta1_mouse <- feature_plot(spf_entero, feature = "Gsta1", high = annotation_colors_epithelium[2])
apoa1_mouse <- feature_plot(spf_entero, feature = "Apoa1", high = annotation_colors_epithelium[1])

gridExtra::grid.arrange(gsta1_mouse, apoa1_mouse, nrow=1)

ggsave(gsta1_mouse, file = "export.dir/mouse_enterocytes_gsta1.pdf", height = 2.5, width = 3.5)
ggsave(apoa1_mouse, file = "export.dir/mouse_enterocytes_apoa1.pdf", height = 2.5, width = 3.5)

```

They do kind of exist in the SPF mice - although not as many Gsta1 positive cells in general.

## GSTA1 and APOA1 metagenes

```{r metagenes, echo=FALSE, message=FALSE}

gsta1_genes <- readRDS("export.dir/gsta1_cluster.Rds")

gsta1_genes <- gsta1_genes[gsta1_genes$Membership > 0.6,]
  
gsta1_genes <- paste0(substr(rownames(gsta1_genes), 1, 1), unlist(tolower(substr(rownames(gsta1_genes), 2, nrow(gsta1_genes)))))

gsta1_genes <- intersect(gsta1_genes, rownames(spf_entero))

spf_entero@meta.data$Gsta1_Metagene <- colMeans(as.data.frame(GetAssayData(spf_entero)[gsta1_genes,]))

FeaturePlot(spf_entero, features = "Gsta1_Metagene", reduction = "pca")

apoa1_genes <- readRDS("export.dir/apoa1_cluster.Rds")

apoa1_genes <- apoa1_genes[apoa1_genes$Membership > 0.6,]
  
apoa1_genes <- paste0(substr(rownames(apoa1_genes), 1, 1), unlist(tolower(substr(rownames(apoa1_genes), 2, nrow(apoa1_genes)))))

apoa1_genes <- intersect(apoa1_genes, rownames(spf_entero))

spf_entero@meta.data$Apoa1_Metagene <- colMeans(as.data.frame(GetAssayData(spf_entero)[apoa1_genes,]))

FeaturePlot(spf_entero, features = c("Gsta1_Metagene", "Apoa1_Metagene"), reduction = "umap")

```

## Is the proportion of GSTA1 postive cells changed in GF relative to SPF

```{r spf vs gf gsta1, echo=FALSE, message=FALSE}

gf <- subset(mouse_ileum, subset = mice == "GF")
gf_entero <- subset(gf, ident = "Enterocyte")


all_entero <- subset(mouse_ileum, ident="Enterocyte")

# Subsample
spf_entero <- spf_entero[, sample(colnames(spf_entero), size=1000, replace=F)]
gf_entero <- gf_entero[, sample(colnames(gf_entero), size=1000, replace=F)]

all_entero_sub <- merge(spf_entero, gf_entero)

all_entero_sub <- RunPCA(FindVariableFeatures(ScaleData(all_entero_sub)))

spf_entero <- subset(all_entero_sub, subset = mice == "SPF")
gf_entero <- subset(all_entero_sub, subset = mice == "GF")

gf_featureplot <- feature_plot(gf_entero,
                  feature = "Gsta1",
                  high = annotation_colors_epithelium[2])

spf_featureplot <- feature_plot(spf_entero,
                   feature = "Gsta1",
                   high = annotation_colors_epithelium[2])


gridExtra::grid.arrange(gf_featureplot, spf_featureplot, nrow = 1)
ggsave(file = "export.dir/gf_spf_gsta1_featureplot.pdf", height = 2.5, width = 6)


ppos_gsta1_mouse <- get_ppos(all_entero_sub, gene = "Gsta1")

ppos_gsta1_mouse$status <- gsub("(_.*)", "", ppos_gsta1_mouse$individual)

ppos_gsta1_mouse$status <- factor(ppos_gsta1_mouse$status, levels = c("SPF", "GF"))
ggplot(ppos_gsta1_mouse, aes(x = status, y=ppos, color = status)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_jitter(height = 0, width = 0.1) +
  theme_classic() +
  ggpubr::stat_compare_means(comparisons = list(c(1,2))) +
  scale_color_manual(values = c("grey", "purple")) +
  theme(legend.position = "none")
  
ggsave("export.dir/spf_gf_ppos_gsta1.pdf", height = 2.5, width = 5)

```

### Pseudobulk with a focus on GSTA1-associated genes

```{r pseudobulk of enterocytes, echo=FALSE, message=FALSE}

agg_entero <- as.data.frame(AggregateExpression(all_entero, group.by="orig.ident"))

# make metadata
spf_gf_meta <- data.frame(sample = colnames(agg_entero),
                          status = gsub("\\.ileum[0-9]", "", gsub("RNA\\.", "", colnames(agg_entero))))
rownames(spf_gf_meta) <- spf_gf_meta$sample

library(DESeq2)

dds <- DESeqDataSetFromMatrix(agg_entero, colData = spf_gf_meta, design = ~status)
colData(dds)$status <- relevel(colData(dds)$status, ref = "SPF")
dds <- DESeq(dds)
results_gf_spf <- lfcShrink(dds,
                     contrast = c("status", "GF", "SPF"),
                     res = results(dds),
                     type = "normal")

results_sig <- subset(results_gf_spf, subset = padj < 0.05 & abs(log2FoldChange) > 1)


# add some labels and coloring things etc
# some padj are 0 so should make them the min where
# the mine != 0
results_gf_spf$padj[results_gf_spf$padj == 0] <- min(results_gf_spf$padj[results_gf_spf$padj !=0 & !is.na(results_gf_spf$padj)])

results_gf_spf$sig <- ifelse(results_gf_spf$padj < 0.05 & abs(results_gf_spf$log2FoldChange) > 1, "Sig", NA)

results_gf_spf$label <- ifelse(rownames(results_gf_spf) %in% c("Fabp1", "Fabp6", "Mt1", "Gsta1", "Rbp2", "Tmsb10"), rownames(results_gf_spf), NA)

ggplot(results_gf_spf, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(size = 0.2) +
  theme_classic() +
  scale_color_manual(values = "blue3") +
  ggrepel::geom_text_repel(label = results_gf_spf$label, color = "black") +
  geom_vline(xintercept = c(-1,1), linetype = "dashed") +
  geom_vline(xintercept = 0) +
  theme(legend.position = "none")

ggsave("export.dir/mouse_gf_spf_volcano.pdf", height = 5, width = 5)
```








